---
title: "Latent ordination in the 200-species model"
author: "Christopher J Walsh"
format: docx
editor: source
bibliography: references.bib
editor_options: 
  chunk_output_type: inline
---

```{r}
#| label: setup
#| echo: false

source("quick_200_spp_data_assembly_for_latent_tests.R")
# This creates data ready for the model for constrained_z
source("cmdstanr_functions.R")
# post processing functions

```

The run_200spp_model() function (loaded in "quick_200_spp_data_assembly_for_latent_tests.R") runs the specified 200-species model and creates a list of objects for post-processing ("species_set", "sample_set", "pred_set", "data",       "fit", "diagnostics", "summary").  This should take ~ 1.5 h.

```{r}
#| eval: false

run_200spp_model(pred_site = pred_site,
                 mod_code = "ifqcdtblrs_fi_qf_qi_ou",
                 mod_path = "stancode/jointspp_pa_no_eps_ou.stan", 
                 cal_loo = TRUE)

```

The need for post-processing is illustrated in @fig-alignment, showing the raw traceplots of the four chains for latent axes 1 and 2 for site 1 (A. z_expanded[1,1], B. z_expanded[1,2]) from the Stan model. They show that the score estimates for each chain drifting independently over the 1200 iterations.  This is because the model does not constrain the ordination to a set rotation (and attempting to do so produced stable, but split chains that did not converge).  The following code chunk applies a Procrustes rotation to the ordination in each iteration, which results in well-mixed, stable traceplots for the aligned z-values (C and D).

```{r}
#| label: fig-alignment
#| echo: false
#| fig-width: 5
#| fig-height: 5
#| fig-cap = "Traceplots of raw estimates of example latent factor score for site 1 (A. axis 1, B. axis 2) for 1200 iterations in the 200-species model.  C and D show the same estimates after Procrustes rotation."

load("~/uomShare/wergStaff/ChrisW/git-data/mw_metabarcoding_model/examp_draws.rda")
par(mfrow = c(2,2), mar = c(4,4,1,1))
base_traceplot(examp_draws$z11_unaligned, "z_expanded[1,1]"); title(main = "A.", adj = 0)
base_traceplot(examp_draws$z12_unaligned, "z_expanded[1,2]"); title(main = "B.", adj = 0)
base_traceplot(examp_draws$z1_aligned[,,1, drop = FALSE], "z_aligned[1,1]"); title(main = "C.", adj = 0)
base_traceplot(examp_draws$z1_aligned[,,2, drop = FALSE], "z_aligned[1,2]"); title(main = "D.", adj = 0)

```

The following code aligns the latent factor parameters z_expanded and lambda.

```{r}

mod_dir <- "~/uomShare/wergStaff/ChrisW/git-data/mw_metabarcoding_model/trial_200spp_models/"
mod_bundle <- load_model_component_bundle(mod_code ="ifqcdtblrs_fi_qf_qi_ou", mod_dir = mod_dir)
mod_fit <-  mod_bundle$fit
mod_data <- mod_bundle$data

# Ensure your dimensions are correctly defined for your active model
n_obs <- mod_data$n_obs
n_site <- mod_data$n_site
n_taxa <- mod_data$n_taxa
n_latent <- mod_data$n_latent
n_chains <- mod_fit$num_chains()
n_iterations <- mod_fit$metadata()$iter_sampling

# dimensions  [1200 iterations, 4 chains, 550 flattened parameters]
raw_z <- mod_fit$draws("z_expanded", format = "array")
raw_lambda <- mod_fit$draws("lambda_raw", format = "array")

# Establish a baseline template using a stable iteration 
# example_z <- mod_fit$draws("z_expanded[1,1]", format = "array")
# base_traceplot(example_z, "z_expanded[1,1]") # all respectable. Pick chain 1
# Save example unaligned (and later aligned) z scores for plotting above.
# examp_draws <- list(z11_unaligned = example_z, 
#                     z12_unaligned = mod_fit$draws("z_expanded[1,2]", format = "array"))
first_vector <- raw_z[500, 4, ]

target_matrix <- matrix(NA, nrow = n_obs, ncol = n_latent)
target_matrix[, 1] <- first_vector[1:n_obs]           # Elements 1 to 275 = Axis 1
target_matrix[, 2] <- first_vector[(n_obs + 1):550]    # Elements 276 to 550 = Axis 2

# Standardise the master frame
target_matrix <- scale(target_matrix) 

# Create empty, 4D arrays to store the aligned results
# dim =iterations, chains, observations, latent axes
z_unpacked_aligned <- array(NA, dim = c(n_iterations, n_chains, n_obs, n_latent))
lambda_unpacked_aligned <- array(NA, dim = c(n_iterations, n_chains, n_taxa, n_latent))

for (c in 1:n_chains) {
  for (i in 1:n_iterations) {
    # current 550-element vector
    current_vector <- raw_z[i, c, ]
    # Create an empty  275 x 2 matrix to be populated
    current_matrix <- matrix(NA, nrow = n_obs, ncol = n_latent)
    current_matrix[, 1] <- current_vector[1:n_obs]            # Dimension 1 (Axis 1)
    current_matrix[, 2] <- current_vector[(n_obs + 1):550]    # Dimension 2 (Axis 2)
    # Standardise to remove cross-chain scale discrepancies
    current_matrix_scaled <- scale(current_matrix)
    if (any(is.na(current_matrix_scaled)) || any(is.infinite(current_matrix_scaled))) {
      next 
    }
    # Rotate the iteration onto the master template: dimensions [275,2]
    rotation <- vegan::procrustes(X = target_matrix, Y = current_matrix_scaled, symmetric = FALSE)
    z_unpacked_aligned[i, c, , ] <- rotation$Yrot
    # Extract the 2 x 2 rotation matrix calculated for this iteration for aligning the species loadings (lambda)
    rotation_matrix <- rotation$rotation
    current_lambda_vector <- raw_lambda[i, c, ]
    current_lambda_matrix <- matrix(current_lambda_vector, nrow = n_taxa, ncol = n_latent)
    # multiply lambda by the same rotation matrix
    lambda_unpacked_aligned[i, c, , ] <- current_lambda_matrix %*% rotation_matrix
  }
}
# Add to data for example plot above
# examp_draws$z1_aligned <- z_unpacked_aligned[,,1,]
# save(examp_draws, file = "~/uomShare/wergStaff/ChrisW/git-data/mw_metabarcoding_model/examp_draws.rda")

####
# 1. Recalculate draws for aligned Z (latent factor scores for sites)
# z_expanded values are identical for observations from the same sites, so just use one value per site
unique_site_ids <- 1:n_site
first_obs_indices <- match(unique_site_ids, mod_data$site)
all_sites_samples_raw <- z_unpacked_aligned[, , first_obs_indices, ] # dim [1200, 4, 223, 2]
# Reshape from a 4-D array into a standard 3-D MCMC array: [1200, 4, 446]
n_parameters <- n_site * n_latent
all_sites_samples_flat <- array(all_sites_samples_raw, dim = c(n_iterations, n_chains, n_parameters))
parameter_names <- character(n_parameters)
idx <- 1
for (s in unique_site_ids) {
  parameter_names[idx]     <- paste0("Site_", s, "_Z1")
  parameter_names[idx + 1] <- paste0("Site_", s, "_Z2")
  idx <- idx + 2
}# Apply the naming dimensions to the reshaped array
dimnames(all_sites_samples_flat) <- list(
  Iteration = 1:n_iterations,
  Chain     = 1:n_chains,
  Parameter = parameter_names
)
# The formal draws object 
all_sites_aligned_draws <- as_draws_array(all_sites_samples_flat)

# 2. Recalculate draws for aligned lambda (latent factor scores for species)
n_lambda_params <- n_taxa * n_latent
lambda_flat <- array(lambda_unpacked_aligned, dim = c(n_iterations, n_chains, n_lambda_params))
species_names <- colnames(mod_data$y)
lambda_names <- character(n_lambda_params)
idx <- 1
for (j in 1:n_taxa) {
  lambda_names[idx]     <- paste0("Species_", species_names[j], "_Load_Axis1")
  lambda_names[idx + 1] <- paste0("Species_", species_names[j], "_Load_Axis2")
  idx <- idx + 2
}
dimnames(lambda_flat) <- list(
  Iteration = 1:n_iterations,
  Chain     = 1:n_chains,
  Parameter = lambda_names
)
# Cast into a formal MCMC draws object
lambda_aligned_draws <- as_draws_array(lambda_flat)
```

From these aligned z and lambda draws, we revised the diagnostic summary for the model using the following code to produce @tbl_diags, which shows the model to be in good order.

```{r}
#| label: "tbl-diags"

lambda_diagnostic_summary <- summarise_draws(lambda_aligned_draws)
z_diagnostic_summary <- summarise_draws(all_sites_aligned_draws)

mod_summary <- mod_bundle$summary[-grep("z_exp|lambda",mod_bundle$summary$variable),]
mod_summary <- rbind(mod_summary, lambda_diagnostic_summary,z_diagnostic_summary)
mod_summary$variable[which(mod_summary$ess_bulk == min(mod_summary$ess_bulk, na.rm = TRUE))]
diag_summary <- data.frame(spec = c("No. sampling iterations","No. divergences",
                                    "No. max_treedepth exceedences",
                                    "BFMI (Chain 1)","BFMI (Chain 2)","BFMI (Chain 3)","BFMI (Chain 4)",
                                    "Min. Bulk ESS", "Min. Tail ESS","Max R-hat",
                                    "Worst sampled parameter"),
                           value = c(n_iterations,
                                     sum(mod_bundle$diagnostics$num_divergent),
                                     sum(mod_bundle$diagnostics$num_max_treedepth),
                                     round(mod_bundle$diagnostics$ebfmi,2),
                                     round(min(mod_summary$ess_bulk, na.rm = TRUE)),
                                     round(min(mod_summary$ess_tail, na.rm = TRUE)),
                                     round(max(mod_summary$rhat, na.rm = TRUE),2),
                                     mod_summary$variable[which(mod_summary$ess_bulk == min(mod_summary$ess_bulk, na.rm = TRUE))])
                           )
ft_to_word(flextable::flextable(diag_summary))

```


```{r}
#| label: fig-latent_factor_ordination
#| echo: false
#| fig-width: 5
#| fig-height: 5
#| fig-cap = "Ordination of latent factors explaining random variation among sites (grey points), overlain by mean locations of species in the ordination space (red points)."

# Plot latent site ordination
mean_obs_z1 <- apply(z_unpacked_aligned[, , , 1], 3, mean, na.rm = TRUE)
mean_obs_z2 <- apply(z_unpacked_aligned[, , , 2], 3, mean, na.rm = TRUE)
ordination_df <- data.frame(Site = mod_data$site, 
  Z1   = mean_obs_z1,
  Z2   = mean_obs_z2
)
# Site scores are unique, so reduce to one value per site
unique_site_ordination <- unique(ordination_df)
plot(unique_site_ordination$Z1, unique_site_ordination$Z2,  pch = 21, bg = grey(0.8), 
     xlab = "Latent Axis 1", 
     ylab = "Latent Axis 2", asp = 1)
abline(v = 0, h = 0, lty = 2)

samples <- mod_bundle$sample_set
sites <- unique(samples[c("reach_v12","site_no")])

# Map species to the ordination - put z and lambda on the same scale first
species <- mod_bundle$species_set
mean_lambda1 <- apply(lambda_unpacked_aligned[, , , 1], 3, mean, na.rm = TRUE)
mean_lambda2 <- apply(lambda_unpacked_aligned[, , , 2], 3, mean, na.rm = TRUE)
max_site_range <- max(abs(c(unique_site_ordination$Z1, unique_site_ordination$Z2)))
max_spp_range  <- max(abs(c(mean_lambda1, mean_lambda2)))
# The scale factor adjusts species to occupy roughly 80% of the site space
scale_factor <- (max_site_range / max_spp_range) * 0.8
species_ord <- data.frame(species = species$taxon[match(colnames(mod_bundle$data$y),species$taxoncode)],
                          taxoncode = colnames(mod_bundle$data$y),
                          axis_1  = mean_lambda1 * scale_factor, 
                          axis_2  = mean_lambda2 * scale_factor)
points(species_ord$axis_1,species_ord$axis_2, pch = 16, cex = 0.5, col = "red")

```


```{r}

# # 5 sites with latent axis 1 scores > 2.5
# # used to manually check them first
# identify(unique_site_ordination$Z1, unique_site_ordination$Z2, unique_site_ordination$Site)
high_z2s <- unique_site_ordination[unique_site_ordination$Z2 > 2,]
sites[sites$site_no %in% high_z2s$Site,]
low_z1s <-  unique_site_ordination[unique_site_ordination$Z1 < -2.1,]
sites[sites$site_no %in% low_z1s$Site,]
# "BNY_24764" "GIS_1106"  "LOL_11925" "PRW_13767" "REI_3934" 
# A selection of sites that don't at first blush have anything in common
# Species associated with those sites
species_ord[species_ord$axis_1 < -2.1,] # all classic disturbance tolerant species
species_ord[species_ord$axis_2 > 2,] # A mix of moderately tolerant and sensitive species
# Perhaps the environmental predictors of these sites suggest they should be good
# but they support some species typical of degraded sites...

# Conduct broader tests for any residual geographical information
coords <- sf::st_coordinates(sites)
dist_matrix <- as.matrix(dist(coords))
# Invert the distance matrix to create a spatial weights matrix (closer = heavier weight)
weights_matrix <- 1 / dist_matrix
diag(weights_matrix) <- 0 
weights_matrix[is.infinite(weights_matrix)] <- 0
# Moran's I to assess for any potential contagious distribution
moran_z1 <- ape::Moran.I(unique_site_ordination$Z1, weights_matrix)
moran_z2 <- ape::Moran.I(unique_site_ordination$Z2, weights_matrix)
# Both >0.5, which suggests randomness without spatial clumping.

# Regressions predicting Axis 1 and 2 from spatial locatoin
Z <- cbind(unique_site_ordination, coords)
spatial_lm1 <- lm(Z1 ~ X * Y, data = Z)
spatial_lm2 <- lm(Z2 ~ X * Y, data = Z)
summary(spatial_lm1)$r.squared  
summary(spatial_lm2)$r.squared  
# Both ~ 0.01, which suggests no residual trend across Euclidean space

```

AUC code

```{r}

library(PRROC)
library(posterior)

# 1. Extract the posterior mean probabilities across all iterations and chains
# This natively returns a clean [n_obs x n_taxa] matrix in R
mean_p_total <- mod_fit$draws("p_total_out")
 mean_p_total <- matrix(mean_p_total, nrow = n_obs, ncol = n_taxa)

mean_p_env   <- summary(mod_fit$draws("p_env_out"))$mean
mean_p_env   <- matrix(mean_p_env, nrow = n_obs, ncol = n_taxa)

# 2. Set up arrays to hold the final AUC metrics for your 200 (or 875) species
roc_auc_total <- numeric(n_taxa)
pr_auc_total  <- numeric(n_taxa)
roc_auc_env   <- numeric(n_taxa)
pr_auc_env    <- numeric(n_taxa)

# 3. Calculate curves species-by-species
for(j in 1:n_taxa) {
  # Actual real-world 0/1 observations for species j
  actual_y <- mod_data$y[, j] 
  
  # Separate probabilities for presences vs absences
  pred_pres_total <- mean_p_total[actual_y == 1, j]
  pred_abs_total  <- mean_p_total[actual_y == 0, j]
  
  pred_pres_env   <- mean_p_env[actual_y == 1, j]
  pred_abs_env    <- mean_p_env[actual_y == 0, j]
  
  # Compute the full-model curves (Environment + Latent factors)
  fg_t <- pred_pres_total; bg_t <- pred_abs_total
  roc_auc_total[j] <- PRROC::roc.curve(scores.class0 = fg_t, scores.class1 = bg_t)$auc
  pr_auc_total[j]  <- PRROC::pr.curve(scores.class0 = fg_t, scores.class1 = bg_t)$auc
  
  # Compute the pure environmental curves (U matrix only)
  fg_e <- pred_pres_env; bg_e <- pred_abs_env
  roc_auc_env[j] <- PRROC::roc.curve(scores.class0 = fg_e, scores.class1 = bg_e)$auc
  pr_auc_env[j]  <- PRROC::pr.curve(scores.class0 = fg_e, scores.class1 = bg_e)$auc
}

# Combine into a gorgeous evaluation summary dataframe!
species_metrics <- data.frame(
  species = species_set$taxon[match(colnames(mod_data$y), species_set$taxoncode)],
  taxoncode = colnames(mod_data$y),
  roc_auc_total = roc_auc_total,  pr_auc_total = pr_auc_total,
  roc_auc_env = roc_auc_env,  pr_auc_env = pr_auc_env)

library(dplyr)

# 1. Calculate the real-world prevalence for each of your 875 species
# colSums(y) gives presences, nrow(y) is 275 total observations
species_prevalence <- colSums(trial_1_data$y) / nrow(trial_1_data$y)

# 2. Attach the prevalence data to your existing AUC summary dataframe
# (Assumes 'species_metrics' is the dataframe generated by our PRROC loop)
species_metrics <- species_metrics %>%
  mutate(
    Prevalence = species_prevalence,
    # Classify each species into its explicit ecological tier
    Tier = case_by(
      Prevalence >= 0.20 ~ "Common (>20%)",
      Prevalence >= 0.05 & Prevalence < 0.20 ~ "Infrequent (5-20%)",
      Prevalence < 0.05  ~ "Truly Rare (<5%)"
    )
  )

# 3. Generate the publication-ready grouped evaluation summary table!
model_evaluation_table <- species_metrics %>%
  group_by(Tier) %>%
  summarise(
    Species_Count = n(),
    Mean_Prevalence = mean(Prevalence),
    # Full Model Metrics (Environment + Latent Space)
    Mean_ROC_Total  = mean(ROC_AUC_Total),
    Mean_PR_Total   = mean(PR_AUC_Total),
    # Pure Environmental Metrics (U Matrix Only)
    Mean_ROC_Env    = mean(ROC_AUC_Env),
    Mean_PR_Env     = mean(PR_AUC_Env)
  )

print(model_evaluation_table)
```

