---
title: "Model data assembly for the 225-site dataset"
author: "Thomas Wilkins"
format: html
editor: source
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false

proj_seed <- 12345 
set.seed(proj_seed)

source("https://tools.thewerg.unimelb.edu.au/mwstr/data/mwstr_network_functions.R")
right <- function(x, n) {substr(x, nchar(x) - n + 1, nchar(x))}
# source("https://raw.githubusercontent.com/tewilkins/mwbugs_functions/refs/heads/main/mwbugs_functions.R")

### Tom, I've commented out these lines in order to explicitly (wherever possible)
###   associate functions with their packages (not always possible with piping
###   in the tidyverse). This means I have likely broken some of the code beyond
###   Section 1.2, which is as far as I have got so far.
# requiredPackages <- c("WriteXLS", "rio", "rstan", "tidyverse", "geosphere")
# load_packages(requiredPackages) # Installs packages if not installed already

# Create a folder to hold temporary data if not already present
system(command = "mkdir -p temp_data/")

# pairs add-on function from pairs() help file
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}

ct <- function (rows, cols, values = NULL, FUN = sum, convertNAToZero = TRUE,...) 
{
  if(!is.vector(rows)) rows <- as.vector(rows)
  if(!is.vector(cols)) cols <- as.vector(cols)
  if(is.null(values)) values <- rep(1,length(rows))
  results <- tapply(values, list(rows, cols), FUN, ...)
  if(convertNAToZero)
    results[is.na(results)] <- 0
  results
}


```

## Introduction

This document contains the assembly of environmental and biological data for the analysis of the 2021-23 macroinvertebrate DNA metabarcoding dataset, prepared for the A. The data is intended for modelling the presence/absences of stream macroinvertebrates with the Stan model `multitaxon_pa_rSAC.stan`, which is readily scalable to larger datasets. The model takes the following declarations as inputs:

-   n_obs: number of observations (sample pairs)
-   n_taxa: number of taxa (species)
-   n_pred: number of predictor variables in the model (will depend on model selection)
-   n_site: number of unique sites (225)
-   n_seasons: number of seasons (2, Autumn and Spring)

And the following derived data as inputs:

-   y: a matrix of presence-absence data for each taxon and observation (n_obs x n_taxa)
-   u: a matrix of predictor data at each observation (n_obs x n_pred)
-   s: a matrix of subsample proportions for each observation (n_obs x n_taxa)
-   M: a matrix of distances between each pair of sites (n_site x n_site)
-   site: a vector of site indices (numbered 1 to n_site) for each observation (n_obs)
-   season: a vector of season indices (1 if Autumn, 2 if Spring) for each observation (n_obs)

### 1.1. Retrieving sample pair and site data from Melbourne Water databases

The location and environmental variables associated with the 225 sites was recorded in the Melbourne Water macroinvertebrate database [@walsh2020]. Environmental data associated with each site was retrieved from the Melbourne Water stream database [@walsh]. The following code chunk retrieves this data from the relevant database tables:

```{r}
#| label: data_retrieval
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

# Open terminal (not in RStudio tab), run:
# ssh -L 5432:localhost:5432 -C -N -l  tewilkins 172.26.150.60

# Connect to MW databases:
mwbugs <- DBI::dbConnect(RPostgres::Postgres(), dbname = "mwbugs",
                          host = "localhost", user = "readonly", password = "reachcode_42")
mwstr <- DBI::dbConnect(RPostgres::Postgres(),dbname = "mwstr",
                          host = "localhost", user = "readonly", password = "reachcode_42")

# Retrieve sample pair data

####  
# Project 3:  Metabarcoded combined-RBA-sample-pair survey of 225 sites across
#   Melbourne 2021-2024 (Nickname MW metabarcoding survey)

samples <- DBI::dbGetQuery(mwbugs, 
                          "SELECT * FROM samples 
                           JOIN sample_project_groups
                           ON samples.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
# returns 275 samples from 225* sites (13 sites have 2 samples, one Autumn and 
# one Spring; the 10 sites from project_no = 1 have up to 5 samples over 3 y)
# * On further investigation, the 5 samples from Toomuc Creek were assigned
#   to 3 different sitecodes. TOO_4269a (trips 377, 396, 401) is really at the
#   top of reach TOO_4332.  TOO_4332a (trip 389) and TOO_4332c (trip 384) are
#   200-250 m downstream in the same reach. All of these locations are upstream
#   of the Henry Road drain from the east, and the Henry Road wetland outlet from
#   the west, and all are downstream of the next large drain upstream (Banjo
#   Patterson Drive). They should therefore all be considered the same site
#   which I will call TOO_4332c. This reduces the number of sites to 223.
samples$sitecode[grep("TOO",samples$sitecode)] <- "TOO-4334-2"
samples$sitecode_v12[grep("TOO",samples$sitecode_v12)] <- "TOO_4332c"
# Check distribution of samples
# x <- aggregate(samples$smpcode, by = list(sitecode = samples$sitecode_v12), FUN = length)
# aggregate(x$sitecode, by = list(n_samps = x$x), FUN = length)
# 200 sites sampled once; 13 sampled twice, 10 (from project 2) sampled 4-5 times.

#### Perhaps also worth considering removing LSN, with a lot of SCMs not accounted for in its EI measurement

biota <- DBI::dbGetQuery(mwbugs, 
                           "SELECT * FROM biota 
                           JOIN sample_project_groups
                           ON biota.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
# returns 20694 records 
# NOTE project 3 includes the samples collected for project 2, which were fully
# processed, with 8-11 replicate MiSeq samples. All species detected in those
# samples are included in the biota table.  The samples can be adjusted to 
# approximate the standard 30% subsampled, 3-replicate bulk-processing method
# by excluding records with the phrase: 
# 'EXCLUDE this record from 3-rep 30% sample' in the notes column 
biota <- biota[!grepl("EXCLUDE this record from 3-rep", 
                          biota$notes),]  #20584
# removes 107 records that were found in other replicates/subsamples
# Check samples have biota:
# nrow(samples[!samples$smpcode %in% unique(biota$smpcode),]) # 0 samples without biota

sites <- sf::st_read(mwbugs, query = 
                      paste0("SELECT * FROM sites WHERE sitecode_v12 IN ('",
                             paste(samples$sitecode_v12, collapse = "','"),"');"))
sites$site_no
# returns 223 sites (a spatial table, so sf:st_read() was used)

```

### 1.2. Between-site distance matrices

We calculated three matrices of distances between sampling sites: 

- Euclidean distances, calculated from Map Grid of Australia 2020 Zone 55 coordinates. Tom, your previous reference to haversine distances, while strictly correct, is an unnecessarily complicated term for a small area such as the Melbourne region. crs 7855 (Map Grid of Australia 2020 Zone 55) is a planar projection applicable to our area, meaning that we can consider the earth (around here) flat.  So, we can just use 7855 coordinates (unit metres) to calculate Euclidean distances.  

- upstream river distances, calculated as the along-stream distance from downstream to upstream for all pairs of sites on the same stream line, using the Melbourne Stream Network database stream lines.  

- downstream river distances, calculated as for upstream river distances, but from upstream sites to downstream sites.

We begin by modelling downstream river distances, as this is likely the most dominant pathway for movement between reaches.  We could potentially add the other two matrices at a later date.

Now, 98.1% of all pairs of the 223 sites in the dataset have Inf distance in these two river distance matrices (unsurprising, as my intention in picking sites was to minimise the possibility of spatial autocorrelation among sites).  I feel like Inf is going to be problematic when you scale the distances....Given the maximum non-Inf value in the dataset is 188 km, I wonder if it would be defensible to just set the Inf values to something like 2 times 188 (or 3 times?), given that a priori, we would expect the influence of sites 188 km apart on each other to be zero... 

I don't really understand your suggested solution

*If that’s going to mess up the scaling of the matrix, then a workaround could be setting the distance to 0km in the distance matrix M (giving an inverse-exponential distance of 1), and then transforming the inverse-weighted matrix W in Stan to set all values of 1 to 0. I’ve done this already for the diagonal values in the distance matrix because this would imply the site is its own neighbour (and thereby conflicting with a_site).*

But I'm happy to go with it if you think it will work...This suggestion also makes me doubt my next point...

I also need to be convinced of the wisdom of calculating a distance matrix on samples (rather than sites) as you do in your model (or perhaps I am misunderstanding what you do?).  I think (but am not 100% sure) that modelling variation within sites using a random site variable is different from (and more correct than?) modelling it using a distance matrix where samples from the same site have distance 0.  I think the more correct way to model this (particularly in this set where we have 23 sites with >1 sample) would be have a random site variable and apply the distance matrix to the sites. (This must be possible surely, given the possibility of applying similarity matrices to the similarly random species?)


```{r}
#| label: site_distance_matrices
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

# 1. Ecuclidean distance matrix
sites$site_no <- 1:nrow(sites)
sites_coords_km <- sf::st_coordinates(sites)/1000
dist_euc_km <- as.matrix(stats::dist(sites_coords_km, diag = TRUE, upper = TRUE))
# dist_euc_scaled <- scale(dist_euc_km, center = FALSE)
# scale() scales by column, which is not what we want to do here.  
# We need to vectorize, scale and return to matrix
x <- as.vector(dist_euc_km)
x <- scale(x, center = FALSE)
dist_euc_scaled <- matrix(x, ncol = ncol(dist_euc_km))
dist_euc_inv_w <- exp(-dist_euc_km)
# set diagonal to zero as in original stan model
for(i in 1:nrow(dist_euc_inv_w)){
  dist_euc_inv_w[i,i] <- 0
}
# Inspect and assess need for scaling
# range(dist_euc_inv_w)  0 to 1, no need for scaling

# 2 and 3. The two river distance matrices (Not yet scaled: needs further consideration)

# Add reach length and site number from mwstr database to sites table
sites$site <- sites$reach_length_m <- NA
for(i in 1:nrow(sites)){
  site_stats <- DBI::dbGetQuery(mwstr, 
                                paste0("SELECT site, length_m FROM streams WHERE reach = '", 
                                       sites$reach_v12[i], "';"))
  sites$reach_length_m[i] <- sum(site_stats$length_m) # some subcs have > 1 line in streams
  sites$site[i] <- site_stats$site
} # takes 10 s

# Load the subcs (reach) table from mwstr and convert it to an igraph network object
subcs <- DBI::dbGetQuery(mwstr, "SELECT reach, site, nextds, tot_length_m FROM subcs;")
network <- subcs[c("site","nextds","reach","tot_length_m")]
network$nextds[network$nextds == -1] <- -(sum(network$nextds == -1):1)
network_i <- apply(network[,1:2], 2, as.character)
network_ig <- igraph::graph_from_data_frame(network_i)
network_df <- igraph::as_long_data_frame(network_ig)

# Subset the network object to link only the 223 sites of interest
network_223 <- data.frame(vertex = match(sites$site, network_i[,1]),
                          site_no = sites$site_no,
                          site = sites$site, 
                          sitecode = sites$sitecode_v12,
                          nextds = NA, nextds_no = NA, dist_km = NA)
network_223 <- network_223[!duplicated(network_223$site),]

for(i in 1:nrow(network_223)){
  all_ds_i <- igraph::subcomponent(network_ig, network_223$vertex[i], 
                                               mode = "out")
  all_ds_223 <- as.numeric(names(all_ds_i)[names(all_ds_i) %in% network_223$site])
  # Next downstream site in pruned network is the second listed subcomponent 
  # (i.e. all downstream)
  network_223$nextds[i] <- all_ds_223[2] #network_223$site_no[network_223$site == all_ds_223[2]]
  if(!is.na(network_223$nextds[i])){
  all_ds_nextds <- as.numeric(names(igraph::subcomponent(network_ig, 
                            all_ds_i[names(all_ds_i) == network_223$nextds[i]], 
                            mode = "out")))
  # all reaches between the site and its nextds (including those two sites)
  reaches_bn <- c(setdiff(names(all_ds_i), all_ds_nextds),network_223$nextds[i])
  ds_sitecode <- sites$sitecode_v12[sites$site == network_223$nextds[i]]
  # Length of all full reaches between the two sites
  network_223$dist_km[i] <- (sum(subcs$tot_length_m[subcs$site %in% reaches_bn]) - 
    # minus the distance upstream of the site on the upstream reach 
    (1.05 - 0.1* match(right(network_223$sitecode[i],1),letters)) * 
    sites$reach_length_m[sites$sitecode_v12 == network_223$sitecode[i]] -
    # minus the distance downstream of the site on the downstream reach
     (0.1* match(right(ds_sitecode,1),letters) - 0.05) * 
    sites$reach_length_m[sites$sitecode_v12 == ds_sitecode])/1000
  network_223$nextds_no[i] <- 
    network_223$site_no[network_223$site == all_ds_223[2]]
  }else{
    network_223$dist_km[i] <- Inf
  }
  } # takes ~10 s

network_223$nextds[is.na(network_223$nextds)] <- -(sum(is.na(network_223$nextds)):1)
network_223$nextds_no[is.na(network_223$nextds_no)] <- -(sum(is.na(network_223$nextds_no)):1)
network_223_i <- apply(network_223[,c("site_no","nextds_no")], 2, as.character)
network_223_ig <- igraph::graph_from_data_frame(network_223_i)
igraph::E(network_223_ig)$weight <- network_223$dist_km

# Downstream river distance matrix
dist_ds_km_223 <- igraph::distances(network_223_ig, 
                                  mode = "out",
                                 weights = network_223$dist_km)
# Remove the terminal sites 
dist_ds_km_223 <- as.matrix(dist_ds_km_223)[1:223,1:223]
# dist_ds_scaled <- dist_ds_km_223
# dist_ds_scaled[dist_ds_scaled == Inf] <- 1000
# # scale() scales by column, which is not what we want to do here.  
# # We need to vectorize, scale and return to matrix
# x <- as.vector(dist_ds_scaled)
# x <- scale(x, center = FALSE)
# dist_ds_scaled <- matrix(x, ncol = ncol(dist_ds_scaled))
dist_ds_inv_w <- exp(-dist_ds_km_223)
# set diagonal to zero as in original stan model
for(i in 1:nrow(dist_ds_inv_w)){
  dist_ds_inv_w[i,i] <- 0
}
# Inspect and assess need for scaling
# range(dist_ds_inv_w) #  0 to 0.88 # no need for scaling

# Downstream river distance matrix
dist_us_km_223 <- igraph::distances(network_223_ig, 
                                  mode = "in",
                                 weights = network_223$dist_km)
# Remove the terminal sites 
dist_us_km_223 <- as.matrix(dist_us_km_223)[1:223,1:223]
dist_us_inv_w <- exp(-dist_us_km_223)
# set diagonal to zero as in original stan model
for(i in 1:nrow(dist_us_inv_w)){
  dist_us_inv_w[i,i] <- 0
}
# Inspect and assess need for scaling
# range(dist_us_inv_w) #  0 to 0.88 # no need for scaling

# sum(dist_ds_km_223 == Inf)/length(dist_ds_km_223) # 98.1% = Inf (i.e. independent)
# sum(dist_us_km_223 == Inf)/length(dist_us_km_223) # same

```

### Compile environmental predictors

```{r}

### --- Sample metadata variables

samples_db <- samples
samples <- samples[c("smpcode","date","sitecode_v12")]
samples$reach_v12 <- substr(samples$sitecode_v12,1,nchar(samples$sitecode_v12)-1)
samples$site_no <- sites$site_no[match(samples$sitecode_v12, sites$sitecode_v12)]
# Retrieve environmental data from mwstr database
samples$season <- 0; samples$season[lubridate::month(samples$date) > 6] <- 1 # Season - 0 = Autumn; 1 = Spring
# regex to identify Riffle habitat in collection_method
samples$riffle <- 0; samples$riffle[grep("riffle", samples_db$collection_method)] <- 1 # Riffle habitat

### --- Forest cover

# AF calculated for sites rather than reaches and stored in mwbugs database
sites_af <- DBI::dbGetQuery(mwbugs, paste0("SELECT sitecode_v12, af_l5_w633_1750, ",
                                           "af_l5_w633_2018, af_l5_w633_2022 ",
                                           "FROM sitecode_env WHERE sitecode_v12 IN ('",
                                           paste(sites$sitecode_v12, collapse = "', '"),"');"))
# Check sites are in sites_af: sum(!sites$sitecode_v12 %in% sites_af$sitecode_v12)  # zero
# sitecode_env table in database was cleaned up 16 May 2026 and all sitecodes match now.

# Join environmental data to sites
# A major storm on 9 June 2021 caused much forest loss https://www.theguardian.com/australia-news/2021/jun/24/dandenong-ranges-power-outage-storm-debris-victoria-emergency-services-warning
# therefore use af_2018 before this date and af_2022 after.
samples$af <- sites_af$af_l5_w633_2022[match(samples$sitecode_v12, sites_af$sitecode_v12)]
samples$af[samples$date < "2021-06-10"] <- sites_af$af_l5_w633_2018[match(samples$sitecode_v12[samples$date < "2021-06-10"],
                         sites_af$sitecode_v12)]
samples$af_ref <- sites_af$af_l5_w633_1750[match(samples$sitecode_v12, sites_af$sitecode_v12)] # reference state for AF

### --- Impervious cover and other environmental variables

# Other variables can be adequately estimated from reach values.
cat_env <- DBI::dbGetQuery(mwstr,paste0("SELECT subcs.reach, subcs.carea_km2, ei_l44_sw1942_2022, c_basalt, ", 
                                         "ti_2022, tf_2018, tf_2022, c_granite, meanq_mm, dor_2010 FROM cat_env ",
                                         "JOIN subcs ON cat_env.site = subcs.site ",
                                         "WHERE cat_env.reach IN ('",
                                         paste(sites$reach_v12, collapse = "', '"), "');"))
subc_env <- DBI::dbGetQuery(mwstr,paste0("SELECT reach, meant_y30_2022, elev_asl_m FROM subc_env WHERE reach IN ('",
                                         paste(sites$reach_v12, collapse = "', '"), "');"))

samples$carea_km2 <- cat_env$carea_km2[match(samples$reach_v12, cat_env$reach)] # Catchment area
samples$dor <- cat_env$dor_2010[match(samples$reach_v12,cat_env$reach)] # Degree of regulation
samples$ei <- cat_env$ei_l44_sw1942_2022[match(samples$reach_v12,cat_env$reach)] # Effective imperviousness
samples$ti <- cat_env$ti_2022[match(samples$reach_v12,cat_env$reach)] # Total imperviousness
samples$tf <- cat_env$tf_2022[match(samples$reach_v12,cat_env$reach)] # Total forest
samples$tf[samples$date < "2021-06-10"] <- 
  cat_env$tf_2018[match(samples$reach_v12[samples$date < "2021-06-10"],
                         cat_env$reach)] # Adjust for storm-related forest loss
samples$meanq <- cat_env$meanq_mm[match(samples$reach_v12,cat_env$reach)] # Mean annual runoff
samples$meant <- subc_env$meant_y30_2022[match(samples$reach_v12,subc_env$reach)] # Mean annual temperature
samples$elev <- subc_env$elev_asl_m[match(samples$reach_v12,subc_env$reach)] # Elevation above sea level
samples$coord_X <- sf::st_coordinates(sites[match(samples$sitecode_v12, sites$sitecode_v12),])[,1]
samples$coord_Y <- sf::st_coordinates(sites[match(samples$sitecode_v12, sites$sitecode_v12),])[,2]

samples <- samples[order(samples$site_no, samples$date),] # Order by site and season
sites_for_xl <- sites
sf::st_geometry(sites_for_xl) <- NULL

# Disconnect from databases
DBI::dbDisconnect(mwbugs)
DBI::dbDisconnect(mwstr)

```

## Exploratory analysis of outliers and multicollinearity among predictor variables.

Candidate predictors: "season","riffle","af","ei,"carea_km2","dor","meanq","meant"

Check also correlations between ti and ei, tf and af, meant and elev

```{r}

plot(samples$af, samples$tf)  
  # cor(samples$af, samples$tf) # = 0.83 (so one or other in a model)
 # wilkins et al 2026 showed af to be more optimal
# ei is dominated by low values, but there are fewer outliers (with potentially high leverage)
#  if ei is log(x + 0.01)-transformed 
# (previous studies used log10(x + 0.001)) 
# this is probably better because values < 0.01 (below which stormwater impacts should be small) are stretched out less
plot(log(samples$ei + 0.01), log(samples$ti+ 0.01))  
  # cor(log(samples$ei + 0.01), log10(samples$ti+ 0.01)) # = 0.95 (so one or other in a model)
# wilkins et al 2026 showed ei to be more optimal
samples$lei <- log(samples$ei + 0.01)

# elevation is positively skewed. No outliers if square-root transformed
# and temperature is negatively skewed by not large outliers that would have potential high leverage
# hist(samples$elev^0.5), hist(samples$meant)
plot(samples$elev^0.5, samples$meant)  
# cor(samples$elev^0.5, samples$meant)  # = 0.90 (one or other in a model)
# (from Walsh 2025) air temperature provides a good measure of stream temperature 
# in the absence of anthropogenic alterations to effects such as tree cover, 
# stormwater drainage and flow modification. It also has the advantage of being 
# compatible with a measure by which climate-warming predictions are communicated.

# Catchment area is highly skewed.  A log-transform is required to normalize it.
# hist(log(samples$carea_km2))  # hist(samples$carea_km2^0.25)
samples$lcarea <- log(samples$carea_km2)

# meanq will be modelled as a quadratic to allow for a non-linear response to meanq
# hist(scale(samples$meanq^0.5) - scale(samples$meanq))
# has no extreme outliers, while scale(samples$meanq) - scale(samples$meanq^0.5)
# (Both produce a model response as required)
samples$smeanq <- samples$meanq^0.5
# To check how scaling affects the distribution of the resulting modal curve:
# plot(samples$smeanq, scale(samples$meanq^0.5) - scale(samples$meanq))
# plot(samples$smeanq, scale(samples$meanq)^0.5 - scale(samples$meanq))
# plot(samples$smeanq, scale(samples$smeanq) - scale(samples$smeanq^2))
# plot(samples$smeanq, scale(samples$smeanq) - scale(samples$smeanq)^2)
# This suggests it is better to independently scale the smeanq and smeanq^2

# dor of most sites ranges between 0 and 1. Eight have dor > 1, including 2 high outliers
# In calculating dor Walsh (2025) capped dor below large service reservoirs, which never spill
# (but release small environmental flows) at 1. We thus do the same for smaller reservoirs with
# dor > 1 to minimise leverage of sites below such reservoirs, while retaining the scale of 
# lower dor sites
# 
# hist(pmin(samples$dor,1))
samples$dor1 <- pmin(samples$dor,1)

pairs(samples[c("af","lei","lcarea","smeanq","dor1")],upper.panel = panel.cor) 
#"season","riffle" are spread evenly across the other variables (although no riffle = 0 at high meanq or dor)
# low correlation between main environmental effects.
# highest 0.69 for af and meanq; and 0.68 for meanq and ldor, both acceptable 

# Potential interactions to include in the model
# To avoid conflation arising from scaling around zero, 
# interactions should be calculated on unscaled predictors and then scaled
# hist(scale(samples$smeanq * samples$af))  # well-behaved
# hist(scale(samples$smeanq * samples$lei)) # well-behaved
# hist(scale(log(samples$smeanq * samples$dor1 + 0.1))) #highly skewed if not log-transformed
# hist(scale(samples$af * samples$lei)) # high (>1.5) low af, high ei; low (<1.25) high af, low ei

samples$q_af <- samples$smeanq * samples$af
samples$q_ei <- samples$smeanq * samples$lei
samples$q_d <- log(samples$smeanq * samples$dor1 + 0.1)
samples$af_ei <- samples$af * samples$lei

pairs(samples[c("af","lei","lcarea","smeanq","dor1","q_af","q_ei","q_d","af_ei")],upper.panel = panel.cor) 

# all interactions introduce high correlations with main effects, but perhaps this
# doesn't matter, and they will acti similarly to the quadratic term for smeanq?

```

## Subset of biota for trial analyses

Begin with a tractable small set of species of varying taxonomic relationships and potential response.  Choose species with moderate prevalance (say between 25 and 75%)

```{r}

# Note range(biota$count) = 1,1
# calculate prevalence on sites rather than samples
biota$reach <- samples$reach_v12[match(biota$smpcode, samples$smpcode)]
biot_by_site <- unique(biota[c("taxoncode","taxon","reach")])
biot_prev <- aggregate(biot_by_site$taxoncode, 
                       by = list(taxoncode = biot_by_site$taxoncode, taxon = biot_by_site$taxon), 
                       FUN = length)
biot_prev$prevalence <- biot_prev$x/length(unique(samples$reach_v12))
biot_prev <- biot_prev[biot_prev$prevalence >= 0.2 & biot_prev$prevalence <= 0.8,]  # 111 species
# about 1 quarter of that might be a good number to start with
# But first quarantine a few congenerics to make the phylogenetic similarity diverse
spp_trial <- biot_prev[grepl("QT2505",biot_prev$taxoncode) | 
                       grepl("QT2511",biot_prev$taxoncode) |
                       grepl("QE0609",biot_prev$taxoncode),]
biot_prev <- biot_prev[!biot_prev$taxoncode %in% spp_trial$taxoncode,] # down to 105
spp_trial <- rbind(spp_trial, biot_prev[seq(4,104,4),])  #33

biota <- biota[biota$taxoncode %in% spp_trial$taxoncode,]  #down to 2882 records

```


## 2. Analysis in Stan

### 2.1. Assembling model data for Stan

The following code chunk assembles the model data for input to the Stan model `multitaxon_pa_rSAC.stan`.

```{r}
#| label: stan_data
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

# Load data from excel file
# data_list <- rio::import_list("temp_data/mw_metabarcoding_225site_model_data.xlsx")
# 
# samples <- data_list$samples
# biota <- data_list$biota

### Preparing predictor data

# scale fixed predictor variables
# keeping them as scaled objects for ease of back-transforming later
I <- scale(samples$lei)
F <- scale(samples$af)
F2 <- scale(samples$af^2)
Q <- scale(samples$smeanq)
Q2 <- scale(samples$smeanq^2)
C <- scale(samples$lcarea)
D <- scale(samples$dor1)
T <- scale(samples$meant)
T2 <- scale(samples$meant^2)
Q_F <- scale(samples$q_af)
Q_I <- scale(samples$q_ei)
Q_D <- scale(samples$q_d)
F_I <- scale(samples$af_ei)
# I don't think it is necessary to scale these two binary variables
nriff <- samples$riffle
season <- samples$season

# Random effects
site <- samples$site_no
 # I don't think site_group is necessary with the distance matrix. For discussion
# site_group <- samples$site_group

# # Coordinates
# coords <- unique(samples[,c("siteno", "coord_X", "coord_Y")])
# coords_XY <- cbind(coords$coord_X, coords$coord_Y)
# distm_XY <- scale(stats::dist(coords_XY), center = FALSE)
# 
### Biota presence/absence, count and subsample percentage data
biota_ct <- with(biota, ct(smpcode, shortcode, count))
biota_ct <- biota_ct[match(samples$smpcode,row.names(biota_ct)),] # ensure c and u rows are ordered the same way

### Predictor matrix (u in the model)
# Most plausible model did not include interactions...perhaps this was because of the scale-confounding problem???
# But equally there is a multicollinearity problem with the interactions....re-assess if they help or not.

Q_F <- scale(samples$q_af)
Q_I <- scale(samples$q_ei)
Q_D <- scale(samples$q_d)
F_I <- scale(samples$af_ei)
# I don't think it is necessary to scale these two binary variables
nriff <- samples$riffle
season <- samples$season

# --- Predictor matrix u for morph samples
I_mod <- as.vector(I)[match(samples$smpcode,row.names(biota_ct))]
F_mod <- as.vector(F)[match(samples$smpcode,row.names(biota_ct))]
F2_mod <- as.vector(F2)[match(samples$smpcode,row.names(biota_ct))]
Q_mod <- as.vector(Q)[match(samples$smpcode,row.names(biota_ct))]
Q2_mod <- as.vector(Q2)[match(samples$smpcode,row.names(biota_ct))]
C_mod <- as.vector(C)[match(samples$smpcode,row.names(biota_ct))]
D_mod <- as.vector(D)[match(samples$smpcode,row.names(biota_ct))]
T_mod <- as.vector(T)[match(samples$smpcode,row.names(biota_ct))]
T2_mod <- as.vector(T2)[match(samples$smpcode,row.names(biota_ct))]
Q_F_mod <- as.vector(Q_F)[match(samples$smpcode,row.names(biota_ct))]
Q_I_mod <- as.vector(Q_I)[match(samples$smpcode,row.names(biota_ct))]
Q_D_mod <- as.vector(Q_D)[match(samples$smpcode,row.names(biota_ct))]
F_I_mod <- as.vector(F_I)[match(samples$smpcode,row.names(biota_ct))]
nriff_mod <- as.vector(nriff)[match(samples$smpcode,row.names(biota_ct))]
season_mod <- as.vector(season)[match(samples$smpcode,row.names(biota_ct))]

samples_u <- cbind(I = I_mod,                   # Effective imperviousness
                   F = F_mod,                   # Attenuated Forest cover
                   Q = Q_mod,                   # mean annual runoff
                   Q2 = Q2_mod,                 # square of Q for modal response
                   C = C_mod,                   # catchment area
                   D = D_mod,                   # Degree of regulation
                   T = T_mod,                   # Temperature  
                   nriff = nriff_mod,           # number of riffles at site (0 or 1)
                   season_mod,                  # Autumn = 0, Spring = 1
                   FI = F_I_mod,                # Interaction betwen I and F
                   QF = Q_F_mod,                # Interaction betwen Q and F
                   Q_I = Q_I_mod,               # Interaction between Q and I          
                   Q_D = Q_D_mod)               # Interaction between Q and D

samples_u <- samples_u[match(samples$smpcode,row.names(biota_ct)),] # ensure c and u rows are ordered the same way

save(samples_u, I, F, F2, Q, Q2, C, D, T, T2, Q_F, Q_I, Q_D, F_I,
     sites, samples, biota, biota_ct, dist_euc_inv_w,dist_ds_inv_w,dist_us_inv_w,
     file = "~/uomShare/wergStaff/ChrisW/temp/223_site_model_data.rda")

samples_u_1 <- cbind(I = I_mod,                   # Effective imperviousness
                   F = F_mod,                   # Attenuated Forest cover
                   F2 = F2_mod,                   # Attenuated Forest cover
                   Q = Q_mod,                   # mean annual runoff
                   Q2 = Q2_mod,                 # square of Q for modal response
                   C = C_mod,                   # catchment area
                   D = D_mod,                   # Degree of regulation
                   T = T_mod,                   # Temperature  
                   T2 = T2_mod,                   # Temperature  
                   nriff = nriff_mod,           # number of riffles at site (0 or 1)
                   season_mod,                  # Autumn = 0, Spring = 1
#                   FI = F_I_mod,                # Interaction betwen I and F
                   QF = Q_F_mod,                # Interaction betwen Q and F
                   Q_I = Q_I_mod)               # Interaction between Q and I          
#                   Q_D = Q_D_mod)               # Interaction between Q and D

samples_u_1 <- samples_u_1[match(samples$smpcode,row.names(biota_ct)),] # ensure c and u rows are ordered the same way

save(samples_u_1, I, F, F2, Q, Q2, C, D, T, T2, Q_F, Q_I, Q_D, F_I,
     sites, samples, biota, biota_ct, dist_euc_inv_w,dist_ds_inv_w,dist_us_inv_w,
     file = "~/uomShare/wergStaff/ChrisW/temp/223_site_model_data_1.rda")

### Assemble stan data:
biota_223site_data <- list(n_obs = nrow(biota_ct),                 # no. samples
                    n_pred = ncol(samples_u_1),                     # no. predictors
                    n_taxa = ncol(biota_ct),                       # no. taxa
                    n_site = as.integer(max(samples$site_no)),     # no. sites
                    u = as.matrix(samples_u_1),               # predictor model matrix
                    y = as.matrix(biota_ct),
                    site = site,                            # site number for each obs
#                    M_euc = as.matrix(dist_euc_inv_w),                            # site number for each obs
                    M = as.matrix(dist_euc_inv_w) )   #t(dist_ds_inv_w)
# I think t(dist_ds_inv_w) is correct, given that the model applies W to sites by row.
# ANS-1611 (site 4) is downstream of ANS-943 (site 5).  
# row 4 of t(dist_ds_inv_w) has a non-zero value at col 5 (thus site 4 has a non-zero W for site 5)
# row 4 of dist_ds_inv_w does not.
# (This doesn't matter for dist_euc_inv_w, because it is symmetric)

model_fit <- rstan::stan(file = "stancode/multitaxon_pa_rSAC_cw.stan", 
                         data = biota_223site_data, 
               #          control = list(adapt_delta = 0.9), 
                         seed = proj_seed, 
                         chains = 1,
                         warmup = 1000,
                         iter = 2000)
save(model_fit, file = "~/uomShare/wergStaff/ChrisW/temp/model_run_1_dist_euc_gamma_F2_T2.rda")
# tds 60 min
# run1_dist_euc and run1_dist_ds each took ~45 min (and maybe <10G RAM)
# multitaxon_pa_mult_rSAC_cw.stan did not work 100% divergent transitions max Rhat NA.
```

```{r}

# Write standata file to WERG drive:
saveRDS(biota_223site_data, file = "~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/223site_standata.rds")

# Store files locally (e.g., for use on SPARTAN):
saveRDS(biota_223site_data, file = "temp_data/223site_standata.rds")

```

### 2.2. Creating directories for model outputs

Normally when running a Stan model, the model outputs are written to a directory on the WERG shared drive, as this drive has sufficient space to store large model outputs and is accessible from multiple computers. However, to run the model on a high-performance computing cluster (such as SPARTAN), it is much easier to set up a local directory, then push the entire local directory to SPARTAN, run the model, then pull the entire directory back to the local computer for analysis.

The following code chunk creates directories for storing model outputs both on the WERG shared drive and locally.

```{r}
#| label: create_output_dirs
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

# Create directory for model outputs on WERG shared drive
## Create local directory for each model set:
system(command = "mkdir -p 225-site_model_objects/")
system(command = "mkdir -p 225-site_model_objects/draws/")
system(command = "mkdir -p 225-site_model_objects/diagnostics/")

```

SPARTAN operates by running batch scripts on a remote server. The batch scripts are run through slurm, a workload manager for Linux clusters. The following is an example slurm batch script that can be used to run the Stan model on SPARTAN (saved as `225site.sbatch`):

```{r}
#| label: create_sbatch_script
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

sbatch_contents <- "#!/bin/bash
#SBATCH -p sapphire                     # Select partition (high CPU)
#SBATCH --job-name=225_site             # Job name
#SBATCH --ntasks=1                      # Number of tasks
#SBATCH --cpus-per-task=4               # Number of CPU cores per task - 4 required for Stan sampler
#SBATCH --mem=500000                    # Memory per node
#SBATCH --time=24:00:00                 # Time limit

# Load the R module
module load foss/2022a R/4.2.2

# Run the R script corresponding to the array job ID
Rscript 225-site_model_objects/225site_model.R


##DO NOT ADD/EDIT BEYOND THIS LINE##
##Job monitor command to list the resource usage
my-job-stats -a -n -s
"

# Write it to a file
writeLines(sbatch_contents, "225-site_model_objects/225site.sbatch")

```

### 2.3. Stan model script

The following is an example R script that can be used to run the Stan model on SPARTAN (saved as `225site_model.R`):

```{r}
#| label: create_stan_rscript 
#| echo: true 
#| message: false 
#| warning: false 
#| results: "hide"

# load stan data

# standata <- readRDS("~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/biota_225site_standata.rds")

standata <- readRDS("temp_data/biota_225site_standata.rds") # local version

# Stop code from accidentally running on local computer

stop() # unblock on SPARTAN or water-prod

library(StanHeaders, lib="~/R_libs/4.2.2") 
library(rstan, lib="~/R_libs/4.2.2") 
source("functions/rstan_functions.R")

# --- Set project seed

proj_seed <- 12345 
set.seed(proj_seed)

# --- Set cores and load stan model:

# rstan_options(auto_write = TRUE)

options(mc.cores = parallel::detectCores()) 
print("Sampling Stan model...") 
model_fit <- rstan::stan(file = "stancode/multitaxon_pa_rSAC_noSS.stan", 
                         data = standata, 
                         control = list(adapt_delta = 0.9), 
                         seed = proj_seed, 
                         chains = 4, 
                         iter = 4000)

print("Model finished sampling.") 
# --- Save model outputs to WERG shared drive: 
print("Saving model draws...") 
draws <- rstan::extract(model_fit, permuted = TRUE) 
saveRDS(draws, file = "225-site_model_objects/draws/draws.rds") 
saveRDS(draws$y_pred, file = "225-site_model_objects/draws/y_pred.rds")
saveRDS(draws$mu, file = "225-site_model_objects/draws/mu_draws.rds") print("Draws saved.")

# --- Generate diagnostics report

print("Generating diagnostics report...") 
mon <- as.data.frame(monitor(model_fit, print = FALSE)) 
write.csv(mon, file = "225-site_model_objects/diagnostics/mon.csv") 
report <- convergence_report(model_fit, mon) 
print_convergence_report(report)

```
