---
title: "Selection of optimal predictor set using 200-species model"
author: "Christopher J Walsh"
format: docx
editor: source
bibliography: references.bib
editor_options: 
  chunk_output_type: inline
---

```{r}
#| label: setup
#| echo: false

source("quick_196_spp_data_assembly_for_ou_model.R")
# This creates data ready for the model for constrained_z
source("cmdstanr_functions.R")
# post processing functions

```

The run_200spp_model() function (loaded in "quick_200_spp_data_assembly_for_latent_tests.R") runs the specified 200-species model and creates a list of objects for post-processing ("species_set", "sample_set", "pred_set", "data",       "fit", "diagnostics", "summary"). 

```{r}
#| eval: false

run_196spp_model(pred_site = pred_site,
                 mod_code = "ifqcdtblrs_fi_qf_qi_ou",
                 mod_path = "stancode/jointspp_pa_no_eps_ou.stan", 
                 calc_loo = TRUE)  
# The full model was used first to validate that all was in order

```

The need for post-processing is illustrated in @fig-alignment, showing the raw traceplots of the four chains for latent axes 1 and 2 for site 1 (A. z_expanded[1,1], B. z_expanded[1,2]) from the Stan model. They show that the score estimates for each chain drifting independently over the 1200 iterations.  This is because the model does not constrain the ordination to a set rotation (and attempting to do so produced stable, but split chains that did not converge).  The following code chunk applies a Procrustes rotation to the ordination in each iteration, which results in well-mixed, stable traceplots for the aligned z-values (C and D).

```{r}
#| label: u_sets

# Core predictors
u_core <- c("I","F","Q","Q2","C","T")   # and "riff","season" in u_obs, which will remain unchanged
u_opts <- c("D","B","L")
u_int_opts <- c("F_I","Q_F","Q_I")  # and maybe (if B or D are important, Q_D and B_I)

# so pred combinations to trial are:
pred_set <- list(u_core, c(u_core, "D"), c(u_core, "B"), c(u_core, "L"), 
                  c(u_core, "D","B"), c(u_core, "D","L"), c(u_core, "B","L"), 
                  c(u_core, "D","B","L"))

names(pred_set) <- c("ifqct", "ifqctd", "ifqctb", "ifqctl", 
                     "ifqctdb", "ifqctdl", "ifqctbl", "ifqctdbl")

for(i in 1:length(pred_set)){
run_196spp_model(pred_site = pred_set[[i]],
                 mod_code = names(pred_set)[i],
                 mod_path = "stancode/jointspp_pa_no_eps_ou.stan", 
                 calc_summary = FALSE, calc_loo = TRUE)  
  gc()
}
```

The following code aligns the latent factor parameters z_expanded and lambda.

```{r}

mod_dir <- "~/uomShare/wergStaff/ChrisW/git-data/mw_metabarcoding_model/trial_200spp_models/"
mod_bundle <- load_model_component_bundle(mod_code ="ifqcdtblrs_fi_qf_qi_ou", mod_dir = mod_dir)
mod_fit <-  mod_bundle$fit
mod_data <- mod_bundle$data

# Ensure your dimensions are correctly defined for your active model
n_obs <- mod_data$n_obs
n_site <- mod_data$n_site
n_taxa <- mod_data$n_taxa
n_latent <- mod_data$n_latent
n_chains <- mod_fit$num_chains()
n_iterations <- mod_fit$metadata()$iter_sampling

# dimensions  [1200 iterations, 4 chains, 550 flattened parameters]
raw_z <- mod_fit$draws("z_expanded", format = "array")
raw_lambda <- mod_fit$draws("lambda_raw", format = "array")

# Establish a baseline template using a stable iteration 
# example_z <- mod_fit$draws("z_expanded[1,1]", format = "array")
# base_traceplot(example_z, "z_expanded[1,1]") # all respectable. Pick chain 1
# Save example unaligned (and later aligned) z scores for plotting above.
# examp_draws <- list(z11_unaligned = example_z, 
#                     z12_unaligned = mod_fit$draws("z_expanded[1,2]", format = "array"))
first_vector <- raw_z[500, 4, ]

target_matrix <- matrix(NA, nrow = n_obs, ncol = n_latent)
target_matrix[, 1] <- first_vector[1:n_obs]           # Elements 1 to 275 = Axis 1
target_matrix[, 2] <- first_vector[(n_obs + 1):550]    # Elements 276 to 550 = Axis 2

# Standardise the master frame
target_matrix <- scale(target_matrix) 

# Create empty, 4D arrays to store the aligned results
# dim =iterations, chains, observations, latent axes
z_unpacked_aligned <- array(NA, dim = c(n_iterations, n_chains, n_obs, n_latent))
lambda_unpacked_aligned <- array(NA, dim = c(n_iterations, n_chains, n_taxa, n_latent))

for (c in 1:n_chains) {
  for (i in 1:n_iterations) {
    # current 550-element vector
    current_vector <- raw_z[i, c, ]
    # Create an empty  275 x 2 matrix to be populated
    current_matrix <- matrix(NA, nrow = n_obs, ncol = n_latent)
    current_matrix[, 1] <- current_vector[1:n_obs]            # Dimension 1 (Axis 1)
    current_matrix[, 2] <- current_vector[(n_obs + 1):550]    # Dimension 2 (Axis 2)
    # Standardise to remove cross-chain scale discrepancies
    current_matrix_scaled <- scale(current_matrix)
    if (any(is.na(current_matrix_scaled)) || any(is.infinite(current_matrix_scaled))) {
      next 
    }
    # Rotate the iteration onto the master template: dimensions [275,2]
    rotation <- vegan::procrustes(X = target_matrix, Y = current_matrix_scaled, symmetric = FALSE)
    z_unpacked_aligned[i, c, , ] <- rotation$Yrot
    # Extract the 2 x 2 rotation matrix calculated for this iteration for aligning the species loadings (lambda)
    rotation_matrix <- rotation$rotation
    current_lambda_vector <- raw_lambda[i, c, ]
    current_lambda_matrix <- matrix(current_lambda_vector, nrow = n_taxa, ncol = n_latent)
    # multiply lambda by the same rotation matrix
    lambda_unpacked_aligned[i, c, , ] <- current_lambda_matrix %*% rotation_matrix
  }
}
# Add to data for example plot above
# examp_draws$z1_aligned <- z_unpacked_aligned[,,161,]
# save(examp_draws, file = "~/uomShare/wergStaff/ChrisW/git-data/mw_metabarcoding_model/examp_draws.rda")

####
# 1. Recalculate draws for aligned Z (latent factor scores for sites)
# z_expanded values are identical for observations from the same sites, so just use one value per site
unique_site_ids <- 1:n_site
first_obs_indices <- match(unique_site_ids, mod_data$site)
all_sites_samples_raw <- z_unpacked_aligned[, , first_obs_indices, ] # dim [1200, 4, 223, 2]
# Reshape from a 4-D array into a standard 3-D MCMC array: [1200, 4, 446]
n_parameters <- n_site * n_latent
all_sites_samples_flat <- array(all_sites_samples_raw, dim = c(n_iterations, n_chains, n_parameters))
parameter_names <- character(n_parameters)
idx <- 1
for (s in unique_site_ids) {
  parameter_names[idx]     <- paste0("Site_", s, "_Z1")
  parameter_names[idx + 1] <- paste0("Site_", s, "_Z2")
  idx <- idx + 2
}# Apply the naming dimensions to the reshaped array
dimnames(all_sites_samples_flat) <- list(
  Iteration = 1:n_iterations,
  Chain     = 1:n_chains,
  Parameter = parameter_names
)
# The formal draws object 
all_sites_aligned_draws <- as_draws_array(all_sites_samples_flat)

# 2. Recalculate draws for aligned lambda (latent factor scores for species)
n_lambda_params <- n_taxa * n_latent
lambda_flat <- array(lambda_unpacked_aligned, dim = c(n_iterations, n_chains, n_lambda_params))
species_names <- colnames(mod_data$y)
lambda_names <- character(n_lambda_params)
idx <- 1
for (j in 1:n_taxa) {
  lambda_names[idx]     <- paste0("Species_", species_names[j], "_Load_Axis1")
  lambda_names[idx + 1] <- paste0("Species_", species_names[j], "_Load_Axis2")
  idx <- idx + 2
}
dimnames(lambda_flat) <- list(
  Iteration = 1:n_iterations,
  Chain     = 1:n_chains,
  Parameter = lambda_names
)
# Cast into a formal MCMC draws object
lambda_aligned_draws <- as_draws_array(lambda_flat)
```

From these aligned z and lambda draws, we revised the diagnostic summary for the model using the following code to produce @tbl_diags, which shows the model to be in good order. In my first re-run of the model (with tightened sigma_taxon prior and corrected data assembly), the lowest ESS was for one of the latent factor site values: not terribly low at 305. Because the non-latent factor variables were all well above 400 (@tbl-diags), this model is adequate for model comparisons.

```{r}
#| label: "tbl-diags"
#| echo: false

lambda_diagnostic_summary <- summarise_draws(lambda_aligned_draws)
z_diagnostic_summary <- summarise_draws(all_sites_aligned_draws)

mod_summary <- mod_bundle$summary[-grep("z_exp|lambda",mod_bundle$summary$variable),]
mod_summary <- rbind(mod_summary, lambda_diagnostic_summary,z_diagnostic_summary)
mod_summary$variable[which(mod_summary$ess_bulk == min(mod_summary$ess_bulk, na.rm = TRUE))]
diag_summary <- data.frame(spec = c("No. sampling iterations","No. divergences",
                                    "No. max_treedepth exceedences",
                                    "BFMI (Chain 1)","BFMI (Chain 2)","BFMI (Chain 3)","BFMI (Chain 4)",
                                    "Min. Bulk ESS", "Min. Tail ESS","Max R-hat",
                                    "Worst sampled parameter"),
                           value = c(n_iterations,
                                     sum(mod_bundle$diagnostics$num_divergent),
                                     sum(mod_bundle$diagnostics$num_max_treedepth),
                                     round(mod_bundle$diagnostics$ebfmi,2),
                                     round(min(mod_summary$ess_bulk, na.rm = TRUE)),
                                     round(min(mod_summary$ess_tail, na.rm = TRUE)),
                                     round(max(mod_summary$rhat, na.rm = TRUE),2),
                                     mod_summary$variable[which(mod_summary$ess_bulk == min(mod_summary$ess_bulk, na.rm = TRUE))])
                           )

if(grepl("Z|lambda", diag_summary$value[diag_summary$spec == "Worst sampled parameter"])){
  non_latent <- mod_summary[-grep("Z|lambda", mod_summary$variable),]
  w1 <- non_latent[order(non_latent$ess_bulk),][1,]
  diag_summary <- rbind(diag_summary, data.frame(spec = c("Min. Bulk ESS (non-latent params)",
                                                          "Worst non-latent param"),
                                                 value = c(round(w1$ess_bulk), w1$variable)))
}
ft_to_word(flextable::flextable(diag_summary))

```

### Assessing the latent-factor ordination

```{r}
#| label: fig-latent_factor_ordination
#| echo: false
#| fig-width: 5
#| fig-height: 5
#| fig-cap: "Ordination of latent factors explaining random variation among sites (grey points), overlain by mean locations of species in the ordination space (red points)."

# Plot latent site ordination
mean_obs_z1 <- apply(z_unpacked_aligned[, , , 1], 3, mean, na.rm = TRUE)
mean_obs_z2 <- apply(z_unpacked_aligned[, , , 2], 3, mean, na.rm = TRUE)
ordination_df <- data.frame(Site = mod_data$site, 
  Z1   = mean_obs_z1,
  Z2   = mean_obs_z2
)
# Site scores are unique, so reduce to one value per site
unique_site_ordination <- unique(ordination_df)
plot(unique_site_ordination$Z1, unique_site_ordination$Z2,  pch = 21, bg = grey(0.8), 
     xlab = "Latent Axis 1", 
     ylab = "Latent Axis 2", asp = 1)
abline(v = 0, h = 0, lty = 2)

samples <- mod_bundle$sample_set
sites <- unique(samples[c("reach_v12","site_no")])

# Map species to the ordination - put z and lambda on the same scale first
species <- mod_bundle$species_set
mean_lambda1 <- apply(lambda_unpacked_aligned[, , , 1], 3, mean, na.rm = TRUE)
mean_lambda2 <- apply(lambda_unpacked_aligned[, , , 2], 3, mean, na.rm = TRUE)
max_site_range <- max(abs(c(unique_site_ordination$Z1, unique_site_ordination$Z2)))
max_spp_range  <- max(abs(c(mean_lambda1, mean_lambda2)))
# The scale factor adjusts species to occupy roughly 80% of the site space
scale_factor <- (max_site_range / max_spp_range) * 0.8
species_ord <- data.frame(species = species$taxon[match(colnames(mod_bundle$data$y),species$taxoncode)],
                          taxoncode = colnames(mod_bundle$data$y),
                          axis_1  = mean_lambda1 * scale_factor, 
                          axis_2  = mean_lambda2 * scale_factor)
points(species_ord$axis_1,species_ord$axis_2, pch = 16, cex = 0.5, col = "red")

```

*The following selection of points may not work without editing for future runs of the model, as the Procrustes rotation may well invert axes*
Most sites are grouped tightly around the origin of the ordination indicating little structure among sites.  Five sites ("BNY_24764", "GIS_1106",  "LOL_11925", "PRW_13767", and "REI_3934") lie to the right of the main mass of sites, with "LOL_11925" the largest outlier in the bottom right (@fig-latent_factor_ordination). These sites do not seem to have features in common.

The species associated with these sites:
                        species taxoncode   axis_1     axis_2
23  Bothrioneurum vejdovskyanum  LO040701 3.529873  0.2591227
94    Parakiefferiella sp. MC-1  QDAF03A1 3.012658 -1.1664422
109       Cricotopus albitarsis  QDAF1508 3.189559  0.6852920
126   Paratanytarsus kathleenae  QDAH0604 3.100098  0.5113616
                          species taxoncode       axis_1    axis_2
101 Thienemanniella sp. B-ACG8705  QDAF07AC  2.228228670 -2.441575
115         Botryocladius grapeth  QDAF3506  2.073556977 -2.160893
118      Orthocladiinae sp. MC-10  QDAF99AA -0.002373152 -3.078418
180             Taschorema evansi  QT011301  1.404124747 -2.971681
185          Asmicridea edwardsii  QT060201  0.826514597 -3.212064

Appear to be primarily urban/disturbance tolerant species. This requires further thought.

```{r}
#| eval: false

# # 5 sites with latent axis 1 scores > 2.5
# # used to manually check them first
# identify(unique_site_ordination$Z1, unique_site_ordination$Z2, unique_site_ordination$Site)
low_z2s <- unique_site_ordination[unique_site_ordination$Z2 < -3,]
sites[sites$site_no %in% low_z2s$Site,]  # LOL-11925
high_z1s <-  unique_site_ordination[unique_site_ordination$Z1 > 2.5,]
sites[sites$site_no %in% high_z1s$Site,]
# "BNY_24764" "GIS_1106"  "LOL_11925" "PRW_13767" "REI_3934" 
set_preds <- data.frame(reach_v12 = NA, mod_bundle$pred_set[high_z1s$Site,])
for(i in 1:nrow(high_z1s)){
  set_preds$reach_v12[i] <- sites$reach_v12[sites$site_no == high_z1s$Site[i]]
}

# Species associated with those sites
species_ord[species_ord$axis_1 >3,] # all classic disturbance tolerant species
species_ord[species_ord$axis_2 < -2.75 | (species_ord$axis_2 < -2 & species_ord$axis_1 > 2),] # A mix of moderately tolerant and sensitive species

# Conduct broader tests for any residual geographical information
coords <- sf::st_coordinates(sites)
dist_matrix <- as.matrix(dist(coords))
# Invert the distance matrix to create a spatial weights matrix (closer = heavier weight)
weights_matrix <- 1 / dist_matrix
diag(weights_matrix) <- 0 
weights_matrix[is.infinite(weights_matrix)] <- 0
# Moran's I to assess for any potential contagious distribution
moran_z1 <- ape::Moran.I(unique_site_ordination$Z1, weights_matrix)
moran_z2 <- ape::Moran.I(unique_site_ordination$Z2, weights_matrix)
# Both >0.5, which suggests randomness without spatial clumping.

# Regressions predicting Axis 1 and 2 from spatial location
Z <- cbind(unique_site_ordination, coords)
spatial_lm1 <- lm(Z1 ~ X * Y, data = Z)
spatial_lm2 <- lm(Z2 ~ X * Y, data = Z)
summary(spatial_lm1)$r.squared  
summary(spatial_lm2)$r.squared  
# Both ~ 0.01, which suggests no residual trend across Euclidean space

```

### Model predictive assessment

### AUC metrics

We calculated two Area-under-the-curve (AUC) metrics to assess model fit.  The area under the receiver-operating-curve (ROC-AUC) measures how well the model separates presences from absences. It calculates the probability of the model assigning a larger probability of occurrence to a randomly chosen site where the species is present than to one where it is absent.  ROC-AUC varies between 0 and 1, with a value of 0.5 meaning the model is no better than a random guess. Because ROC-AUC includes true negatives, it is inflated for rare species, whereby a model that predicts absences at all sites will score a high value by virtue of predicting the predominance of absences in the dataset.  

The area under the precision-recall curve (PR-AUC) is more a more useful metric for rare species as it only considers presences and measures mean ranking of site probabilities of sites at which the species is present. It varies between the species prevalence (meaning the model assigns the lowest probability to the sites at which the species was observed) to 1 (meaning the model assigns the highest probability to those sites).  

ROC-AUC shows that the full model, including latent factors, predicts presence-absence of all 200 species well (@fig-auc A). While for most species, the predictions using only the fixed environmental predictors generally reduced ROC-AUC (red points in @fig-auc), the prediction of presence-absence remained strong for all species.  PR-AUC similarly shows good predictions of presences for most species, except for some rare species (@fig-auc B).  @fig-auc C focuses on the rarest species (occurring in <7 sites) and shows that for all but a few species, the model's prediction is substantially better than a random guess.

@fig-auc C reveals 3 types of species:

- those for which PR-AUC is much higher when calculated using the full model (about half of the rare species, such as *Macrogyrus* sp. B-ABW3343 group and *Polypedilum nubifer*).  For these, the latent factors (likely unmeasured local variation) explain their distributions as much or more than the fixed environmental variables of the model.

- those for which PR-AUC is unchanged when the latent factors are removed (~10 species, such as *Plumatella fungosa* and *Fluviolanatus subtortus*). The distributions of these species are predominately predicted by the fixed environmental variables.

- those for which PR-AUC is lower when calculated using the full model (only a few, e.g. *Cricotopus* sp. B-ABX4071 and *Aulodrilus pigueti*). The distributions of these species are independent of the assemblage structure modelled by the latent factors, and are perhaps driven by random dispersal.


```{r}
#| label: fig-auc
#| echo: false
#| fig-width: 5
#| fig-height: 8
#| fig-cap: "A. ROC-AUC and B. PR-AUC for all 200 modelled species. Black points show AUC when probability of occurrence is modelled using latent-factor site variables. Red points are AUC calculated only using the fixed environmental predictors. C shows the two PR-AUC measures for the 39 species with low prevalence (below the gray horizontal line in A and B: present in <7 sites), with the two measures connected by a red horizontal line for clarity. The vertical line in A. shows the AUC score at which the model would be no better than chance.  The diagonal lines in B and C show the lowest possible PR-AUC value (equal to species prevalence)."

# p_total_array_raw <- mod_fit$draws("p_total_out", format = "array")
# p_env_array_raw   <- mod_fit$draws("p_env_out",   format = "array")
# mean_vector_total <- apply(p_total_array_raw, 3, mean)
# mean_vector_env   <- apply(p_env_array_raw,   3, mean)
# mean_p_total <- matrix(mean_vector_total, nrow = n_obs, ncol = n_taxa)
# mean_p_env   <- matrix(mean_vector_env,   nrow = n_obs, ncol = n_taxa)
# 
# # Arrays to hold the final AUC metrics for all species
# roc_auc_total <- numeric(n_taxa)
# pr_auc_total  <- numeric(n_taxa)
# roc_auc_env   <- numeric(n_taxa)
# pr_auc_env    <- numeric(n_taxa)
# # Calculate curve for each species
# for(j in 1:n_taxa) {
#   # Observed presence-absence
#   obs_y <- mod_data$y[, j] 
#   # Separate probabilities for presences and absences
#   pred_pres_total <- mean_p_total[obs_y == 1, j]
#   pred_abs_total  <- mean_p_total[obs_y == 0, j]
#   pred_pres_env   <- mean_p_env[obs_y == 1, j]
#   pred_abs_env    <- mean_p_env[obs_y == 0, j]
#   # Calculate the full-model curves (Environment + Latent factors)
#   fg_t <- pred_pres_total
#   bg_t <- pred_abs_total
#   roc_auc_total[j] <- PRROC::roc.curve(scores.class0 = fg_t, scores.class1 = bg_t)$auc
#   pr_auc_total[j]  <- PRROC::pr.curve(scores.class0 = fg_t, scores.class1 = bg_t)$auc.integral
#   # Calculate the pure environmental curves (U matrix only)
#   fg_e <- pred_pres_env; bg_e <- pred_abs_env
#   roc_auc_env[j] <- PRROC::roc.curve(scores.class0 = fg_e, scores.class1 = bg_e)$auc
#   pr_auc_env[j]  <- PRROC::pr.curve(scores.class0 = fg_e, scores.class1 = bg_e)$auc.integral
# }
# 
# # Combine into a gorgeous evaluation summary dataframe!
# auc_tab <- data.frame(
#   species = species_set$taxon[match(colnames(mod_data$y), species_set$taxoncode)],
#   taxoncode = colnames(mod_data$y),
#   prevalence = species_set$prevalence[match(colnames(mod_data$y), species_set$taxoncode)],
#   roc_auc_total = roc_auc_total,  pr_auc_total = pr_auc_total,
#   roc_auc_env = roc_auc_env,  pr_auc_env = pr_auc_env)
# save(auc_tab, file = "~/uomShare/wergStaff/ChrisW/temp/auc_tab.rda")

load("~/uomShare/wergStaff/ChrisW/temp/auc_tab.rda")
layout(matrix(c(1,2,3,4,4,4), c(2,3), byrow = TRUE), widths = c(2,12,12), heights = c(6,6))
par(mar = c(0,0,0,0)); plot.new(); title(ylab = "Prevalence by site (%)", line = -2)

par(mar = c(4,2,1,1))
auc_tab <- auc_tab[order(auc_tab$prevalence, decreasing = TRUE),]
with(auc_tab, plot(roc_auc_total, prevalence, xlim = c(0.5,1), xlab = "ROC-AUC", axes = FALSE))
with(auc_tab, points(roc_auc_env, prevalence, col = "red"))
abline(h = 6/223, lty = 1, col = "gray");
abline(v = 0.5, lty = 3); title(main = "A.", adj = 0)
axis(1); axis(2, las = 1, at = seq(0,0.8,0.1), labels = seq(0,80,10)); box(bty = "l")

with(auc_tab, plot(pr_auc_total, prevalence, xlim = c(min(prevalence),1), xlab = "PR-AUC", axes = FALSE))
with(auc_tab, points(pr_auc_env, prevalence, col = "red"))
with(auc_tab, lines(prevalence, prevalence, lty = 3)); title(main = "B.", adj = 0)
legend("topleft", pch = 1, col = c("black","red"), legend = c("Full model","Environmental predictors only"), cex = 0.9)
axis(1); axis(2, las = 1, at = seq(0,0.8,0.1), labels = seq(0,80,10)); box(bty = "l"); 
abline(h = 6/223, lty = 1, col = "gray");

low_prev_species <- auc_tab[auc_tab$prevalence <= 6/223,]
par(mar = c(4,12,1,1))
low_prev_species$index <- nrow(low_prev_species):1
with(low_prev_species, plot(pr_auc_total, index, xlim = c(0,1), 
                            axes = FALSE, xlab = "PR-AUC", ylab = ""))
with(low_prev_species, points(pr_auc_env, index, col = "red"))
for(i in 1:nrow(low_prev_species)){
  lines(c(low_prev_species$pr_auc_total[i],low_prev_species$pr_auc_env[i]),rep(low_prev_species$index[i],2), col = "red")
}
lines(low_prev_species$prevalence, low_prev_species$index, type = "s", lty = 3); title(main = "C.", adj = 0)
axis(1); axis(2, at = 1:39, labels = rev(low_prev_species$species), las = 1, cex.axis= 0.75); box(bty = "l")
```

#### Observed vs Predicted species richness

Predicted species richness per sample of the 200 species in the model was well predicted by the model (@fig-exp_richness - thank you, Tom - your tightening of the sigma_taxon prior did the trick!). Sample 384-UYT-199-4-DP is the one extreme low outlier with many more species observed than predicted.  This is an odd result, given the other 4 samples taken from this site were well predicted (the other samples from this site had observed, and predicted richness between 36 and 48).  This has me suspecting a data mix-up....I will look into that possibility some more.

There are 4 or 5 other samples for which the 95% CLs do not cross the 1:1 line. I will postpone investigating these further until the model predictors are nailed down.

```{r}
#| label: fig-exp_richness
#| echo: false
#| fig-width: 5
#| fig-height: 5
#| fig-cap: "Mean predicted species richness per sample with 95% confidence limits against observed richness.  The dotted line is 1:1."

# Extract simulated data
y_rep_array <- y_rep_matrix <- mod_fit$draws("y_rep", format = "matrix")
y_rep_matrix <- as.matrix(y_rep_matrix)
# 2. Rebuild the 3D array: 24000 draws, 200 species, 275 samples
# R fills arrays column-first, which perfectly matches how Stan flattens arrays:
# y_rep[1,1], y_rep[2,1], ..., y_rep[33,1], y_rep[1,2]...
y_rep_array <- array(y_rep_matrix, dim = c(4800, 275, 200))
expected_richness_draws <- apply(y_rep_array, c(1, 2), sum) # Dimensions: 24000 x 275

# Compute statistics column-wise for the 275 samples
richness_summary <- data.frame(
  smpcode = row.names(mod_data$y),
  sample_id     = 1:275,
  obs_richness = apply(mod_data$y, 1, FUN = sum),
  mean_richness = colMeans(expected_richness_draws),
  ci_lower      = apply(expected_richness_draws, 2, quantile, probs = 0.025),
  ci_upper      = apply(expected_richness_draws, 2, quantile, probs = 0.975),
  row.names     = NULL
)
# save(richness_summary, file = "~/uomShare/wergStaff/ChrisW/temp/exp_richness_200spp.rda")
# load("~/uomShare/wergStaff/ChrisW/temp/exp_richness_200spp.rda")

plot(richness_summary$obs_richness, richness_summary$mean_richness, 
     ylim = c(0,120), xlim = c(0,120), xlab = "Observed richness",
     ylab = "Predicted richness", axes = FALSE)
abline = c(0,1)
for(i in 1:nrow(richness_summary)){
  lines(rep(richness_summary$obs_richness[i],2),
    c(richness_summary$ci_lower[i], richness_summary$ci_upper[i]))
}
abline(0,1,lty = 3); axis(1); axis(2, las = 1); box(bty = "l")
# identify(richness_summary$obs_richness,richness_summary$mean_richness, richness_summary$smpcode)
# richness_summary[grep("UYT", richness_summary$smpcode),]

# # Alternative approach using predicted 0s and 1s
# simulated_presence <- array(
#   rbinom(length(y_rep_array), size = 1, prob = y_rep_array),
#   dim = dim(y_rep_array)
# )
# 
# # Sum presence/absence over species (dimension 2)
# sim_richness_draws <- apply(simulated_presence, c(1, 2), sum)
# 
# # Compute statistics
# sim_richness_summary <- data.frame(
#   sample_id     = 1:275,
#   obs_richness = apply(biota_ct, 1, FUN = sum),
#   mean_richness = colMeans(sim_richness_draws),
#   ci_lower      = apply(sim_richness_draws, 2, quantile, probs = 0.025),
#   ci_upper      = apply(sim_richness_draws, 2, quantile, probs = 0.975),
#   row.names     = NULL
# )
# 
# plot(sim_richness_summary$obs_richness, sim_richness_summary$mean_richness, 
#      ylim = range(c(sim_richness_summary$ci_upper, sim_richness_summary$ci_lower)))
# abline = c(0,1)
# for(i in 1:nrow(sim_richness_summary)){
#   lines(rep(sim_richness_summary$obs_richness[i],2),
#     c(sim_richness_summary$ci_lower[i], sim_richness_summary$ci_upper[i]))
# }
# abline(0,1,lty = 3)
# # # Extract the numeric index from the string (e.g., "3" from "y_rep[1,3]") and sort
# # sample_numbers <- as.numeric(gsub(".*,([0-9]+)\\]", "\\1", richness_summary$sample_name))
# # richness_summary <- richness_summary[order(sample_numbers), ]
# 
```

# Phylogenetic covariance

The OU model is the preferable model because it sampled more efficiently, and OU has a stronger theoretical basis of OU over Brownian [@sharma_etal_2025].  Mean $\alpha_{phylo}$ was 6.8 (95% confidence intervals 5.5,8.4).  From this, and the correlation values between species from the phylogenetic covariance matrix $\mathbf{\Sigma}_{\text{phylo}}$, we can find the predicted correlation between species sharing different taxonomic ranks (@fig-corr_by_rank).  

This suggests only a moderate correlation between species sharing the same genus or tribe, with little correlation between species sharing subfamilies or families, and near-zero correlation between species sharing orders. This suggests substantial evolutionary drift in niches within families suggesting substantial noise in models based on family-level taxonomy.

It remains possible that these correlation values will change when the model is applied to the full 883 species in the dataset.

```{r}
#| label: "fig-corr_by_rank"
#| fig-width: 5
#| fig-height: 5
#| fig-cap: "Predicted mean correlation (points) between species sharing a common taxonomic rank, with 80%  and 95% confidence limits (thick lines and think lines respectively"

spp_vcv_ou <- mod_data$phylo_cor
# # Congenerics (e.g Oecetis)  corr = 0.9114853
# spp_vcv_ou[grep("Oecetis", row.names(spp_vcv_ou)), grep("Oecetis", colnames(spp_vcv_ou))]
# # Same tribe corr (e.g Chironomini)= 0.8870648
# spp_vcv_ou[grep("Chironomus|Polypedilum", row.names(spp_vcv_ou)), grep("Chironomus|Polypedilum", colnames(spp_vcv_ou))]
# # Same subfamily (e.g. Chiornominae) corr = 0.8157158
# spp_vcv_ou[grep("Conochironomus|Polypedilum", row.names(spp_vcv_ou)), grep("Conochironomus|Polypedilum", colnames(spp_vcv_ou))]
# # Confamilials e.g Tateidae, Hydropsychidae corr =  0.7521885
# spp_vcv_ou[grep("pyrgus", row.names(spp_vcv_ou)), grep("pyrgus", colnames(spp_vcv_ou))]
# spp_vcv_ou[grep("Cheumatopsyche|Asmicridea", row.names(spp_vcv_ou)), grep("Cheumatopsyche|Asmicridea", colnames(spp_vcv_ou))]
# # Same order e.g Plecoptera = 0.5720539
# spp_vcv_ou[grep("Dinotoperla|Cosmioperla", row.names(spp_vcv_ou)), grep("Dinotoperla|Cosmioperla", colnames(spp_vcv_ou))]
# # Interpret the alpha_phylo value

alpha_draws <- mod_fit$draws(variables = "alpha_phylo")
# mean(alpha_draws) # 6.826893; 
# quantile(alpha_draws, probs = c(0.025,0.975)) # 5.514889 8.437285

corrs_by_rank <- data.frame(common_rank = c("genus","tribe","subfamily","family","order"),
                            phylo_cor = c(0.9114853,0.8870648,0.8157158,0.7521885,0.5720539),
                            mean = NA, cl_025 = NA, cl_100 = NA, cl_900 = NA, cl_975 = NA)
for(i in 1:nrow(corrs_by_rank)){
  corri <- exp(-2 * alpha_draws * (1- corrs_by_rank$phylo_cor[i]))
  corrs_by_rank$mean[i] <- mean(corri)
  corrs_by_rank[i,4:7] <- quantile(corri, probs = c(0.025, 0.1, 0.9, 0.975))
}

plot(1:5, corrs_by_rank$mean, ylim = c(0,0.4), xlim = c(0.5,5.5), 
     axes = FALSE, xlab = "Common taxonomic rank", 
     ylab = "Correlation between related species") 
for(i in 1:nrow(corrs_by_rank)){
  lines(c(i,i),c(corrs_by_rank$cl_025[i], corrs_by_rank$cl_975[i]))
  lines(c(i,i),c(corrs_by_rank$cl_100[i], corrs_by_rank$cl_900[i]), lwd = 3)
}
points(1:5, corrs_by_rank$mean, pch = 21, bg = "grey", cex = 1.25)
axis(1, at = 1:5, labels = corrs_by_rank$common_rank); axis(2, las = 1); box(bty = 'l')
abline(h = 0, lty = 3)
```
