---
title: "Model data assembly for the 225-site dataset"
author: "Thomas Wilkins and Christopher J Walsh"
format: html
editor: source
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false

suppressPackageStartupMessages(library(cmdstanr))
# Appends the threading variable to your Stan makefile configuration
cpp_options <- list(stan_threads = TRUE)
options(cmdstanr_max_num_threads = 4)
suppressPackageStartupMessages(library(posterior))
suppressPackageStartupMessages(library(data.table))
library(loo)

# Random seed to fix Stan results
proj_seed <- 12345 
set.seed(proj_seed)

# Network functions used herein: 
source("https://tools.thewerg.unimelb.edu.au/mwstr/data/mwstr_network_functions.R")
# Bug database functions used herein: ct(), 
source("https://tools.thewerg.unimelb.edu.au/mwbugs/data/bug_database_functions.R")
right <- function(x, n) {substr(x, nchar(x) - n + 1, nchar(x))}

```

## Introduction

This is a cutdown version of MW_metabarcoding_data.qmd for compiling and testing a 200-species subset of data in preparation for scaling up to the full 875-species dataset

### Retrieve site and sample from the Melbourne Water macroinvertebrate database

```{r}
#| label: mwbugs_data_retrieval_1
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

# Connect to MW databases:
mwbugs <- DBI::dbConnect(RPostgres::Postgres(), dbname = "mwbugs", host = "localhost")
mwstr <- DBI::dbConnect(RPostgres::Postgres(), dbname = "mwstr", host = "localhost")
# Project 3:  Metabarcoded combined-RBA-sample-pair survey of 223 sites across
#   Melbourne 2021-2024 (Nickname MW metabarcoding survey)
samples <- DBI::dbGetQuery(mwbugs, 
                          "SELECT * FROM samples 
                           JOIN sample_project_groups
                           ON samples.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
samples$sitecode[grep("TOO",samples$sitecode)] <- "TOO-4334-2"
samples$sitecode_v12[grep("TOO",samples$sitecode_v12)] <- "TOO_4332c"
sites <- sf::st_read(mwbugs, query = 
                      paste0("SELECT * FROM sites WHERE sitecode_v12 IN ('",
                             paste(samples$sitecode_v12, collapse = "','"),"');"))
# Create an integer site number for use in the Stan model
sites <- dplyr::mutate(sites, site_no = 1:nrow(sites), .before = sitecode)

```

<!--# CW note to self: Perhaps also worth considering removing LSN, with a lot of SCMs not accounted for in its EI measurement -->

```{r}
#| label: mwbugs_data_retrieval_2
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

biota_all <- DBI::dbGetQuery(mwbugs, 
                           "SELECT * FROM biota 
                           JOIN sample_project_groups
                           ON biota.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
biota_all <- biota_all[!grepl("EXCLUDE this record from 3-rep", 
                          biota_all$notes),]  #20584
species <- unique(biota_all[c("taxoncode","taxon")])
# nrow(samples[!samples$smpcode %in% unique(biota_all$smpcode),])  # 0
# confirms there are 0 samples without biota.

```

<!--# CW note to self: chase down these bibtex records.-->

The biota table contained 20694 occurrence records of 882 species from the 275 samples. The dataset included samples from the bulk-processing experiment, which was conducted to develop and optimize the sample-processing methods [@carew_etal_2026, @wilkins_etal_2026a].  The samples for that experiment were were fully processed, with 8-11 replicate MiSeq samples. All species detected in those samples are included in the biota table.  The samples can be adjusted to  approximate the standard 30% subsampled, 3-replicate bulk-processing method by excluding records with the phrase:  'EXCLUDE this record from 3-rep 30% sample' in the notes column, as per the code above. 

### Subset of biota for trial analyses

We began with a tractable small set of species of varying taxonomic relationships and potential response, choosing species with moderate prevalance among sites (between 2 and 80%).

```{r}

# Note range(biota$count) = 1,1
# calculate prevalence on sites rather than samples
# add reach_v12 to samples for matching biota 
samples$reach_v12 <- sites$reach_v12[match(samples$sitecode, sites$sitecode)]
biota_all$reach <- samples$reach_v12[match(biota_all$smpcode, samples$smpcode)]
biot_by_site <- unique(biota_all[c("taxoncode","taxon","reach")])
biot_prev <- aggregate(biot_by_site$taxoncode, 
                       by = list(taxoncode = biot_by_site$taxoncode, taxon = biot_by_site$taxon), 
                       FUN = length)
biot_prev$prevalence <- biot_prev$x/length(unique(samples$reach_v12))
biot_prevalent <- biot_prev[biot_prev$prevalence >= 0.2 & biot_prev$prevalence <= 0.8,]  # 111 species
biot_non_prevalent <- biot_prev[!biot_prev$taxoncode %in% biot_prevalent$taxoncode,] #777 species
# about 1 quarter of that might be a good number to start with
# But first quarantine a few congenerics to make the phylogenetic similarity diverse
spp_trial <- biot_prevalent[grepl("QT2505",biot_prevalent$taxoncode) | 
                            grepl("QT2511",biot_prevalent$taxoncode) |
                            grepl("QE0609",biot_prevalent$taxoncode),]
spp_trial_1 <- biot_prevalent
biot_prevalent <- biot_prevalent[!biot_prevalent$taxoncode %in% spp_trial$taxoncode,] # down to 104
spp_trial <- rbind(spp_trial, biot_prevalent[seq(4,104,4),])  
# subset of 33 for first preliminary trial. I want to make sure that spp_trial is a subset of the larger spp_trial_1
# sum(!spp_trial$taxoncode %in% spp_trial_1$taxoncode) # 0 # all good
# A secondary subset for model selection without epsilon parameter: all prevalent species + 
# Add less (or more) prevalent species, with increased chance of congenerics by selecting the adjacent species
for(i in 1:nrow(biot_prevalent)){
spp_trial_1 <- unique(rbind(spp_trial_1, biot_prev[which(biot_prev$taxon == biot_prevalent$taxon[i]) + 1,]))
} # 197  = ~ 20% of the full set
# I also want to include some lowland species to give the "L" predictor a fair go
lowland_spp <- c("OP0599A1",  # 1 record - Cannings St ford
"IB020101","KG021202") # both with lots of lowland records
# sum(lowland_spp %in%spp_trial_1$taxoncode )  # None included so far, so add them
spp_trial_1 <- unique(rbind(spp_trial_1, biot_prev[biot_prev$taxoncode %in% lowland_spp,]))  #200 spp

biota <- biota_all[biota_all$taxoncode %in% spp_trial$taxoncode,]  #down to 2780 records for the first 33 spp trial
biota_1 <- biota_all[biota_all$taxoncode %in% spp_trial_1$taxoncode,]  #11194 records

### Biota presence/absence matrix
biota_1_ct <- with(biota_1, ct(smpcode, shortcode, count))

# Two species shown to be predicted well in the first trial to aid the latent-factor ordination
well_behaved_sp_pr <- c("QT250713", "IH010101")

```

### Phylogenetic variance-covariance matrix

```{r}
#| label: fig-spp_tree
#| echo: true
#| message: false
#| warning: false
#| fig-height: 7
#| fig-width: 6
#| fig-cap: "The taxonomic tree for the trial subset of 33 species modelled here" 

# Make sure the taxonomic tree matrix is ordered the same as biota_1_ct
spp_trial_1 <- spp_trial_1[match(colnames(biota_1_ct),spp_trial_1$taxoncode),]
# load clossification object compiled in ~/uomShare/wergStaff/ChrisW/git-data/mwbugs/build_phylo_vcv.R
spp_class_all <- readRDS(
  url("https://tools.thewerg.unimelb.edu.au/mwbugs/data/spp_classes_itis.rds"))
# Select the subset of taxa relevant for the current analysis (or use all 982 spp)
spp_class_1 <- spp_class_all[c(spp_trial_1$taxon)]
spp_tree_1 <- taxize::class2tree(spp_class_1)
# compute branch lengths to reflect evolutionary divergence
spp_vcv <- ape::vcv(spp_tree_1$phylo, corr = TRUE)
# par(mar = c(0,1,0,1)); plot(spp_tree_1, cex = 0.25)

# Using spp_vcv resulted in a mean rho_phylo value of 0.268 (90% CLs 0.184,0.359)
# Try censoring the matrix so that only family and below levels have values > 0
# (family-family corr = 0.7518585) - hypothesis being that niche conservatism is
# only likely at family level or below...
spp_vcv_f <- spp_vcv
spp_vcv_f[spp_vcv_f < 0.75] <- 0
# The model 

# The following approach, borrowing from the taxonomic index of Clarke and Warwick
# with equally spaced correlation for taxonomic levels genus-tribe-subfamily-family
# and zero above that, sampled more efficiently (min ess_bulk 486 with 1200 sampling 
# iteratoins) than a model using either of the above versions of spp_vcv and 
# produced a mean rho_phylo of 0.273 (90% CLs 0.187, 0.365)

taxonomy_df <- data.frame(species = spp_trial_1$taxon, genus = NA,  
                          tribe = NA, subfamily = NA, family = NA) 
for(i in 1:nrow(taxonomy_df)) {
  spi <- spp_trial_1$taxon[i]
  spi_df <- spp_class_1[[spi]]
  if (!is.null(spi_df)) {
    t_match <- spi_df$name[spi_df$rank == "genus"]
    taxonomy_df$genus[i] <- if(length(t_match) > 0) t_match[1] else paste0("unknown_gen_", i)
    t_match <- spi_df$name[spi_df$rank == "tribe"]
    taxonomy_df$tribe[i] <- if(length(t_match) > 0) t_match[1] else paste0("unknown_tribe_", i)
    t_match <- spi_df$name[spi_df$rank == "subfamily"]
    taxonomy_df$subfamily[i] <- if(length(t_match) > 0) t_match[1] else paste0("unknown_subfam_", i)
    t_match <- spi_df$name[spi_df$rank == "family"]
    taxonomy_df$family[i] <- if(length(t_match) > 0) t_match[1] else paste0("unknown_fam_", i)
  } else {
    taxonomy_df$genus[i] <- paste0("unresolved_gen_", i)
    taxonomy_df$tribe[i] <- paste0("unresolved_tribe_", i)
    taxonomy_df$subfamily[i] <- paste0("unresolved_subfam_", i)
    taxonomy_df$family[i] <- paste0("unresolved_fam_", i)
  }
}

taxonomic_cor <- matrix(0, nrow = nrow(spp_trial_1), ncol = nrow(spp_trial_1))
rownames(taxonomic_cor) <- taxonomy_df$species
colnames(taxonomic_cor) <- taxonomy_df$species
for (i in 1:nrow(spp_trial_1)) {
  for (j in 1:nrow(spp_trial_1)) {
    if (i == j) {
      taxonomic_cor[i, j] <- 1.00 # Self-correlation
    } else if (taxonomy_df$genus[i] == taxonomy_df$genus[j] && !grepl("^No", taxonomy_df$genus[i])) {
      taxonomic_cor[i, j] <- 0.80 # genus-genus pairs
    } else if (taxonomy_df$tribe[i] == taxonomy_df$tribe[j] && !grepl("^No", taxonomy_df$tribe[i])) {
      taxonomic_cor[i, j] <- 0.60 # tribe-tribe pairs
    } else if (taxonomy_df$subfamily[i] == taxonomy_df$subfamily[j] && !grepl("^No", taxonomy_df$subfamily[i])) {
      taxonomic_cor[i, j] <- 0.40 # subfamily-subfamily pairs
    } else if (taxonomy_df$family[i] == taxonomy_df$family[j] && !grepl("^No", taxonomy_df$family[i])) {
      taxonomic_cor[i, j] <- 0.20 # family-family pairs
    } else {
      taxonomic_cor[i, j] <- 0.00 # Anything higher than family is zero
    }
  }
}
# Matrix validation for Stan's Cholesky factorisation
spp_vcv_suborder <- as.matrix(Matrix::nearPD(taxonomic_cor, corr = TRUE)$mat)

```


### Spatial autocorrelation among sites

Removed from this consideration.  See MW_metabarcoding_data_compilation.qmd in case of wanting to add it back in.

### Compile environmental predictors

```{r}
#| label: env_data_retrieval
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

# Add fields for matching with other tables to the samples table
samples$reach_v12 <- substr(samples$sitecode_v12,1,nchar(samples$sitecode_v12)-1)
# numeric site identifier for stan (site_no)
samples$site_no <- sites$site_no[match(samples$sitecode_v12, sites$sitecode_v12)]

# Derive season and riffle fields from sample specifications
samples$season <- 0; samples$season[lubridate::month(samples$date) > 6] <- 1 
# Season - 0 = Autumn; 1 = Spring
# samples with collection_method mentionnig riffle, include 1 riffle RBA sample
samples$riffle <- 0; 
samples$riffle[grep("riffle", samples$collection_method)] <- 1 

# Reduce the samples table to relevant fields and add sample-level environmental 
# data ("season" and "riffle")
samples <- samples[,c("smpcode","date","sitecode_v12","reach_v12",
                     "site_no", "season", "riffle")]

# Attenuated Forest cover: AF - optimal parameterization from wilkins_etal_2026 
# (See also Walsh (2023) Chapter 7)-  drawn from the mwbugs database,
# which accounts for location of the sample site within the reach
sites_af <- DBI::dbGetQuery(mwbugs, 
              paste0("SELECT sitecode_v12, af_l5_w633_1750, ",
                     "af_l5_w633_2018, af_l5_w633_2022 ",
                     "FROM sitecode_env WHERE sitecode_v12 IN ('",
                     paste(sites$sitecode_v12, collapse = "', '"),"');"))
# Check all sites are in sites_af: 
# sum(!sites$sitecode_v12 %in% sites_af$sitecode_v12)  # zero missing. all good.
sites$af <- sites_af$af_l5_w633_2022[match(sites$sitecode_v12, sites_af$sitecode_v12)]
# A major storm on 9 June 2021 caused much forest loss https://www.theguardian.com/australia-news/2021/jun/24/dandenong-ranges-power-outage-storm-debris-victoria-emergency-services-warning
# I originally used af_2018 before this date and af_2022 after, but on checking noted that it only changed af values for 
# urban sites for which the forest had no effect (and DNG_7957, where af was slightly higher in 2022 than 2018)
# Therefore leave af as invariate across samples from each site (Which also makes modelling easier)
# samples$af[samples$date < "2021-06-10"] <- 
# sites_af$af_l5_w633_2018[match(samples$sitecode_v12[samples$date < "2021-06-10"],
#                          sites_af$sitecode_v12)]
sites$af_ref <- sites_af$af_l5_w633_1750[match(sites$sitecode_v12, sites_af$sitecode_v12)] 
# reference state for AF to be used for predictions from the ultimate model

# Impervious cover and other environmental variables from the mwstr database
cat_env <- DBI::dbGetQuery(mwstr,
  paste0("SELECT subcs.reach, subcs.carea_km2, ei_l44_sw1942_2022, c_basalt, ", 
         "ti_2022, tf_2018, tf_2022, meanq_mm, dor_2010 FROM cat_env ",
         "JOIN subcs ON cat_env.site = subcs.site ","WHERE cat_env.reach IN ('",
         paste(sites$reach_v12, collapse = "', '"), "');"))

subc_env <- DBI::dbGetQuery(mwstr,
               paste0("SELECT reach, meant_y30_2022, elev_asl_m FROM subc_env ",
                      "WHERE reach IN ('",
                      paste(sites$reach_v12, collapse = "', '"), "');"))
# Catchment area in sq km. 
sites$carea_km2 <- cat_env$carea_km2[match(sites$reach_v12, cat_env$reach)] 
# Proportion of catchment with basalt geology See Walsh (2023) Chapter 5.7
sites$basalt <- cat_env$c_basalt[match(sites$reach_v12, cat_env$reach)]
# Degree of regulation: see Walsh (2023) Chapter 9
sites$dor <- cat_env$dor_2010[match(sites$reach_v12,cat_env$reach)] 
# Effective imperviousness: see Walsh (2023) Chapter 6
sites$ei <- cat_env$ei_l44_sw1942_2022[match(sites$reach_v12,cat_env$reach)] 
# Total imperviousness:  see Walsh (2023) Chapter 6
sites$ti <- cat_env$ti_2022[match(sites$reach_v12,cat_env$reach)] 
# Total forest cover:  see Walsh (2023) Chapter 7
sites$tf <- cat_env$tf_2022[match(sites$reach_v12,cat_env$reach)] 
# Mean annual runoff depth (mm). See Walsh (2023) Chapter 8
sites$meanq <- cat_env$meanq_mm[match(sites$reach_v12,cat_env$reach)] 
# Mean annual air temperature (degrees C). See Walsh (2023) Chapter 5.8
sites$meant <- subc_env$meant_y30_2022[match(sites$reach_v12,subc_env$reach)] 
# Elevation above sea-level (m). See Walsh (2023) Chapter 5.9
sites$elev <- subc_env$elev_asl_m[match(sites$reach_v12,subc_env$reach)]
# A lowland/estuarine influenc indicator
sites$lowness <- -1
sites$lowness[sites$elev < 30] <- -0.5
sites$lowness[sites$elev < 15] <- 0
sites$lowness[sites$elev < 7.5] <- 0.5
sites$lowness[sites$elev < 2.5] <- 1

# Transformations and interactions prior to scaling
sites$lcarea <- log(sites$carea_km2)
sites$smeanq <- sites$meanq^0.5
sites$dor1 <- pmin(sites$dor,1)
sites$lei <- log(sites$ei + 0.001)
sites$q_af <- sites$smeanq * sites$af
sites$q_ei <- sites$smeanq * sites$lei
sites$b_ei <- sites$basalt * sites$lei
sites$q_d <- log(sites$smeanq * sites$dor1 + 0.1)
sites$af_ei <- sites$af * sites$lei

# Keep a summary table of predictors

```

### Exploratory analysis of outliers and multicollinearity among predictor variables.

See MW_metabarcoding_data_compilation.qmd  for the exploratory analysis used to reduce the set of candidate predictors

### Assembling model data for Stan


```{r}
#| label: stan_data
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

biota_1_ct <- biota_1_ct[match(samples$smpcode,row.names(biota_1_ct)),] # ensure c and u rows are ordered the same way
# make first two well-predicted prevalent species to stabilize the lambda_diag constraint

# Ensure samples and biota_1_ct rows are in the same order
samples <- samples[match(row.names(biota_1_ct), samples$smpcode),]
sum(!samples$smpcode == row.names(biota_1_ct))  # should be zero

# Index for sites in samples table
site <- samples$site_no 

### Two predictor matrices: one for site-specific variables (u_site), 
###                         one for sample(obs)-specific variables (u_obs)

# scale fixed predictor variables in 
# keeping them as scaled objects for ease of back-transforming later
I <- scale(sites$lei)
F <- scale(sites$af)
Q <- scale(sites$smeanq)
Q2 <- scale(sites$smeanq^2)
C <- scale(sites$lcarea)
T <- scale(sites$meant)
D <- scale(sites$dor1)
B <- scale(sites$basalt)
Q_F <- scale(sites$q_af)
Q_I <- scale(sites$q_ei)
Q_D <- scale(sites$q_d)
F_I <- scale(sites$af_ei)
B_I <- scale(sites$b_ei)
L <- sites$lowness

# Save scaling attributes for all scaled parameters
pred_specs <- data.frame(db_name = c("ei_l44_sw1942_2022","af_l5_w633_2022","meanq_mm","meanq_mm","carea_km2","dor_2010","meant_y30_2022",
                                     "elev_asl_m","basalt","ei*af","meanq*af","meanq*ei","basalt*ei","meanq*dor"),
                         short_name = c("ei","af","meanq","meanq2","carea_km2","dor","meant","lowness","c_basalt",
                                        "if","qf","qi","bi","qd"),
                         transf_name = c("lei","af","smeanq","meanq","lcarea","dor1","meant","lowness","basalt",
                                         "af_ei","q_af","q_ei","b_ei","q_d"),
                         transformation = c("log(x+0.001)",NA,"x^0.5",NA,"log(x)","pmin(x,1)",NA,"bespoke scale",NA,"lei*af","smeanq*af","smeanq*lei","basalt*lei","log(smeanq*dor1+0.1"),
                         scaled_name = c("I","F","Q","Q2","C","D","T","L","B","F_I","Q_F","Q_I","B_I","Q_D"),
                         scale = NA, center = NA)

for(i in 1:nrow(pred_specs)){
  pred_specs$scale[i] <- ifelse(pred_specs$scaled_name[i] %in% c("B","L"), NA,
                                  attr(get(pred_specs$scaled_name[i]), "scaled:scale"))
  pred_specs$center[i] <- ifelse(pred_specs$scaled_name[i] %in% c("B","L"), NA,
                                   attr(get(pred_specs$scaled_name[i]), "scaled:center"))
}

# Save the scaled parameters for use in predictions
write.csv(pred_specs, "~/uomShare/wergStaff/ChrisW/git-data/mw_metabarcoding_model/parameter_scaling_attributes.csv")

u_site <- unique(as.matrix(data.frame(I = I,      # Effective imperviousness
                 F = F,                   # Attenuated Forest cover
                 Q = Q,                   # mean annual runoff
                 Q2 = Q2,                 # square of Q for modal response
                 C = C,                   # catchment area
                 D = D,                   # Degree of regulation
                 T = T,                   # Temperature 
                 L = L,                   # lowness
                 B = B,                   # Basalt
                 F_I = F_I,                # Interaction betwen I and F
                 Q_F = Q_F,                # Interaction betwen Q and F
                 Q_I = Q_I)))             # Interaction between Q and D
 # diff(site) # all zeros and 1s meaning that unique(u_site) is ordered by site
u_obs <- as.matrix(data.frame(riff = samples$riffle, season = samples$season))

# Matrix for modelling latent variation unrelated to predictors
# Compute the Hat projection matrix onto the site covariate space
H_site <- u_site %*% solve(t(u_site) %*% u_site) %*% t(u_site)
# Compute the Residual Projection Matrix (The Null-Space Operator)
M_site_res <- diag(nrow(u_site)) - H_site

# Assemble biota and predictor data for post-processing purposes
species_set <- colnames(biota_1_ct)
species_set <- biot_prev[match(species_set, biot_prev$taxoncode),]
sample_set <- row.names(biota_1_ct)
sample_set <- samples[match(sample_set, samples$smpcode),c("smpcode","reach_v12","site_no")]

# Assemble stan data for second trial set
master_data <- list(n_obs = nrow(biota_1_ct),          # no. samples
                 n_site_pred = ncol(u_site),           # no. site-specific predictors
                 n_obs_pred = ncol(u_obs),             # no. sample-specific predictors
                 n_taxa = ncol(biota_1_ct),            # no. taxa
                 n_site = max(site),                   # no. sites with a site upstream
                 n_latent = 2,                         # no. latent factors for site effect
                 u_site = u_site,                      # site-specific predictor matrix
                 u_obs = u_obs,                        # sample-specific predictor matrix
                 y = as.matrix(biota_1_ct),
                 site  = site,
                 M_site_res = M_site_res,              # Residual projection matrix for latent-factor estimation
                 phylo_cor = spp_vcv_f)                # Phylogenetic covariance   

```

### Models

```{r}
params_to_summarise <- c(
  "mu_beta_site", "scale_beta_site", "beta_site",  # Fixed site-level effects
  "mu_beta_obs", "scale_beta_obs", "beta_obs",  # Fixed sample-level effects
  "mu_taxon", "sigma_taxon",        # Species hyper-parameters
   "z_expanded", "lambda",          # latent factors - these do not converge as they are unconstrained
  "a_taxon", "rho_phylo",           # Random species effect, strength of phylogeny effect
  "tjurs_r2_total",                 # Explanatory power using fixed predictors and latent random site variation
  "tjurs_r2_env",                   # Explanatory power solely using fixed predictors
  "y_rep"                           # For observed vs predicted plots
) # to reduce the RAM required to build the summary table

pred_site <- c("I","F","Q","Q2","C","D","T","B","L","F_I","Q_F","Q_I")
pred_obs <- c("riff","season") # included in all candidate models
mod_code <- "ifqcdtblrs_fi_qf_qi_spp_vcv_f" 
pred_set <- u_site[,match(pred_site, colnames(u_site))]
mod_data <- master_data
mod_data$n_site_pred <- length(pred_site)
mod_data$u_site <- pred_set
mod_bundle <- list(species_set = species_set, sample_set = sample_set, 
                   pred_set = pred_set, data = mod_data)
mod <- cmdstan_model(stan_file = "stancode/jointspp_pa_no_eps.stan",
                     compile_model_methods = TRUE) #, 
                    #  force_recompile = TRUE )
mod_dir <-paste0("~/uomShare/wergStaff/ChrisW/git-data/",
              "mw_metabarcoding_model/trial_200spp_models/", mod_code)
if (!dir.exists(mod_dir)) dir.create(mod_dir)
model_fit <- mod$sample(data = mod_data, seed = proj_seed,
                        output_dir = mod_dir,
                        chains = 4, 
                        parallel_chains = 4,
                        iter_warmup = 800,  
                        iter_sampling = 1200)
mod_bundle$fit <- model_fit
mod_bundle$diagnostics <- model_fit$diagnostic_summary()
saveRDS(mod_bundle, file = paste0(mod_dir, "/", mod_code, "model_bundle.rds"))
mod_summary <- model_fit$summary(variables = params_to_summarise); gc()
mod_bundle$summary <- mod_summary
saveRDS(mod_bundle, file = paste0(mod_dir, "/", mod_code, "model_bundle.rds"))
log_lik_draws <- model_fit$draws(variables = "log_lik", format = "draws_array")
ess <- apply(log_lik_draws, 3, ess_bulk)
total_draws <- posterior::ndraws(log_lik_draws)
r_eff <- ess / total_draws
rm(log_lik_draws); gc()
mod_loo <- model_fit$loo(variables = "log_lik", r_eff = r_eff, cores = 1); gc() # can spike RAM usage
mod_bundle$loo <- mod_loo
saveRDS(mod_bundle, file = paste0(mod_dir, "/", mod_code, "model_bundle.rds"))
rm(mod_bundle,mod_summary, model_fit, mod_loo); gc()

```



```{r}
source("cmdstanr_functions.R")
x <- compile_model_diagnostics(mod_code = "ifqcdtblrs_fi_qf_qi_spp_vcv", 
                                mod_dir = paste0("~/uomShare/wergStaff/ChrisW/git-data/",
                                                 "mw_metabarcoding_model/trial_200spp_models/"))
x
summary <-  load_model_component_bundle(mod_code = "ifqcdtblrs_fi_qf_qi_spp_vcv", 
                                   mod_dir = paste0("~/uomShare/wergStaff/ChrisW/git-data/",
                                                    "mw_metabarcoding_model/trial_200spp_models/"))$summary
summary_non_z <- summary[!grepl("z|lambda",summary$variable),]
summary_non_z[order(summary_non_z$ess_bulk),][1:2,]
summary_non_z[order(summary_non_z$ess_tail),][1:2,]

summary[summary$variable == "rho_phylo",]
# Extract draws as an array
draws_array <- model_fit$draws(variables = "z_expanded[14,2]")
# Create trace plots for specific parameters
bayesplot::mcmc_trace(draws_array)

# Extract your raw MCMC arrays from cmdstanr
# Dimensions: [Iterations, Chains, Sites, Latent_Axes]
z_raw_array <- model_fit$draws("z_expanded", format = "array")

# Pick your anchor reference matrix (e.g., Chain 1, Iteration 500)
target_matrix <- z_raw_array[500, 1, ]

# Create a blank array to hold the newly aligned, drift-free scores
aligned_z <- array(NA, dim = dim(z_raw_array))

# Loop through every iteration and chain to un-spin the drift cell-by-cell
for (c in 1:4) {
  for (i in 1:1000) {
    current_matrix <- z_raw_array[i, c, ]
    
    # procrustes calculates the exact rotation matrix needed for THIS step
    rotation <- procrustes(X = target_matrix, Y = current_matrix, symmetric = FALSE)
    
    # Store the rotated, perfectly aligned matrix
    aligned_z[i, c, ] <- rotation$Yrot
  }
}

par(mar = c(4,4,1,1))
plot(aligned_z[, 1, 1], type = "l", col = "red", ylim = range(aligned_z[, , 1], na.rm = TRUE),
     xlab = "Iteration", ylab = "Z Score", main = "Raw Trace: Observation 1, Axis 1")
lines(aligned_z[, 2, 1], col = "blue")
lines(aligned_z[, 3, 1], col = "green")
lines(aligned_z[, 4, 1], col = "purple")

# Chain 1 (Red) and Chain 4 (Purple) carry the actual latent signal
healthy_chains <- c(1, 4) 
n_healthy <- length(healthy_chains)

# Array to store the unpacked, aligned results
n_obs = 275; n_latent = 2
z_unpacked_aligned <- array(NA, dim = c(1200, n_healthy, n_obs, n_latent))

# Establish your master template using a stable iteration from your best chain (Chain 1)
first_vector <- aligned_z[600, 1, ]
target_matrix <- matrix(first_vector, nrow = n_obs, ncol = n_latent)
target_matrix <- scale(target_matrix) # Standardize the master template

# Run the loop ONLY across Chain 1 and Chain 4
for (idx in seq_along(healthy_chains)) {
  c <- healthy_chains[idx]
  
  for (i in 1:1200) {
    current_vector <- aligned_z[i, c, ]
    current_matrix <- matrix(current_vector, nrow = n_obs, ncol = n_latent)
    
    # Standardize to remove any scale discrepancies between Red and Purple
    current_matrix_scaled <- scale(current_matrix)
    
    if (any(is.na(current_matrix_scaled)) || any(is.infinite(current_matrix_scaled))) {
      # Skip this specific corrupted iteration and jump straight to the next one
      next 
    }
    # Rotate the iteration onto the master template
    rotation <- procrustes(X = target_matrix, Y = current_matrix_scaled, symmetric = FALSE)
    
    # Store the result cleanly
    z_unpacked_aligned[i, idx, , ] <- rotation$Yrot
  }
}

# Extract the clean posterior mean across just these two active chains
mean_obs_z1 <- apply(z_unpacked_aligned[, , , 1], 3, mean, na.rm = TRUE)
mean_obs_z2 <- apply(z_unpacked_aligned[, , , 2], 3, mean, na.rm = TRUE)

plot(mean_obs_z1, mean_obs_z2, pch = 19, 
     col = rgb(0, 0, 1, 0.6), xlab = "Latent Axis 1 (Residual Site Gradients)", 
     ylab = "Latent Axis 2 (Residual Site Gradients)", main = "Aligned Site Ordination Space")

# Map the 275 observation averages back to your 223 unique site IDs
ordination_df <- data.frame(
  Site = mod_data$site, # Your site index vector from Stan
  Z1   = mean_obs_z1,
  Z2   = mean_obs_z2
)
ybPal <- RColorBrewer::brewer.pal(9,"YlOrBr")
ordination_df$col_z1 <- ybPal[as.numeric(cut(ordination_df$Z1, breaks = 9))]
ordination_df$col_z2 <- ybPal[as.numeric(cut(ordination_df$Z2, breaks = 9))]
ord_site <- data.frame(site = 1:223, unique(ordination_df[,2:5]))
# Create a color gradient based on a missing variable (e.g., Longitude)
sites$col_z1 <- ord_site$Z1[match(sites$site_no, ord_site$site)]
sites$col_z2 <- ord_site$Z2[match(sites$site_no, ord_site$site)]
sites$z1 <- ord_site$Z1[match(sites$site_no, ord_site$site)]
sites$z2 <- ord_site$Z2[match(sites$site_no, ord_site$site)]
plot(sites$geom, pch = 21, bg = sites$col_z1)
plot(sites$geom, pch = 21, bg = sites$col_z2)
# Re-plot using the unique site data and the color gradient
plot(sf::st_coordinates(sites)[,2], sites$z1)
abline(lm(sites$z1 ~ sf::st_coordinates(sites)[,1]))
plot(sf::st_coordinates(sites)[,2], sites$z2)
abline(lm(sites$z2 ~ sf::st_coordinates(sites)[,1]))
summary(lm(sites$z1 ~ sf::st_coordinates(sites)[,1]))

plot(ordination_df$Z1, ordination_df$Z2, 
     pch = 19, col = site_colors,
     xlab = "Latent Axis 1", ylab = "Latent Axis 2",
     main = "Ordination Colored by Missing Gradient")

fit <- model_fit
fit <- load_model_component_bundle(mod_code = "ifqcdtblrs_fi_qf_qi", 
                                   mod_dir = paste0("~/uomShare/wergStaff/ChrisW/git-data/",
                                                    "mw_metabarcoding_model/trial_200spp_models/"))$fit
tjur_draws <- fit$draws(variables = c("tjurs_r2_env","tjurs_r2_total"),format = "draws_matrix")
tjur_quantiles <- data.frame(matrixStats::colQuantiles(tjur_draws,
                                        probs = c(0.025, 0.5, 0.975)))
par(mar = c(4,4,1,1)); 
plot(tjur_quantiles$X50.[1:200],1:200, xlim = c(0,1))
points(tjur_quantiles$X50.[201:400],1:200, col = "red")
abline(v = 0.1, lty = 3); abline(v = 0.5, lty = 3)
for(i in 1:200){
  lines(c(tjur_quantiles$X2.5.[i + 200],tjur_quantiles$X97.5.[i + 200]), rep(i,2), col = "red")
}

spno <- which(species_set$taxoncode == well_behaved_sp_pr[1])
spno_beta_names <- c(paste0("beta_site[",1:length(pred_site),",",spno,"]"),
                     paste0("beta_obs[",1:length(pred_obs),",",spno,"]"))
spno_beta_draws <- fit$draws(variables = spno_beta_names,format = "draws_matrix")
spno_beta_quantiles <- data.frame(matrixStats::colQuantiles(spno_beta_draws, 
                                                  probs = c(0.025, 0.5, 0.975)))
par(mfrow = c(1,1),bg = "white", mar = c(4,4,1,1))
plot(spno_beta_quantiles$X50., 1:length(spno_beta_names), 
     xlim = range(spno_beta_quantiles), axes = FALSE, 
     xlab = "Estimate", ylab = "")
for(i in 1:length(spno_beta_names)){
  coli <- ifelse(spno_beta_quantiles$X97.5.[i] < 0, "#1b9e77", 
                 ifelse(spno_beta_quantiles$X2.5.[i] > 0, "#d95f02", "black"))
  points(spno_beta_quantiles$X50.[i], i, pch =21, bg = coli)
  lines(c(spno_beta_quantiles$X2.5.[i],spno_beta_quantiles$X97.5.[i]), 
        rep(i,2), col = coli)
        }
axis(1); abline(v = 0, lty = 3); box(bty = "l")
axis(2, at = 1:length(spno_beta_names), labels = c(pred_site, pred_obs), las = 1, cex.lab = 0.8)

```

