---
title: "Latent ordination in the 200-species model"
author: "Christopher J Walsh"
format: docx
editor: source
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false

source("quick_200_spp_data_assembly_for_latent_tests.R")
# This creates data ready for the model for constrained_z
source("cmdstanr_functions.R")
# post processing functions

```


"ifqcdtblrs_fi_qf_qi_constr_z" trialled constraining the latent ordination by selecting two species with divergent responses to the unconstrained ordination in "ifqcdtblrs_fi_qf_qi_spp_vcv_ou".  The resulting model suffered from split chains and very low Effective Sample Size for tjur_r2 of the selected anchor species.  This indicates that the constrained approach is not applicable to this high-dimension dataset.  We thus implemented an unconstrained, non-centered latent factor ordination, which requires post-processing [@assman_etal_2014].

```{r}
#| eval: false

pred_site <- c("I","F","Q","Q2","C","D","T","B","L","F_I","Q_F","Q_I")
pred_obs <- c("riff","season") # included in all candidate models

 # "ifqcdtblrs_fi_qf_qi_constr_z" # Note taxa re-ordered to facilitate constrained z
run_200spp_model(pred_site = pred_site,
                 mod_code = "ifqcdtblrs_fi_qf_qi_constr_z")
# This did not work well.

```

The following therefore first repeats the code used to run the optimal model using the OU model for phylogenetic covariance, followed by post-processing in R to align the z_lambda drifting latent ordinations. 

```{r}
#| eval: false

#### standardized phylogenetic matrix phylo_cor
spp_class_all <- readRDS(
  url("https://tools.thewerg.unimelb.edu.au/mwbugs/data/spp_classes_itis.rds"))
spp_trial_1 <- spp_trial_1[match(colnames(biota_1_ct),spp_trial_1$taxoncode),]
spp_class_200 <- spp_class_all[spp_trial_1$taxon]
# Select the subset of taxa relevant for the current analysis (or use all 982 spp)
spp_tree_1 <- taxize::class2tree(spp_class_200)
# Scale the tree so total root-to-tip height is exactly 1.0
spp_tree_1_raw <- spp_tree_1$phylo
max_height <- max(ape::node.depth.edgelength(spp_tree_1_raw))
scaled_tree <- spp_tree_1_raw
scaled_tree$edge.length <- scaled_tree$edge.length / max_height
spp_vcv_ou <- ape::vcv(scaled_tree)

master_data$phylo_cor <- spp_vcv_ou
params_to_summarise[params_to_summarise == "rho_phylo"] <- "alpha_phylo"
run_200spp_model(pred_site = pred_site,
                 mod_code = "ifqcdtblrs_fi_qf_qi_ou",
                 mod_path = "stancode/jointspp_pa_no_eps_ou.stan")

```

The need for post-processing is illustrated in @fig-alignment, showing the raw traceplots of the four chains for latent axes 1 and 2 for site 1 (A. z_expanded[1,1], B. z_expanded[1,2]) from the Stan model. They show that the score estimates for each chain drifting over the 1200 iterations.  This is because the model does not constrain the ordination to a set rotation (and attempting to do so resulted in very low effective sample sizes).  The following code chunk applies a Procrustes rotation to the ordination in each iteration, which results in well-mixed, stable traceplots for the aligned z-values (C and D).

```{r}
#| label: fig-alignment
#| echo: false
#| fig-width = 5
#| fig-height = 5
#| fig-cap = "Traceplots of raw estimates of example latent factor score for site 1 (A. axis 1, B. axis 2) for 1200 iterations in the 200-species model.  C and D show the same estimates after Procrustes rotation."

load("~/uomShare/wergStaff/ChrisW/git-data/mw_metabarcoding_model/examp_draws.rda")
par(mfrow = c(2,2), mar = c(4,4,1,1))
base_traceplot(examp_draws$z11_unaligned, "z_expanded[1,1]"); title(main = "A.", adj = 0)
base_traceplot(examp_draws$z12_unaligned, "z_expanded[1,2]"); title(main = "B.", adj = 0)
base_traceplot(examp_draws$z1_aligned[,,1, drop = FALSE], "z_aligned[1,1]"); title(main = "C.", adj = 0)
base_traceplot(examp_draws$z1_aligned[,,2, drop = FALSE], "z_aligned[1,2]"); title(main = "D.", adj = 0)

```

The following code aligns the latent factor parameters z_expanded and lambda.

```{r}

mod_dir <- "~/uomShare/wergStaff/ChrisW/git-data/mw_metabarcoding_model/trial_200spp_models/"
mod_bundle <- load_model_component_bundle(mod_code ="ifqcdtblrs_fi_qf_qi_ou", mod_dir = mod_dir)
mod_fit <-  mod_bundle$fit
mod_data <- mod_bundle$data

# Ensure your dimensions are correctly defined for your active model
n_obs <- mod_data$n_obs
n_site <- mod_data$n_site
n_taxa <- mod_data$n_taxa
n_latent <- mod_data$n_latent
n_chains <- mod_fit$num_chains()
n_iterations <- mod_fit$metadata()$iter_sampling

# dimensions  [1200 iterations, 4 chains, 550 flattened parameters]
raw_z <- mod_fit$draws("z_expanded", format = "array")
raw_lambda <- mod_fit$draws("lambda_raw", format = "array")

# Establish a baseline template using a stable iteration 
# example_z <- mod_fit$draws("z_expanded[1,1]", format = "array")
# base_traceplot(example_z, "z_expanded[1,1]") # all respectable. Pick chain 1
# Save example unaligned (and later aligned) z scores for plotting above.
# examp_draws <- list(z11_unaligned = example_z, 
#                     z12_unaligned = mod_fit$draws("z_expanded[1,2]", format = "array"))
first_vector <- raw_z[500, 4, ]

target_matrix <- matrix(NA, nrow = n_obs, ncol = n_latent)
target_matrix[, 1] <- first_vector[1:n_obs]           # Elements 1 to 275 = Axis 1
target_matrix[, 2] <- first_vector[(n_obs + 1):550]    # Elements 276 to 550 = Axis 2

# Standardise the master frame
target_matrix <- scale(target_matrix) 

# Create empty, 4D arrays to store the aligned results
# dim =iterations, chains, observations, latent axes
z_unpacked_aligned <- array(NA, dim = c(n_iterations, n_chains, n_obs, n_latent))
lambda_unpacked_aligned <- array(NA, dim = c(n_iterations, n_chains, n_taxa, n_latent))

for (c in 1:n_chains) {
  for (i in 1:n_iterations) {
    # current 550-element vector
    current_vector <- raw_z[i, c, ]
    # Create an empty  275 x 2 matrix to be populated
    current_matrix <- matrix(NA, nrow = n_obs, ncol = n_latent)
    current_matrix[, 1] <- current_vector[1:n_obs]            # Dimension 1 (Axis 1)
    current_matrix[, 2] <- current_vector[(n_obs + 1):550]    # Dimension 2 (Axis 2)
    # Standardise to remove cross-chain scale discrepancies
    current_matrix_scaled <- scale(current_matrix)
    if (any(is.na(current_matrix_scaled)) || any(is.infinite(current_matrix_scaled))) {
      next 
    }
    # Rotate the iteration onto the master template: dimensions [275,2]
    rotation <- vegan::procrustes(X = target_matrix, Y = current_matrix_scaled, symmetric = FALSE)
    z_unpacked_aligned[i, c, , ] <- rotation$Yrot
    # Extract the 2 x 2 rotation matrix calculated for this iteration for aligning the species loadings (lambda)
    rotation_matrix <- rotation$rotation
    current_lambda_vector <- raw_lambda[i, c, ]
    current_lambda_matrix <- matrix(current_lambda_vector, nrow = n_taxa, ncol = n_latent)
    # multiply lambda by the same rotation matrix
    lambda_unpacked_aligned[i, c, , ] <- current_lambda_matrix %*% rotation_matrix
  }
}
# Add to data for example plot above
# examp_draws$z1_aligned <- z_unpacked_aligned[,,1,]
# save(examp_draws, file = "~/uomShare/wergStaff/ChrisW/git-data/mw_metabarcoding_model/examp_draws.rda")

####
# 1. Recalculate draws for aligned Z (latent factor scores for sites)
# z_expanded values are identical for observations from the same sites, so just use one value per site
unique_site_ids <- 1:n_site
first_obs_indices <- match(unique_site_ids, mod_data$site)
all_sites_samples_raw <- z_unpacked_aligned[, , first_obs_indices, ] # dim [1200, 4, 223, 2]
# Reshape from a 4-D array into a standard 3-D MCMC array: [1200, 4, 446]
n_parameters <- n_site * n_latent
all_sites_samples_flat <- array(all_sites_samples_raw, dim = c(n_iterations, n_chains, n_parameters))
parameter_names <- character(n_parameters)
idx <- 1
for (s in unique_site_ids) {
  parameter_names[idx]     <- paste0("Site_", s, "_Z1")
  parameter_names[idx + 1] <- paste0("Site_", s, "_Z2")
  idx <- idx + 2
}# Apply the naming dimensions to the reshaped array
dimnames(all_sites_samples_flat) <- list(
  Iteration = 1:n_iterations,
  Chain     = 1:n_chains,
  Parameter = parameter_names
)
# The formal draws object 
all_sites_aligned_draws <- as_draws_array(all_sites_samples_flat)

# 2. Recalculate draws for aligned lambda (latent factor scores for species)
n_lambda_params <- n_taxa * n_latent
lambda_flat <- array(lambda_unpacked_aligned, dim = c(n_iterations, n_chains, n_lambda_params))
species_names <- colnames(mod_data$y)
lambda_names <- character(n_lambda_params)
idx <- 1
for (j in 1:n_taxa) {
  lambda_names[idx]     <- paste0("Species_", species_names[j], "_Load_Axis1")
  lambda_names[idx + 1] <- paste0("Species_", species_names[j], "_Load_Axis2")
  idx <- idx + 2
}
dimnames(lambda_flat) <- list(
  Iteration = 1:n_iterations,
  Chain     = 1:n_chains,
  Parameter = lambda_names
)
# Cast into a formal MCMC draws object
lambda_aligned_draws <- as_draws_array(lambda_flat)
```

From these aligned z and lambda draws, we revised the diagnostic summary for the model using the following code to produce @tbl_diags, which shows the model to be in good order.

```{r}
#| label: "tbl-diags"

lambda_diagnostic_summary <- summarise_draws(lambda_aligned_draws)
z_diagnostic_summary <- summarise_draws(all_sites_aligned_draws)

mod_bundle$diagnostics$num_divergent; 
mod_bundle$diagnostics$num_max_treedepth
mod_bundle$diagnostics$ebfmi
mod_summary <- mod_bundle$summary[-grep("z_exp|lambda",mod_bundle$summary$variable),]
mod_summary <- rbind(mod_summary, lambda_diagnostic_summary,z_diagnostic_summary)
min(mod_summary$ess_bulk, na.rm = TRUE)
min(mod_summary$ess_tail, na.rm = TRUE)
mod_summary$variable[which(mod_summary$ess_bulk == min(mod_summary$ess_bulk, na.rm = TRUE))]
diag_summary <- data.frame(spec = c("No. sampling iterations","No. divergences",
                                    "No. max_treedepth exceedences",
                                    "BFMI (Chain 1)","BFMI (Chain 2)","BFMI (Chain 3)","BFMI (Chain 4)",
                                    "Min. Bulk ESS", "Min. Tail ESS","Max R-hat",
                                    "Worst sampled parameter"),
                           value = c(n_iterations,
                                     sum(mod_bundle$diagnostics$num_divergent),
                                     sum(mod_bundle$diagnostics$num_max_treedepth),
                                     round(mod_bundle$diagnostics$ebfmi,2),
                                     round(min(mod_summary$ess_bulk, na.rm = TRUE)),
                                     round(min(mod_summary$ess_tail, na.rm = TRUE)),
                                     round(max(mod_summary$rhat, na.rm = TRUE),2),
                                     mod_summary$variable[which(mod_summary$ess_bulk == min(mod_summary$ess_bulk, na.rm = TRUE))])
                           )
ft_to_word(flextable::flextable(diag_summary))

```


```{r}
# --- 3. EXTRACT POSTERIOR MEANS DROPPING THE SKIPPED SECTIONS ---
# The margin '3' calculates the mean across the 275 observations, 
# seamlessly aggregating across all 1200 iterations and all 4 chains.
mean_obs_z1 <- apply(z_unpacked_aligned[, , , 1], 3, mean, na.rm = TRUE)
mean_obs_z2 <- apply(z_unpacked_aligned[, , , 2], 3, mean, na.rm = TRUE)

# --- 4. MAP TO COLLAPSED SITE-LEVEL DATA FRAME ---
# Bind the observation-level averages back to your original site indices
ordination_df <- data.frame(
  Site = mod_data$site, # Replace with your exact site integer array
  Z1   = mean_obs_z1,
  Z2   = mean_obs_z2
)

# Collapse repeat visits down to get exactly 223 rows (one per unique site)
unique_site_ordination <- unique(ordination_df)

# --- 5. GENERATE THE FINAL CLEAN SCATTER PLOT ---
plot(unique_site_ordination$Z1, unique_site_ordination$Z2, 
     pch = 21, bg = grey(0.8), 
     xlab = "Latent Axis 1 (Residual Site Gradients)", 
     ylab = "Latent Axis 2 (Residual Site Gradients)", 
     main = "Aligned Multi-Chain Site Ordination")
abline(v = 0, h = 0, lty = 2)

####

# 1. Build a real-world geographic distance matrix between your 223 sites
coords <- sf::st_coordinates(sites)
dist_matrix <- as.matrix(dist(coords))

# 2. Invert the distance matrix to create a spatial weights matrix (closer = heavier weight)
weights_matrix <- 1 / dist_matrix
diag(weights_matrix) <- 0 
weights_matrix[is.infinite(weights_matrix)] <- 0

# 3. Run Moran's I on Latent Axis 1 and Axis 2
moran_z1 <- ape::Moran.I(unique_site_ordination$Z1, weights_matrix)
moran_z2 <- ape::Moran.I(unique_site_ordination$Z2, weights_matrix)

# Inspect the results
print(moran_z1$p.value)
print(moran_z2$p.value)
# Both >0.5, which strongly suggests randomness without spatial clumping.

# Run a regression predicting Axis 1 from geography
Z <- cbind(unique_site_ordination, coords)
spatial_lm1 <- lm(Z1 ~ X * Y, data = Z)
spatial_lm2 <- lm(Z2 ~ X * Y, data = Z)

# Look at the Multiple R-squared values
summary(spatial_lm1)$r.squared  # 0.01
summary(spatial_lm2)$r.squared  # 0.01

```



