---
title: "Model data assembly for the 225-site dataset"
author: "Thomas Wilkins"
format: html
editor: visual
bibliography: references.bib
---

```{r}
#| label: setup
#| echo: false

# source("https://tools.thewerg.unimelb.edu.au/data/mwstr_v13/mwstr_network_functions.R")
source("https://raw.githubusercontent.com/tewilkins/mwbugs_functions/refs/heads/main/mwbugs_functions.R")

requiredPackages <- c("WriteXLS", "rio", "rstan", "tidyverse", "geosphere")
load_packages(requiredPackages) # Installs packages if not installed already

# Create a folder to hold temporary data if not already present
system(command = "mkdir -p temp_data/")

```

## 1. Introduction

This document contains the assembly of environmental and biological data for the analysis of the 2021-23 macroinvertebrate DNA metabarcoding dataset, prepared for the A.
The data is intended for modelling the presence/absences of stream macroinvertebrates with the Stan model `multitaxon_pa_rSAC.stan`, which is readily scalable to larger datasets.
The model takes the following declarations as inputs:

-   n_obs: number of observations (sample pairs)
-   n_taxa: number of taxa (species)
-   n_pred: number of predictor variables in the model (will depend on model selection)
-   n_site: number of unique sites (225)
-   n_seasons: number of seasons (2, Autumn and Spring)

And the following derived data as inputs:

-   y: a matrix of presence-absence data for each taxon and observation (n_obs x n_taxa)
-   u: a matrix of predictor data at each observation (n_obs x n_pred)
-   s: a matrix of subsample proportions for each observation (n_obs x n_taxa)
-   M: a matrix of haversine distances between each pair of sites (n_site x n_site)
-   site: a vector of site indices (numbered 1 to n_site) for each observation (n_obs)
-   season: a vector of season indices (1 if Autumn, 2 if Spring) for each observation (n_obs)

### 1.1. Retrieving sample pair and site data from Melbourne Water databases

The location and environmental variables associated with the 225 sites was recorded in the Melbourne Water macroinvertebrate database [@walsh2020].
Environmental data associated with each site was retrieved from the Melbourne Water stream database [@walsh].
The following code chunk retrieves this data from the relevant database tables:

```{r}
#| label: data_retrieval
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: false

# Open terminal (not in RStudio tab), run:
# ssh -L 5432:localhost:5432 -C -N -l  tewilkins 172.26.150.60

# Connect to MW databases:
mwbugs <- DBI::dbConnect(RPostgres::Postgres(), dbname = "mwbugs",
                          host = "localhost", user = "readonly", password = "reachcode_42")
mwstr <- DBI::dbConnect(RPostgres::Postgres(),dbname = "mwstr",
                          host = "localhost", user = "readonly", password = "reachcode_42")

# Retrieve sample pair data

####  
# Project 3:  Metabarcoded combined-RBA-sample-pair survey of 225 sites across
#   Melbourne 2021-2024 (Nickname MW metabarcoding survey)

samples <- DBI::dbGetQuery(mwbugs, 
                          "SELECT * FROM samples 
                           JOIN sample_project_groups
                           ON samples.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
# returns 275 samples from 225 sites (some sites have 2 samples, one in Autumn and one in Spring)

biota <- DBI::dbGetQuery(mwbugs, 
                           "SELECT * FROM biota 
                           JOIN sample_project_groups
                           ON biota.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
# returns 19429 rows of biota data from 275 samples

# Check samples have biota:
# Drop samples without invertebrate entries
# nrow(samples[!samples$smpcode %in% unique(biota$smpcode),]) # 0 samples without biota

# Only include samples which are part of the metabarcoding project
samples <- samples[samples$processing_method == "two samples combined, homogenized, DNA-metabarcoded",]
# returns 265 samples from 225 sites

biota <- biota[biota$smpcode %in% samples$smpcode,] # Reduced from 19429 to 18560 rows
sites <- sf::st_read(mwbugs, query = paste0("SELECT * FROM sites WHERE sitecode IN ('",
                              paste(unique(samples$sitecode), collapse = "','"), "');"))

### --- Site coordinates for distance matrix
sites_wgs84 <- sf::st_transform(sites$geom, crs = 4326) # Transform geometry into WGS84
sites_coord <- as.data.frame(sf::st_coordinates(sites_wgs84)) # Extract coordinates
sites$coord_X <- sites_coord$X # Add X coordinates to sites
sites$coord_Y <- sites_coord$Y # Add Y coordinates to sites

### --- Sample metadata variables

# Retrieve environmental data from mwstr database
samples$season <- "A"; samples$season[lubridate::month(samples$date) > 6] <- "S" # Season
# regex to identify Riffle habitat in collection_method
samples$riffle <- 0; samples$riffle[grep("riffle", samples$collection_method)] <- 1 # Riffle habitat

### --- Forest cover

# AF calculated for sites rather than reaches and stored in mwbugs database
sites_af <- DBI::dbReadTable(mwbugs, "sitecode_env")

# Check sites are in sites_af:
sites$sitecode_v12[which(!sites$sitecode_v12 %in% sites_af$sitecode_v12)]
# Some sites are not matched:
# "MUC_987b"  "BRL_728e"  "TNK_2042d" "UOZ_1497c" "SPL_263d"  "TY9_79e"   "WMR_82c"   "BLT_87a" 

# Check if there are some sites on the same reach:
sites_af$sitecode_v12[which(substr(sites_af$sitecode_v12,1,nchar(sites_af$sitecode_v12)-1) %in% 
                              c("MUC_987", "BRL_728", "TNK_2042", "UOZ_1497", "SPL_263", "TY9_79", "WMR_82", "BLT_87"))]
# "BRL_728g"  "MUC_987i"  "UOZ_1497h" "SPL_263b"  "TY9_79a"   "WMR_82d"   "BLT_87c" 

# Correct these sitecodes in sites_af to match those on the same reach (if possible):
sites_af$sitecode_v12_corr <- sites_af$sitecode_v12
sites_af$sitecode_v12_corr[which(sites_af$sitecode_v12 %in% 
                              c("BRL_728g",  "MUC_987i",  "UOZ_1497h", "SPL_263b",  "TY9_79a",   "WMR_82d",   "BLT_87c"))] <-
  c("BRL_728e",  "MUC_987b",  "UOZ_1497c", "SPL_263d",  "TY9_79e",   "WMR_82c",   "BLT_87a")

# Check all sites are now matched
sites$sitecode_v12[which(!sites$sitecode_v12 %in% sites_af$sitecode_v12_corr)]
# Still one unmatched: "TNK_2042d"
# Remove for now
sites <- sites[sites$sitecode_v12 != "TNK_2042d",]
samples <- samples[samples$sitecode_v12 != "TNK_2042d",]
biota <- biota[biota$smpcode %in% samples$smpcode,]

# Join environmental data to sites
# A major storm on 9 June 2021 caused much forest loss https://www.theguardian.com/australia-news/2021/jun/24/dandenong-ranges-power-outage-storm-debris-victoria-emergency-services-warning
# therefore use af_2018 before this date and af_2022 after.
samples$af <- sites_af$af_2022[match(samples$sitecode_v12, sites_af$sitecode_v12_corr)]
samples$af[samples$date < "2021-06-10"] <- sites_af$af_2018[match(samples$sitecode_v12[samples$date < "2021-06-10"],
                         sites_af$sitecode_v12)]

### --- Impervious cover and other environmental variables

# Other variables can be adequately estimated from reach values.
cat_env <- DBI::dbGetQuery(mwstr,paste0("SELECT *, subcs.carea_km2 FROM cat_env ",
                                         "JOIN subcs ON cat_env.site = subcs.site ",
                                         "WHERE cat_env.reach IN ('",
                                         paste(sites$reach_v12, collapse = "', '"), "');"))
subc_env <- DBI::dbGetQuery(mwstr,paste0("SELECT * FROM subc_env WHERE reach IN ('",
                                         paste(sites$reach_v12, collapse = "', '"), "');"))

samples$reach_v12 <- sites$reach_v12[match(samples$sitecode, sites$sitecode)] # Add reach_v12 to samples
samples$carea_km2 <- cat_env$carea_km2[match(samples$reach_v12, cat_env$reach)] # Catchment area
samples$dor <- cat_env$dor_2010[match(samples$reach_v12,cat_env$reach)] # Degree of regulation
samples$ei <- cat_env$ei_2022[match(samples$reach_v12,cat_env$reach)] # Effective imperviousness
samples$ti <- cat_env$ti_2022[match(samples$reach_v12,cat_env$reach)] # Total imperviousness
samples$tf <- cat_env$tf_2022[match(samples$reach_v12,cat_env$reach)] # Total forest
samples$tf[samples$date < "2021-06-10"] <- 
  cat_env$tf_2018[match(samples$reach_v12[samples$date < "2021-06-10"],
                         cat_env$reach)] # Adjust for storm-related forest loss
samples$meanq <- cat_env$meanq_mm[match(samples$reach_v12,cat_env$reach)] # Mean annual runoff
samples$meant <- subc_env$meant_y30_2022[match(samples$reach_v12,subc_env$reach)] # Mean annual temperature
samples$elev <- subc_env$elev_asl_m[match(samples$reach_v12,subc_env$reach)] # Elevation above sea level
samples$sitecode <- as.factor(samples$sitecode)
samples$siteno <- as.numeric(as.factor(samples$sitecode_v12))
samples$coord_X <- sites$coord_X[match(samples$sitecode, sites$sitecode)]
samples$coord_Y <- sites$coord_Y[match(samples$sitecode, sites$sitecode)]

samples <- samples[order(samples$siteno, samples$monthcode),] # Order by site and season

# Disconnect from databases
DBI::dbDisconnect(mwbugs)
DBI::dbDisconnect(mwstr)

# Write to excel file:
writexl::write_xlsx(list("samples" = samples,
                         "biota" = biota,
                         "sites" = sites),
                   path = "temp_data/mw_metabarcoding_225site_model_data.xlsx")

# To WERG drive:
writexl::write_xlsx(list("samples" = samples,
                          "biota" = biota,
                          "sites" = sites),
                    path = "~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/mw_metabarcoding_225site_model_data.xlsx")

```

## 2. Analysis in Stan

### 2.1. Assembling model data for Stan

The following code chunk assembles the model data for input to the Stan model `multitaxon_pa_rSAC.stan`.

```{r}
#| label: stan_data
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

# Load data from excel file
data_list <- rio::import_list("temp_data/mw_metabarcoding_225site_model_data.xlsx")

samples <- data_list$samples
biota <- data_list$biota

### Preparing predictor data

# scale fixed predictor variables
# keeping them as scaled objects for ease of back-transforming later
I <- scale(log(samples$ei + 0.01))
F <- scale(samples$af)
Q <- scale(samples$meanq^0.5)
Q2 <- scale(samples$meanq^0.5)^2
C <- scale(log(samples$carea_km2))
D <- scale(pmin(samples$dor, 1))
T <- scale(samples$meant)
nriff <- scale(samples$riffle)

# Random effects
site <- samples$siteno
site_group <- samples$site_group
season <- as.numeric(as.factor(samples$season)) # Autumn = 1, Spring = 2

# Coordinates
coords <- unique(samples[,c("siteno", "coord_X", "coord_Y")])
coords_XY <- cbind(coords$coord_X, coords$coord_Y)
distm_XY <- scale(distm(coords_XY), center = FALSE)

### Biota presence/absence, count and subsample percentage data
biota_ct <- with(biota, ct(smpcode, shortcode, count))
biota_ct <- biota_ct[match(samples$smpcode,row.names(biota_ct)),] # ensure c and u rows are ordered the same way

### Predictor matrix (u in the model)
# Full predictor set: I, F, Q, C, D, T, nriff, Q2, IF, QF, IQ
# Interactions: IF, QF, IQ (all first order interactions only
# Quadratic: Q2 (for modal response of biota to runoff, Q)
# Most plausible model did not include interactions, should remove:

# --- Predictor matrix u for morph samples
I_mod <- as.vector(I)[match(samples$smpcode,row.names(biota_ct))]
F_mod <- as.vector(F)[match(samples$smpcode,row.names(biota_ct))]
Q_mod <- as.vector(Q)[match(samples$smpcode,row.names(biota_ct))]
Q2_mod <- as.vector(Q2)[match(samples$smpcode,row.names(biota_ct))]
C_mod <- as.vector(C)[match(samples$smpcode,row.names(biota_ct))]
D_mod <- as.vector(D)[match(samples$smpcode,row.names(biota_ct))]
T_mod <- as.vector(T)[match(samples$smpcode,row.names(biota_ct))]
nriff_mod <- as.vector(nriff)[match(samples$smpcode,row.names(biota_ct))]

samples_u <- cbind(I = I_mod,                   # Effective imperviousness
                   F = F_mod,                   # Attenuated Forest cover
                   Q = Q_mod,                   # mean annual runoff
                   C = C_mod,                   # catchment area
                   D = D_mod,                   # Degree of regulation
                   T = T_mod,                   # Temperature  
                   nriff = nriff_mod,           # number of riffles at site
                   Q2 = Q2_mod,                 # square of Q for modal response
                   IF = I_mod * F_mod,          # Interaction betwen I and F
                   FQ = F_mod * (Q_mod + Q2_mod),      # Interaction betwen Q and F
                   IQ = I_mod * (Q_mod + Q2_mod))      # Interaction betwen I and Q            

samples_u <- samples_u[match(samples$smpcode,row.names(biota_ct)),] # ensure c and u rows are ordered the same way
samples_u <- samples_u[,c("I","F","Q","C","D","T","nriff","Q2")] # final predictor set doesn't include interactions
# Final predictor set: I, F, Q, C, D, T, nriff, Q2

### Assemble stan data:
biota_225site_data <- list(n_obs = nrow(biota_ct),                 # no. samples
                    n_pred = ncol(samples_u),                     # no. predictors
                    n_taxa = ncol(biota_ct),                       # no. taxa
                    n_site = as.integer(max(samples$siteno)),     # no. sites
                    n_seasons = as.integer(max(season)),                # no. seasons included in the dataset
                    u = as.matrix(samples_u),               # predictor model matrix
                    y = as.matrix(biota_ct),
                    site = site,                            # site number for each obs
                    season = season,                        # sampling either in Autumn (1) or Spring (2)
                    M = as.matrix(distm_XY))

# Write standata file to WERG drive:
saveRDS(biota_225site_data, file = "~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/225site_standata.rds")

# Store files locally (e.g., for use on SPARTAN):
saveRDS(biota_225site_data, file = "temp_data/225site_standata.rds")



```

### 2.2. Creating directories for model outputs

Normally when running a Stan model, the model outputs are written to a directory on the WERG shared drive, as this drive has sufficient space to store large model outputs and is accessible from multiple computers.
However, to run the model on a high-performance computing cluster (such as SPARTAN), it is much easier to set up a local directory, then push the entire local directory to SPARTAN, run the model, then pull the entire directory back to the local computer for analysis.

The following code chunk creates directories for storing model outputs both on the WERG shared drive and locally.

```{r}
#| label: create_output_dirs
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

# Create directory for model outputs on WERG shared drive
## Create local directory for each model set:
system(command = "mkdir -p 225-site_model_objects/")
system(command = "mkdir -p 225-site_model_objects/draws/")
system(command = "mkdir -p 225-site_model_objects/diagnostics/")

```

SPARTAN operates by running batch scripts on a remote server.
The batch scripts are run through slurm, a workload manager for Linux clusters.
The following is an example slurm batch script that can be used to run the Stan model on SPARTAN (saved as `225site.sbatch`):

```{r}
#| label: create_sbatch_script
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

sbatch_contents <- "#!/bin/bash
#SBATCH -p sapphire                     # Select partition (high CPU)
#SBATCH --job-name=225_site             # Job name
#SBATCH --ntasks=1                      # Number of tasks
#SBATCH --cpus-per-task=4               # Number of CPU cores per task - 4 required for Stan sampler
#SBATCH --mem=500000                    # Memory per node
#SBATCH --time=24:00:00                 # Time limit

# Load the R module
module load foss/2022a R/4.2.2

# Run the R script corresponding to the array job ID
Rscript 225-site_model_objects/225site_model.R


##DO NOT ADD/EDIT BEYOND THIS LINE##
##Job monitor command to list the resource usage
my-job-stats -a -n -s
"

# Write it to a file
writeLines(sbatch_contents, "225-site_model_objects/225site.sbatch")

```

### 2.3. Stan model script

The following is an example R script that can be used to run the Stan model on SPARTAN (saved as `225site_model.R`):

```{r}
#| label: create_stan_rscript 
#| echo: true 
#| message: false 
#| warning: false 
#| results: "hide"

# load stan data

# standata <- readRDS("~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/biota_225site_standata.rds")

standata <- readRDS("temp_data/biota_225site_standata.rds") # local version

# Stop code from accidentally running on local computer

stop() # unblock on SPARTAN or water-prod

library(StanHeaders, lib="~/R_libs/4.2.2") 
library(rstan, lib="~/R_libs/4.2.2") 
source("functions/rstan_functions.R")

# --- Set project seed

proj_seed <- 12345 
set.seed(proj_seed)

# --- Set cores and load stan model:

# rstan_options(auto_write = TRUE)

options(mc.cores = parallel::detectCores()) 
print("Sampling Stan model...") 
model_fit <- rstan::stan(file = "stancode/multitaxon_pa_rSAC_noSS.stan", 
                         data = standata, 
                         control = list(adapt_delta = 0.9), 
                         seed = proj_seed, 
                         chains = 4, 
                         iter = 4000)

print("Model finished sampling.") 
# --- Save model outputs to WERG shared drive: 
print("Saving model draws...") 
draws <- rstan::extract(model_fit, permuted = TRUE) 
saveRDS(draws, file = "225-site_model_objects/draws/draws.rds") 
saveRDS(draws$y_pred, file = "225-site_model_objects/draws/y_pred.rds")
saveRDS(draws$mu, file = "225-site_model_objects/draws/mu_draws.rds") print("Draws saved.")

# --- Generate diagnostics report

print("Generating diagnostics report...") 
mon <- as.data.frame(monitor(model_fit, print = FALSE)) 
write.csv(mon, file = "225-site_model_objects/diagnostics/mon.csv") 
report <- convergence_report(model_fit, mon) 
print_convergence_report(report)

```
