---
title: "Model data assembly for the 225-site dataset"
author: "Thomas Wilkins and Christopher J Walsh"
format: html
editor: source
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false

# Random seed to fix Stan results
proj_seed <- 12345 
set.seed(proj_seed)

# Network functions used herein: 
source("https://tools.thewerg.unimelb.edu.au/mwstr/data/mwstr_network_functions.R")
# Bug database functions used herein: ct(), 
source("https://tools.thewerg.unimelb.edu.au/mwbugs/data/bug_database_functions.R")
right <- function(x, n) {substr(x, nchar(x) - n + 1, nchar(x))}

# Create a folder to hold temporary data if not already present
system(command = "mkdir -p temp_data/")

# pairs add-on function from pairs() help examples
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}

```

## Introduction

This document contains the assembly of environmental and biological data for the analysis of the 2021-23 macroinvertebrate DNA metabarcoding dataset. The data is intended for modelling the presence/absences of stream macroinvertebrates with the Stan model `multitaxon_pa_rSAC.stan`, which is readily scalable to larger datasets. The model takes the following declarations as inputs:

-   n_obs: number of observations (sample pairs)
-   n_taxa: number of taxa (species)
-   n_pred: number of predictor variables in the model (will depend on model selection)
-   n_site: number of unique sites (223)

And the following derived data as inputs:

-   y: a matrix of presence-absence data for each taxon and observation (n_obs x n_taxa)
-   u: a matrix of predictor data at each observation (n_obs x n_pred)
-   s: a matrix of subsample proportions for each observation (n_obs x n_taxa)
-   M: a matrix of distances between each pair of sites (n_site x n_site)
-   site: a vector of site indices (numbered 1 to n_site) for each observation (n_obs)
-   season: a vector of season indices (1 if Autumn, 2 if Spring) for each observation (n_obs)

In this document version, the model was developed using a subset of 33 species from the full dataset (which contains )

### Retrieving site and sample from the Melbourne Water macroinvertebrate database

We drew the site locations, sample specifications and species occurrence data in each sample from the Melbourne Water macroinvertebrate database [@walsh2025].  The following code retrieves these data from the database.  See @walsh_2023 for access instructions.

```{r}
#| echo: false

# Open terminal (not in RStudio tab), run:
# ssh -L 5432:localhost:5432 -C -N -l  tewilkins 172.26.150.60
```

```{r}
#| label: mwbugs_data_retrieval_1
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

# Connect to MW databases:
mwbugs <- DBI::dbConnect(RPostgres::Postgres(), dbname = "mwbugs")

####  
# Project 3:  Metabarcoded combined-RBA-sample-pair survey of 223 sites across
#   Melbourne 2021-2024 (Nickname MW metabarcoding survey)

samples <- DBI::dbGetQuery(mwbugs, 
                          "SELECT * FROM samples 
                           JOIN sample_project_groups
                           ON samples.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
samples$sitecode[grep("TOO",samples$sitecode)] <- "TOO-4334-2"
samples$sitecode_v12[grep("TOO",samples$sitecode_v12)] <- "TOO_4332c"

sites <- sf::st_read(mwbugs, query = 
                      paste0("SELECT * FROM sites WHERE sitecode_v12 IN ('",
                             paste(samples$sitecode_v12, collapse = "','"),"');"))

# Create an integer site number for use in the Stan model
sites <- dplyr::mutate(sites, site_no = 1:nrow(sites), .before = sitecode)

```

```{r}
#| label: sample_nos
#| echo: false

# x <- aggregate(samples$smpcode, by = list(sitecode = samples$sitecode_v12), FUN = length)
# aggregate(x$sitecode, by = list(n_samps = x$x), FUN = length)
#   n_samps   x
# 1       1 202
# 2       2  13
# 3       3   1
# 4       4   1
# 5       5   8

```

The SQL query retrieved 275 samples from 225 sites. The 5 samples from Toomuc Creek were assigned to 3 different sitecodes. TOO_4269a (trips 377, 396, 401) is really at the top of reach TOO_4332.  TOO_4332a (trip 389) and TOO_4332c (trip 384) are 200-250 m downstream in the same reach. All of these locations are upstream of the Henry Road drain from the east, and the Henry Road wetland outlet from the west, and all are downstream of the next large drain upstream (Banjo Patterson Drive). They should therefore all be considered the same site which we assigned to "TOO_4332c" in the code above. This reduced the number of sites to 223.  Among the 275 samples, 200 sites were sampled once, 13 twice, and 10 3-5 times.

<!--# CW note to self: Perhaps also worth considering removing LSN, with a lot of SCMs not accounted for in its EI measurement -->


```{r}
#| label: mwbugs_data_retrieval_2
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

biota <- DBI::dbGetQuery(mwbugs, 
                           "SELECT * FROM biota 
                           JOIN sample_project_groups
                           ON biota.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
biota <- biota[!grepl("EXCLUDE this record from 3-rep", 
                          biota$notes),]  #20584

species <- unique(biota[c("taxoncode","taxon")])
# nrow(samples[!samples$smpcode %in% unique(biota$smpcode),])  # 0
# confirms there are 0 samples without biota.

```

<!--# CW note to self: chase down these bibtex records.-->

The biota table contained 20694 occurrence records of 882 species from the 275 samples. The dataset included samples from the bulk-processing experiment, which was conducted to develop and optimize the sample-processing methods [@carew_etal_2026, @wilkins_etal_2026a].  The samples for that experiment were were fully processed, with 8-11 replicate MiSeq samples. All species detected in those samples are included in the biota table.  The samples can be adjusted to  approximate the standard 30% subsampled, 3-replicate bulk-processing method by excluding records with the phrase:  'EXCLUDE this record from 3-rep 30% sample' in the notes column, as per the code above. 

## Subset of biota for trial analyses

We began with a tractable small set of species of varying taxonomic relationships and potential response, choosing species with moderate prevalance among sites (between 2 and 80%).

```{r}

# Note range(biota$count) = 1,1
# calculate prevalence on sites rather than samples
# add reach_v12 to samples for matching biota 
samples$reach_v12 <- sites$reach_v12[match(samples$sitecode, sites$sitecode)]
biota$reach <- samples$reach_v12[match(biota$smpcode, samples$smpcode)]
biot_by_site <- unique(biota[c("taxoncode","taxon","reach")])
biot_prev <- aggregate(biot_by_site$taxoncode, 
                       by = list(taxoncode = biot_by_site$taxoncode, taxon = biot_by_site$taxon), 
                       FUN = length)
biot_prev$prevalence <- biot_prev$x/length(unique(samples$reach_v12))
biot_prev <- biot_prev[biot_prev$prevalence >= 0.2 & biot_prev$prevalence <= 0.8,]  # 111 species
# about 1 quarter of that might be a good number to start with
# But first quarantine a few congenerics to make the phylogenetic similarity diverse
spp_trial <- biot_prev[grepl("QT2505",biot_prev$taxoncode) | 
                       grepl("QT2511",biot_prev$taxoncode) |
                       grepl("QE0609",biot_prev$taxoncode),]
biot_prev <- biot_prev[!biot_prev$taxoncode %in% spp_trial$taxoncode,] # down to 105
spp_trial <- rbind(spp_trial, biot_prev[seq(4,104,4),])  #33

biota <- biota[biota$taxoncode %in% spp_trial$taxoncode,]  #down to 2882 records

```

### Phylogenetic variance-covariance matrix

We constructed a classification object for use in the R package taxise, primarily drawing on the Integrated Taxonomic Information System (ITIS) database, validated against the taxonomic tables of the Melbourne Water Macroinvertebrate Database. We constructed taxonomic tables consistent with ITIS conventions for taxa missing from ITIS (using the code: ~/uomShare/wergStaff/ChrisW/git-data/mwbugs/build_phylo_vcv.R). 

The resulting classification object can be accessed and used as shown in the following code. We used the `taxize::class2tree()` function to create a taxonomic tree, and the `ape` package to compute branch lengths to reflect evolutionary divergence and produce a variance-covariance matrix of the species.

```{r}
#| label: fig-spp_tree
#| echo: true
#| message: false
#| warning: false
#| fig-height: 7
#| fig-width: 6
#| fig-cap: "The taxonomic tree for the trial subset of 33 species modelled here" 

spp_class <- readRDS(
  url("https://tools.thewerg.unimelb.edu.au/mwbugs/data/spp_classes_itis.rds"))
# Select the subset of taxa relevant for the current analysis (or use all 982 spp)
spp_class <- spp_class[c(spp_trial$taxon)]
spp_tree <- taxize::class2tree(spp_class)
# compute branch lengths to reflect evolutionary divergence
spp_tree <- ape::compute.brlen(spp_tree$phylo)
# compute the variance-covariance matrix
spp_vcv <- ape::vcv.phylo(spp_tree)
plot(spp_tree, cex = 0.75)
```


### Spatial autocorrelation among sites

Conceptually there are two types of distance of ecological relevance to in-stream biota.

#### 1. Euclidean distances

Direct distances between sites, calculated as the Euclidean distance between the site coordinates. The sites table is projected in Map Grid of Australia 2020 Zone 55, and we divided the coordinates by 1000 for distances in km.  Direct distances are an appropriate indicator of potential movement between sites by species with aerial dispersal.  The model `multitaxon_pa_ac_euc.stan` reads the site coordinates as a two-vector array, and calculates a spatial variance-covariance matrix with exponential decay using the Stan function `gp_exponential_cov` (or squared exponential decay from `gp_exp_quad_cov`), from which it estimates a spatial-autocorrelation effect. In the code below, we calculated square distance matrices in R to explore the distribution of between-site distances and arrive at appropriate prior for the exponential decay parameter (a different prior was optimal for squared exponential decay: see comments in the Stan model).

```{r}
#| label: euc_distance_matrix
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

# 1. Euclidean distance (dist_euc_inv_w)
sites_coords_km <- sf::st_coordinates(sites)/1000
dist_euc_km <- as.matrix(stats::dist(sites_coords_km, diag = TRUE, upper = TRUE))


```

```{r}
#| echo: false

## Assessment of appropriate priors for rho, where range of nearest-neighbour distances are:
dist_range <- c(0.15,170)

## Google advice  https://share.google/aimode/SVs3hXymcnHtv0qfO
# mode of invgamma is scale/(shape + 1), and the mode should be near or slightly larger than the minimum distance
# shape <= 2 results in infinite variance, permitting esxploration of long-range decay patters
alpha <- 1
5/(2.5 -1)
rho_prior <- invgamma::rinvgamma(1000, shape = 2.5, rate = 3)  # rate for this function is what is usually called scale
weight_sq_inv_half_mind <- alpha^2 * exp(-0.5 * (dist_range[1]/(2*rho_prior))^2)
weight_sq_inv_mind <- alpha^2 * exp(-0.5 * (dist_range[1]/rho_prior)^2)
weight_sq_inv_maxd <- alpha^2 * exp(-0.5 * (dist_range[2]/rho_prior)^2)
# posterior value...
plot(seq(0.1,20,0.1), alpha^2 * exp(-0.5 * (seq(0.1,20,0.1)/(2*min(rho_prior)))^2))
# 2.5, 3 limits attenuation to >~1.5 km, which is close to the limits of the dataset
par(mfrow = c(2,2))
hist(rho_prior, breaks = 100)
hist(weight_sq_inv_maxd, breaks = 100)
hist(weight_sq_inv_mind, breaks = 100)
hist(weight_sq_inv_half_mind, breaks = 100)

alpha <- 1
rho_prior <- invgamma::rinvgamma(1000, shape = 2.5, rate = 2.5)
weight_inv_half_mind <- alpha^2 * exp(-dist_range[1]/(2*rho_prior))
weight_inv_mind <- alpha^2 * exp(-dist_range[1]/rho_prior)
weight_inv_maxd <- alpha^2 * exp(-dist_range[2]/rho_prior)
par(mfrow = c(2,2))
hist(rho_prior, breaks = 100)
hist(weight_inv_maxd, breaks = 100)
hist(weight_inv_mind, breaks = 100)
hist(weight_inv_half_mind, breaks = 100)

```



The model (using either decay form) sampled quickly, but failed to converge. Several possible causes are:

- Binary data insufficient to permit inference

- Collinearity with predictors that vary across the landscape (qmean the primary one perhaps)

- little or inconsistent euclidean spatial correlation across the study area (Aerial dispersal much more likely over watersheds [in the original sense: i.e. the ridge between two catchments] in forested areas, and likely to be severely limited in cleared and especially urban areas - cite work of Downes and Lancaster)

#### 2. River distances

River distances were calculated as the along-stream distance between each pair of sites on the same stream line, using stream lines from  the Melbourne Stream Network database [@walsh_2023].  The river distance of primary ecological importance is downstream distance, as that is the dominant direction of movement for stream macroinvertebrates through drift. Some upstream movement is also possible *, and might be considered if I can get downstream working*.

Catchment area is the primary variable that may compete for variance with spatial autocorrelation along rivers.  As an approach to minimize this problem, I regressed catchment area against the longest stream distance upstream of each site, and created a new variable carea_r from the residual of that relationship.


```{r}
#| label: river_distance_matrix
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

mwstr <- DBI::dbConnect(RPostgres::Postgres(), dbname = "mwstr")

# Add reach length and site number (i.e. the unique numeric code for each reach) 
# from mwstr database to sites table
sites$site <- sites$reach_length_m <- NA
for(i in 1:nrow(sites)){
  site_stats <- DBI::dbGetQuery(mwstr, 
                    paste0("SELECT site, length_m FROM streams WHERE reach = '", 
                          sites$reach_v12[i], "';"))
  # some subcs have > 1 line in streams, hence sum and unique in the following
  sites$reach_length_km[i] <- sum(site_stats$length_m) / 1000
  sites$site[i] <- unique(site_stats$site)
} # takes 10 s

# Load the subcs (reach) table from mwstr & convert to an igraph network object
subcs <- DBI::dbGetQuery(mwstr, 
                         "SELECT reach, site, nextds, tot_length_m FROM subcs;")
subcs$tot_length_km <- subcs$tot_length_m/1000
network <- subcs[c("site","nextds","reach","tot_length_km")]
network$nextds[network$nextds == -1] <- -(sum(network$nextds == -1):1)
network_i <- apply(network[,1:2], 2, as.character)
network_ig <- igraph::graph_from_data_frame(network_i)
network_df <- igraph::as_long_data_frame(network_ig)

# Subset the network object to link only the 223 sites of interest
network_223 <- data.frame(vertex = match(sites$site, network_i[,1]),
                          site_no = sites$site_no,
                          site = sites$site, 
                          sitecode = sites$sitecode_v12,
                          nextds = NA, nextds_no = NA, dist_km = NA)
network_223 <- network_223[!duplicated(network_223$site),]

for(i in 1:nrow(network_223)){
  all_ds_i <- igraph::subcomponent(network_ig, network_223$vertex[i], 
                                               mode = "out") # ie. u/s and d/s
  # "out" = downstream. "all" does not return the symmetric matrix required.
  # This needs to be done by adding the "out" matrix to the "in" matrix
  all_ds_223 <- as.numeric(names(all_ds_i)[names(all_ds_i) %in% network_223$site])
  # Next downstream site in pruned network is the second listed subcomponent 
  # (i.e. all downstream)
  network_223$nextds[i] <- all_ds_223[2] 
  if(!is.na(network_223$nextds[i])){
  all_ds_nextds <- as.numeric(names(igraph::subcomponent(network_ig, 
                            all_ds_i[names(all_ds_i) == network_223$nextds[i]], 
                            mode = "out")))
  # all reaches between the site and its nextds (including those two sites)
  reaches_bn <- c(setdiff(names(all_ds_i), all_ds_nextds),network_223$nextds[i])
  ds_sitecode <- sites$sitecode_v12[sites$site == network_223$nextds[i]]
  # Length of all full reaches between the two sites
  network_223$dist_km[i] <- sum(subcs$tot_length_km[subcs$site %in% reaches_bn]) - 
    # minus the distance upstream of the site on the upstream reach 
    (1.05 - 0.1* match(right(network_223$sitecode[i],1),letters)) * 
    sites$reach_length_km[sites$sitecode_v12 == network_223$sitecode[i]] -
    # minus the distance downstream of the site on the downstream reach
     (0.1* match(right(ds_sitecode,1),letters) - 0.05) * 
    sites$reach_length_km[sites$sitecode_v12 == ds_sitecode]
  network_223$nextds_no[i] <- 
    network_223$site_no[network_223$site == all_ds_223[2]]
  }else{
    network_223$dist_km[i] <- Inf
  }
  } # takes ~10 s

network_223$nextds[is.na(network_223$nextds)] <- 
  -(sum(is.na(network_223$nextds)):1)
network_223$nextds_no[is.na(network_223$nextds_no)] <- 
  -(sum(is.na(network_223$nextds_no)):1)
network_223_i <- apply(network_223[,c("site_no","nextds_no")], 2, as.character)
network_223_ig <- igraph::graph_from_data_frame(network_223_i)
igraph::E(network_223_ig)$weight <- network_223$dist_km
M_rivds_raw <- igraph::distances(network_223_ig, mode = "out", 
                                 weights = igraph::E(network_223_ig)$weight)
# Stan requires D[i, j] to be FROM j TO i, so transpose
M_rivds_raw <- t(M_rivds_raw)
# Replace Inf values with a large number for Stan numerical stability
M_rivds_raw[is.infinite(M_rivds_raw)] <- 99999.0

n_ds_pairs <- sum(M_rivds_raw != 99999 & M_rivds_raw > 0)   # 718
M_vec <- as.vector(M_rivds_raw[M_rivds_raw != 99999 & M_rivds_raw > 0])
# hist(M_vec, breaks = 100)
# range(M_vec) #  0.12, 190
# I checked the 0.12 and it's right  BH4_1657 is only 100 m or so downstream of BH1_417
# sum(M_vec < 10)  # 69 pairs < 10 km apart
# sum(M_vec < 5)  # 30 pairs < 5 km apart

```

### Compile environmental predictors

We drew environmental data associated with each site largely from the Melbourne Water stream database [@walsh_2025].  The environmental data in the stream database are derived for the bottom of reaches (typically ~250 m long), which is adequately accurate for all environmental variables except for attenuated forest cover (AF), which can vary along reaches.  Thus, site-specific measures of AF were calculated for each site, and stored in the macroinvertebrate database [@walsh_2023].  The two binary variables, season and riffle, were drawn from the samples table, described aove.
 
```{r}
#| label: env_data_retrieval
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

# Add fields for matching with other tables to the samples table
samples$reach_v12 <- substr(samples$sitecode_v12,1,nchar(samples$sitecode_v12)-1)
samples$site_no <- sites$site_no[match(samples$sitecode_v12, sites$sitecode_v12)]

# Derive season and riffle fields from sample specifications
samples$season <- 0; samples$season[lubridate::month(samples$date) > 6] <- 1 
# Season - 0 = Autumn; 1 = Spring
# samples with collection_method mentionnig riffle, include 1 riffle RBA sample
samples$riffle <- 0; 
samples$riffle[grep("riffle", samples$collection_method)] <- 1 

# Reduce the samples table to relevant fields and add environmental data
samples <- samples[,c("smpcode","date","sitecode_v12","reach_v12",
                     "site_no", "season", "riffle")]

# Attenuated Forest cover: AF - optimal parameterization from wilkins_etal_2026 
# (See also Walsh (2023) Chapter 7)-  drawn from the mwbugs database,
# which accounts for location of the sample site within the reach
sites_af <- DBI::dbGetQuery(mwbugs, 
              paste0("SELECT sitecode_v12, af_l5_w633_1750, ",
                     "af_l5_w633_2018, af_l5_w633_2022 ",
                     "FROM sitecode_env WHERE sitecode_v12 IN ('",
                     paste(sites$sitecode_v12, collapse = "', '"),"');"))
# Check all sites are in sites_af: 
# sum(!sites$sitecode_v12 %in% sites_af$sitecode_v12)  # zero missing. all good.
samples$af <- sites_af$af_l5_w633_2022[match(samples$sitecode_v12, 
                                             sites_af$sitecode_v12)]
# A major storm on 9 June 2021 caused much forest loss https://www.theguardian.com/australia-news/2021/jun/24/dandenong-ranges-power-outage-storm-debris-victoria-emergency-services-warning
# therefore use af_2018 before this date and af_2022 after.
samples$af[samples$date < "2021-06-10"] <- 
sites_af$af_l5_w633_2018[match(samples$sitecode_v12[samples$date < "2021-06-10"],
                         sites_af$sitecode_v12)]
samples$af_ref <- sites_af$af_l5_w633_1750[match(samples$sitecode_v12, 
                                sites_af$sitecode_v12)] 
# reference state for AF to be used for predictions from the ultimate model

# Impervious cover and other environmental variables from the mwstr database
cat_env <- DBI::dbGetQuery(mwstr,
  paste0("SELECT subcs.reach, subcs.carea_km2, ei_l44_sw1942_2022, c_basalt, ", 
         "ti_2022, tf_2018, tf_2022, meanq_mm, dor_2010 FROM cat_env ",
         "JOIN subcs ON cat_env.site = subcs.site ","WHERE cat_env.reach IN ('",
         paste(sites$reach_v12, collapse = "', '"), "');"))

subc_env <- DBI::dbGetQuery(mwstr,
               paste0("SELECT reach, meant_y30_2022, elev_asl_m FROM subc_env ",
                      "WHERE reach IN ('",
                      paste(sites$reach_v12, collapse = "', '"), "');"))
# Catchment area in sq km. 
samples$carea_km2 <- cat_env$carea_km2[match(samples$reach_v12, cat_env$reach)] 
# Proportion of catchment with basalt geology See Walsh (2023) Chapter 5.7
samples$basalt <- cat_env$c_basalt[match(samples$reach_v12, cat_env$reach)]
# Degree of regulation: see Walsh (2023) Chapter 9
samples$dor <- cat_env$dor_2010[match(samples$reach_v12,cat_env$reach)] 
# Effective imperviousness: see Walsh (2023) Chapter 6
samples$ei <- cat_env$ei_l44_sw1942_2022[match(samples$reach_v12,cat_env$reach)] 
# Total imperviousness:  see Walsh (2023) Chapter 6
samples$ti <- cat_env$ti_2022[match(samples$reach_v12,cat_env$reach)] 
# Total forest cover:  see Walsh (2023) Chapter 7
samples$tf <- cat_env$tf_2022[match(samples$reach_v12,cat_env$reach)] 
# Adjust for storm-related forest loss, as above for AF
samples$tf[samples$date < "2021-06-10"] <- 
  cat_env$tf_2018[match(samples$reach_v12[samples$date < "2021-06-10"],
                         cat_env$reach)] 
# Mean annual runoff depth (mm). See Walsh (2023) Chapter 8
samples$meanq <- cat_env$meanq_mm[match(samples$reach_v12,cat_env$reach)] 
# Mean annual air temperature (degrees C). See Walsh (2023) Chapter 5.8
samples$meant <- subc_env$meant_y30_2022[match(samples$reach_v12,subc_env$reach)] 
# Elevation above sea-level (m). See Walsh (2023) Chapter 5.9
samples$elev <- subc_env$elev_asl_m[match(samples$reach_v12,subc_env$reach)]

# Order by site and season
samples <- samples[order(samples$site_no, samples$date),] 

```

### Exploratory analysis of outliers and multicollinearity among predictor variables.

Broadly the candidate environmental predictors for the model were: 

- season: a binary variable assessing if species were more likely to occur in spring/early summer (spring) or late summer/autumn (autumn).

- riffle: a binary variable assessing if species were more likely to occur in samples that include riffle habitat.

- forest cover: @wilkins_etal_2026 showed attenuated forest cover (AF) to be a better predictor of assemblage composition than total catchment forest cover (TF), both proportions. They also determined an optimal attenuation parameterization (half-decay distance overland of 5 m and along the stream of 633 m), which we use.  @fig-explor_1 shows that AF and TF are likely too correlated to use both in a model (primarily because of the strong agreement at either extreme).  There are, however, many sites where TF and AF diverge strongly between 0.2 and 0.9...maybe we *should* include both!

- imperviousness: @wilkins_etal_2026 determined an optimal parameterizatino of effective imperviousness EI, which we use. However, they used a log(x + 0.01) transformation which differs from earlier work, which used log(x _ 0.001). The latter is preferred because it gives greater emphasis to differences between EI and total imperviousness (TI) at low levels of EI (< 1% EI, <5% TI) (@fig-explor_1 C, D).  I will assess if the latter is better. If so, it is another argument for re-doing the optimization.

- Air temperature: @walsh_2025 argued that air temperature provides a good measure of stream temperature in the absence of anthropogenic alterations to effects such as tree cover, stormwater drainage and flow modification. It also has the advantage of being compatible with a measure by which climate-warming predictions are communicated.  The latter argument makes air temperature a superior choice of predictor than elevation with which it is highly correlated (@fig-explor_1 B).

- catchment area (carea_km2) and mean annual discharge depth (meanq). Discharge depth is used rather than volume to separate the effects of flow derived from catchment rainfall (of which meanq is a direct indicator) from river size (which is indicated by catchment area).  meanq is also a good indicator of flow permanence.  We log-transform carea_km2 to reduce the leverage of the small number of large catchments in the dataset. @walsh_webb_2016 found macroinvertebrate response to discharge depth to often be non-linear.  We thus modelled Q as a quadratic function, by including two variables meanq^0.5 (which we term Q) and meanq (which we term Q2). 

```{r}
#| echo: false
#| eval: false

# Working used to arrive at the above conclusions.
# hist(log(samples$carea_km2))

# meanq will be modelled as a quadratic to allow for a non-linear response to meanq
hist(samples$meanq)
# hist(scale(samples$meanq^0.5) - scale(samples$meanq))
# has no extreme outliers, while scale(samples$meanq) - scale(samples$meanq^0.5)
# (Both produce a model response as required)
# To check how scaling affects the distribution of the resulting modal curve:
# plot(samples$meanq^0.5, scale(samples$meanq^0.5) - scale(samples$meanq))
# plot(samples$meanq^0.5, scale(samples$meanq)^0.5 - scale(samples$meanq))
# plot(samples$meanq^0.5, scale(samples$meanq^0.5) - scale(samples$meanq))
# This suggests it is better to independently scale the smeanq and smeanq^2

# hist(pmin(samples$dor,1))

```

- Degree of regulation (DoR) is calculated as the volume of storages in a catchment as a proportion of annual catchment runoff volume. DoR of most sites in this dataset ranged between 0 and 1, but eight had dor > 1, including 2 high outliers. In calculating dor Walsh (2025) capped dor below large service reservoirs, which never spill (but release small environmental flows) at 1. We thus did the same for smaller reservoirs with dor > 1 to minimise leverage of sites below such reservoirs, while retaining the scale of lower dor sites

- Proportion of catchment with basalt geology (basalt) in the Melbourne region is a good predictor of the Western Volcanic Plains streams, which tend to have elevated salinity. Most sites have zero basalt, but 88 sites in the dataset have an even spread of basalt to 1 (@fig-pairs_1. 


```{r}
#| label: fig-explor_1
#| fig-height: 6
#| fig-width: 6
#| fig-cap: "Relationships between similar candidate predictors A. Forest cover, B. Air temperature v Elevation, C, D Imperviousness using two possible transformations."
#| echo: false
#| eval: true

par(mfrow = c(2,2))
plot(samples$af, samples$tf, xlab = "Attenuated forest cover (AF)", 
     ylab = "Total forest cover (TF)", axes = FALSE)
axis(1); axis(2, las = 1); abline(0,1, lty = 3); box(bty = "l")
title(main = paste("A. Forest cover metrics: R =", round(cor(samples$af, samples$tf),2)), 
      font.main =1, cex.main = 0.8, adj = 0)
plot(samples$meant, samples$elev, xlab = "Mean air temperature (deg C)", 
     ylab = "Square-root Elevation ASL (m)", axes = FALSE)
axis(1); axis(2, las = 1); abline(0,1, lty = 3); box(bty = "l")
title(main = paste("B. Air temperarature v. elevation: R =", round(cor(samples$meant, samples$elev),2)), 
      font.main =1, cex.main = 0.8, adj = 0)

plot(log(samples$ei + 0.01), log(samples$ti + 0.01), xlab = "Effective imperviousness (EI) %", 
     ylab = "Total imperviousness (TI) %", axes = FALSE)
axis(1, at = log((c(0,0.1,0.3,1,3,10,30,100) + 1)/100), labels = c(0,0.1,0.3,1,3,10,30,100))
axis(2, at = log((c(0,0.1,0.3,1,3,10,30,100) + 1)/100), labels = c(0,0.1,0.3,1,3,10,30,100), las = 1)
abline(0,1, lty = 3); box(bty = "l")
title(main = paste("C. EI v TI (log[x+0.01]): R =", round(cor(log(samples$ei + 0.01), log(samples$ti + 0.01)),2)), 
      font.main =1, cex.main = 0.8, adj = 0)

plot(log(samples$ei + 0.001), log(samples$ti + 0.001), xlab = "Effective imperviousness (EI) %", 
     ylab = "Total imperviousness (TI) %", axes = FALSE)
axis(1, at = log((c(0,0.1,0.3,1,3,10,30,100) + 0.1)/100), labels = c(0,0.1,0.3,1,3,10,30,100))
axis(2, at = log((c(0,0.1,0.3,1,3,10,30,100) + 0.1)/100), labels = c(0,0.1,0.3,1,3,10,30,100), las = 1)
abline(0,1, lty = 3); box(bty = "l")
title(main = paste("D. EI v TI (log[x+0.001]): R =", round(cor(log(samples$ei + 0.001), log(samples$ti + 0.001)),2)), 
      font.main =1, cex.main = 0.8, adj = 0)

```

From these assessments of predictor distributions, we calculated transformed variables for potential inclusion in the model. We then assessed potential multicollinearity among predictors. With a maximum correlation of 0.67 (between AF and meanq @fig-pairs_1), the level of multicollinearity among the primary predictors is acceptably low.

We also trialled several interactions for inclusion in the model. I won't document the logic until we are sure they could be included.  

All interactions introduce high correlations with main effects, but perhaps this doesn't matter, and they will acti similarly to the quadratic term for smeanq?...

```{r}
#| label: fig-pairs_1
#| fig-height: 7
#| fig-width: 7
#| fig-cap: "A pairs plot of 7 of the primary candidate environmental predictors for the model. season and riffle are not shown, but they are both spread evenly across the other variables (although there are no cases of riffle = 0 at high meanq or dor)."
#| echo: true
#| eval: true

samples$lcarea <- log(samples$carea_km2)
samples$smeanq <- samples$meanq^0.5
samples$dor1 <- pmin(samples$dor,1)
samples$lei_tom <- log(samples$ei + 0.01)
samples$lei <- log(samples$ei + 0.001)
samples$q_af <- samples$smeanq * samples$af
samples$q_ei <- samples$smeanq * samples$lei
samples$b_ei <- samples$basalt * samples$lei
samples$q_d <- log(samples$smeanq * samples$dor1 + 0.1)
samples$af_ei <- samples$af * samples$lei

pairs(samples[c("af","lei","lcarea","smeanq","dor1","basalt")],upper.panel = panel.cor) 

```
```{r}
#| label: fig-pairs_2
#| fig-height: 7
#| fig-width: 7
#| fig-cap: "A pairs plot of 7 of the primary candidate environmental predictors for the model and selected interactions. Not plotted in rendered version"
#| echo: false
#| eval: false

pairs(samples[c("af","lei","lcarea","smeanq","dor1","q_af","q_ei","q_d","af_ei")],upper.panel = panel.cor) 

```


## Analysis in Stan

### Assembling model data for Stan

The following code chunk assembles the model data for input to the Stan model `multitaxon_pa_rSAC.stan`.

```{r}
#| label: stan_data
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

# scale fixed predictor variables
# keeping them as scaled objects for ease of back-transforming later
I <- scale(samples$lei)
F <- scale(samples$af)
Q <- scale(samples$smeanq)
Q2 <- scale(samples$smeanq^2)
C <- scale(samples$lcarea)
D <- scale(samples$dor1)
T <- scale(samples$meant)
B <- scale(samples$basalt)
Q_F <- scale(samples$q_af)
Q_I <- scale(samples$q_ei)
Q_D <- scale(samples$q_d)
F_I <- scale(samples$af_ei)
B_I <- scale(samples$b_ei)
riffle <- samples$riffle
season <- samples$season

# Random effects
site <- samples$site_no

### Biota presence/absence, count and subsample percentage data
biota_ct <- with(biota, ct(smpcode, shortcode, count))
biota_ct <- biota_ct[match(samples$smpcode,row.names(biota_ct)),] # ensure c and u rows are ordered the same way

### Predictor matrix (u in the model)
# Most plausible model did not include interactions...perhaps this was because of the scale-confounding problem???
# But equally there is a multicollinearity problem with the interactions....re-assess if they help or not.

# --- Predictor matrix u for morph samples
I_mod <- as.vector(I)[match(samples$smpcode,row.names(biota_ct))]
F_mod <- as.vector(F)[match(samples$smpcode,row.names(biota_ct))]
Q_mod <- as.vector(Q)[match(samples$smpcode,row.names(biota_ct))]
Q2_mod <- as.vector(Q2)[match(samples$smpcode,row.names(biota_ct))]
C_mod <- as.vector(C)[match(samples$smpcode,row.names(biota_ct))]
D_mod <- as.vector(D)[match(samples$smpcode,row.names(biota_ct))]
T_mod <- as.vector(T)[match(samples$smpcode,row.names(biota_ct))]
B_mod <- as.vector(B)[match(samples$smpcode,row.names(biota_ct))]
Q_F_mod <- as.vector(Q_F)[match(samples$smpcode,row.names(biota_ct))]
Q_I_mod <- as.vector(Q_I)[match(samples$smpcode,row.names(biota_ct))]
Q_D_mod <- as.vector(Q_D)[match(samples$smpcode,row.names(biota_ct))]
F_I_mod <- as.vector(F_I)[match(samples$smpcode,row.names(biota_ct))]
B_I_mod <- as.vector(B_I)[match(samples$smpcode,row.names(biota_ct))]
riff_mod <- as.vector(riffle)[match(samples$smpcode,row.names(biota_ct))]
season_mod <- as.vector(season)[match(samples$smpcode,row.names(biota_ct))]

# reduce parameters while grappling with spatial autocorrelation...
samples_u <- cbind(I = I_mod,                   # Effective imperviousness
                   F = F_mod,                   # Attenuated Forest cover
                   Q = Q_mod,                   # mean annual runoff
                   Q2 = Q2_mod,                 # square of Q for modal response
                   C = C_mod,                   # catchment area
                   # D = D_mod,                   # Degree of regulation
                   # T = T_mod,                   # Temperature 
                   # B = B_mod,                   # Basalt
                   nriff = riff_mod,            # number of riffles at site (0 or 1)
                   season_mod)                  # Autumn = 0, Spring = 1
                   # FI = F_I_mod,                # Interaction betwen I and F
                   # QF = Q_F_mod,                # Interaction betwen Q and F
                   # Q_I = Q_I_mod,               # Interaction between Q and I          
#                   B_I = B_I_mod)               # Interaction between Q and D

save(samples_u, I, F, Q, Q2, C, D, T, B, Q_F, Q_I, Q_D, F_I, B_I, #saved for scaling parameters
     sites, samples, biota, biota_ct, dist_euc_km, 
     file = "~/uomShare/wergStaff/ChrisW/temp/model_allvars_data.rda")


### Assemble stan data:
biota_223site_data <- list(n_obs = nrow(biota_ct),                 # no. samples
                           n_pred = ncol(samples_u),             # no. predictors
                           n_taxa = ncol(biota_ct),                # no. taxa
                           n_site = max(site),                     # no. sites with a site upstream
                           u = as.matrix(samples_u),             # predictor model matrix
                           y = as.matrix(biota_ct),
                           site = site,                            # site number for each obs
#                           M = dist_euc_km)   
                           coords = sf::st_coordinates(sites)/1000)  # distances in km

model_fit <- rstan::stan(file = "~/git/mw_metabarcoding_model/stancode/multitaxon_pa_ac_euc.stan",
                         data = biota_223site_data,
                         save_warmup = TRUE,
               #          control = list(adapt_delta = 0.9),
                         seed = proj_seed,
                         chains = 1,
                         warmup = 2500,
                         iter = 3500)
# save(model_fit, file = "~/uomShare/wergStaff/ChrisW/temp/model_ifqcrs_ac_euc.rda")

# rstan::stan_trace(model_fit, pars = c("rho"))  #, "alpha"
```

```{r}

# Write standata file to WERG drive:
saveRDS(biota_223site_data, file = "~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/223site_standata.rds")

# Store files locally (e.g., for use on SPARTAN):
saveRDS(biota_223site_data, file = "temp_data/223site_standata.rds")

```

### 2.2. Creating directories for model outputs

Normally when running a Stan model, the model outputs are written to a directory on the WERG shared drive, as this drive has sufficient space to store large model outputs and is accessible from multiple computers. However, to run the model on a high-performance computing cluster (such as SPARTAN), it is much easier to set up a local directory, then push the entire local directory to SPARTAN, run the model, then pull the entire directory back to the local computer for analysis.

The following code chunk creates directories for storing model outputs both on the WERG shared drive and locally.

```{r}
#| label: create_output_dirs
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

# Create directory for model outputs on WERG shared drive
## Create local directory for each model set:
system(command = "mkdir -p 225-site_model_objects/")
system(command = "mkdir -p 225-site_model_objects/draws/")
system(command = "mkdir -p 225-site_model_objects/diagnostics/")

```

SPARTAN operates by running batch scripts on a remote server. The batch scripts are run through slurm, a workload manager for Linux clusters. The following is an example slurm batch script that can be used to run the Stan model on SPARTAN (saved as `225site.sbatch`):

```{r}
#| label: create_sbatch_script
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

sbatch_contents <- "#!/bin/bash
#SBATCH -p sapphire                     # Select partition (high CPU)
#SBATCH --job-name=225_site             # Job name
#SBATCH --ntasks=1                      # Number of tasks
#SBATCH --cpus-per-task=4               # Number of CPU cores per task - 4 required for Stan sampler
#SBATCH --mem=500000                    # Memory per node
#SBATCH --time=24:00:00                 # Time limit

# Load the R module
module load foss/2022a R/4.2.2

# Run the R script corresponding to the array job ID
Rscript 225-site_model_objects/225site_model.R


##DO NOT ADD/EDIT BEYOND THIS LINE##
##Job monitor command to list the resource usage
my-job-stats -a -n -s
"

# Write it to a file
writeLines(sbatch_contents, "225-site_model_objects/225site.sbatch")

```

### 2.3. Stan model script

The following is an example R script that can be used to run the Stan model on SPARTAN (saved as `225site_model.R`):

```{r}
#| label: create_stan_rscript 
#| echo: true 
#| message: false 
#| warning: false 
#| results: "hide"

# load stan data

# standata <- readRDS("~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/biota_225site_standata.rds")

standata <- readRDS("temp_data/biota_225site_standata.rds") # local version

# Stop code from accidentally running on local computer

stop() # unblock on SPARTAN or water-prod

library(StanHeaders, lib="~/R_libs/4.2.2") 
library(rstan, lib="~/R_libs/4.2.2") 
source("functions/rstan_functions.R")

# --- Set project seed

proj_seed <- 12345 
set.seed(proj_seed)

# --- Set cores and load stan model:

# rstan_options(auto_write = TRUE)

options(mc.cores = parallel::detectCores()) 
print("Sampling Stan model...") 
model_fit <- rstan::stan(file = "stancode/multitaxon_pa_rSAC_noSS.stan", 
                         data = standata, 
                         control = list(adapt_delta = 0.9), 
                         seed = proj_seed, 
                         chains = 4, 
                         iter = 4000)

print("Model finished sampling.") 
# --- Save model outputs to WERG shared drive: 
print("Saving model draws...") 
draws <- rstan::extract(model_fit, permuted = TRUE) 
saveRDS(draws, file = "225-site_model_objects/draws/draws.rds") 
saveRDS(draws$y_pred, file = "225-site_model_objects/draws/y_pred.rds")
saveRDS(draws$mu, file = "225-site_model_objects/draws/mu_draws.rds") print("Draws saved.")

# --- Generate diagnostics report

print("Generating diagnostics report...") 
mon <- as.data.frame(monitor(model_fit, print = FALSE)) 
write.csv(mon, file = "225-site_model_objects/diagnostics/mon.csv") 
report <- convergence_report(model_fit, mon) 
print_convergence_report(report)

```
