---
title: "Model data assembly for the 225-site dataset"
author: "Thomas Wilkins"
format: html
editor: source
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false

source("https://tools.thewerg.unimelb.edu.au/mwstr/data/mwstr_network_functions.R")
right <- function(x, n) {substr(x, nchar(x) - n + 1, nchar(x))}
# source("https://raw.githubusercontent.com/tewilkins/mwbugs_functions/refs/heads/main/mwbugs_functions.R")

### Tom, I've commented out these lines in order to explicitly (wherever possible)
###   associate functions with their packages (not always possible with piping
###   in the tidyverse). This means I have likely broken some of the code beyond
###   Section 1.2, which is as far as I have got so far.
# requiredPackages <- c("WriteXLS", "rio", "rstan", "tidyverse", "geosphere")
# load_packages(requiredPackages) # Installs packages if not installed already

# Create a folder to hold temporary data if not already present
system(command = "mkdir -p temp_data/")

```

## 1. Introduction

This document contains the assembly of environmental and biological data for the analysis of the 2021-23 macroinvertebrate DNA metabarcoding dataset, prepared for the A. The data is intended for modelling the presence/absences of stream macroinvertebrates with the Stan model `multitaxon_pa_rSAC.stan`, which is readily scalable to larger datasets. The model takes the following declarations as inputs:

-   n_obs: number of observations (sample pairs)
-   n_taxa: number of taxa (species)
-   n_pred: number of predictor variables in the model (will depend on model selection)
-   n_site: number of unique sites (225)
-   n_seasons: number of seasons (2, Autumn and Spring)

And the following derived data as inputs:

-   y: a matrix of presence-absence data for each taxon and observation (n_obs x n_taxa)
-   u: a matrix of predictor data at each observation (n_obs x n_pred)
-   s: a matrix of subsample proportions for each observation (n_obs x n_taxa)
-   M: a matrix of distances between each pair of sites (n_site x n_site)
-   site: a vector of site indices (numbered 1 to n_site) for each observation (n_obs)
-   season: a vector of season indices (1 if Autumn, 2 if Spring) for each observation (n_obs)

### 1.1. Retrieving sample pair and site data from Melbourne Water databases

The location and environmental variables associated with the 225 sites was recorded in the Melbourne Water macroinvertebrate database [@walsh2020]. Environmental data associated with each site was retrieved from the Melbourne Water stream database [@walsh]. The following code chunk retrieves this data from the relevant database tables:

```{r}
#| label: data_retrieval
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

# Open terminal (not in RStudio tab), run:
# ssh -L 5432:localhost:5432 -C -N -l  tewilkins 172.26.150.60

# Connect to MW databases:
mwbugs <- DBI::dbConnect(RPostgres::Postgres(), dbname = "mwbugs",
                          host = "localhost", user = "readonly", password = "reachcode_42")
mwstr <- DBI::dbConnect(RPostgres::Postgres(),dbname = "mwstr",
                          host = "localhost", user = "readonly", password = "reachcode_42")

# Retrieve sample pair data

####  
# Project 3:  Metabarcoded combined-RBA-sample-pair survey of 225 sites across
#   Melbourne 2021-2024 (Nickname MW metabarcoding survey)

samples <- DBI::dbGetQuery(mwbugs, 
                          "SELECT * FROM samples 
                           JOIN sample_project_groups
                           ON samples.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
# returns 275 samples from 225* sites (13 sites have 2 samples, one Autumn and 
# one Spring; the 10 sites from project_no = 1 have up to 5 samples over 3 y)
# * On further investigation, the 5 samples from Toomuc Creek were assigned
#   to 3 different sitecodes. TOO_4269a (trips 377, 396, 401) is really at the
#   top of reach TOO_4332.  TOO_4332a (trip 389) and TOO_4332c (trip 384) are
#   200-250 m downstream in the same reach. All of these locations are upstream
#   of the Henry Road drain from the east, and the Henry Road wetland outlet from
#   the west, and all are downstream of the next large drain upstream (Banjo
#   Patterson Drive). They should therefore all be considered the same site
#   which I will call TOO_4332c. This reduces the number of sites to 223.
samples$sitecode[grep("TOO",samples$sitecode)] <- "TOO-4334-2"
samples$sitecode_v12[grep("TOO",samples$sitecode_v12)] <- "TOO_4332c"
# Check distribution of samples
# x <- aggregate(samples$smpcode, by = list(sitecode = samples$sitecode_v12), FUN = length)
# aggregate(x$sitecode, by = list(n_samps = x$x), FUN = length)
# 202 sites sampled once; 13 sampled twice, 10 (from project 2) sampled 3-5 times.

biota <- DBI::dbGetQuery(mwbugs, 
                           "SELECT * FROM biota 
                           JOIN sample_project_groups
                           ON biota.smpcode = sample_project_groups.smpcode
                           WHERE sample_project_groups.project_code = 3;")
# returns 20691 records 
# NOTE project 3 includes the samples collected for project 2, which were fully
# processed, with 8-11 replicate MiSeq samples. All species detected in those
# samples are included in the biota table.  The samples can be adjusted to 
# approximate the standard 30% subsampled, 3-replicate bulk-processing method
# by excluding records with the phrase: 
# 'EXCLUDE this record from 3-rep 30% sample' in the notes column 
biota <- biota[!grepl("EXCLUDE this record from 3-rep", 
                          biota$notes),]  #20584
# removes 107 records that were found in other replicates/subsamples
# Check samples have biota:
# nrow(samples[!samples$smpcode %in% unique(biota$smpcode),]) # 0 samples without biota

sites <- sf::st_read(mwbugs, query = 
                      paste0("SELECT * FROM sites WHERE sitecode_v12 IN ('",
                             paste(samples$sitecode_v12, collapse = "','"),"');"))
# returns 223 sites (a spatial table, so sf:st_read() was used)

```

### 1.2. Between-site distance matrices

We calculated three matrices of distances between sampling sites (Tom, I just calculated them - we can discuss which of them we use): 

- Euclidean distances, calculated from Map Grid of Australia 2020 Zone 55 coordinates. Tom, your previous reference to haversine distances, while strictly correct, is an unnecessarily complicated term for a small area such as the Melbourne region. crs 7855 (Map Grid of Australia 2020 Zone 55) is a planar projection applicable to our area, meaning that we can consider the earth (around here) flat.  So, we can just use 7855 coordinates (unit metres) to calculate Euclidean distances.  

- upstream river distances, calculated as the along-stream distance from downstream to upstream for all pairs of sites on the same stream line, using the Melbourne Stream Network database stream lines.  
- downstream river distances, calculated as for upstream river distances, but from upstream sites to downstream sites.

Tom, we need to be explicit about what we want to model with these matrices. I am not taken by your argument (from your email):

*I have doubts about how much the similarity between sites overland could be attributed to species-specific aerial dispersal - I think this would be better addressed with some sort of population genomics approach (see Razeng et al., 2017, attached), because, in my thinking, the overland distance doesn’t strictly imply connectivity between sites, just that there is some non-randomness in the spatial arrangement of sites that can be accounted for… though it would be nice to follow up on this with a smaller dataset.*  

I'm not suggesting that we conduct a study of aerial dispersal, but we need to have some biological/ecological/environmental basis for predictors that are seeking to explain non-randomness. If we are seeking to explain a species' distribution, then spatial patterns of distribution have to be either a result of spatial patterns in environmental effects (which we aim to portray with our environmental variables) or movement of species between sites.  In my view, in a well-structured model, spatial autocorrelation should be modelling the latter (movement between sites).  In that case, autocorrelation along Euclidean lines between sites is most likely to be modelling aerial (or possibly overland) dispersal. I find it unsatisfactory to consider autocorrelation with euclidean distance to be just explaining some unspecified non-randomness.

For stream macroinvertebrates, a second likely vector of movement is in-stream.  Generally, downstream movement is considered the dominant vector of dispersal, but up-stream movement is also possible. I have calculated both upstream and downstream matrices because it was easy to do, but if we were to use only one, it would be the upstream river distance matrix (dist_us_km_223), which models the likely effect of movement from upstream sites. (In dist_us_km_233, the most upstream sites have Inf distance from all other sites - i.e. no possible influence from upstream; conversely in dist_ds_km_233 the most downstream sites have Inf distance from all other sites - i.e. no possible influence from downstream...clear as mud?)

Now, 98.1% of all pairs of the 223 sites in the dataset have Inf distance in these two river distance matrices (unsurprising, as my intention in picking sites was to minimise the possibility of spatial autocorrelation among sites).  I feel like Inf is going to be problematic when you scale the distances....Given the maximum non-Inf value in the dataset is 188 km, I wonder if it would be defensible to just set the Inf values to something like 2 times 188 (or 3 times?), given that a priori, we would expect the influence of sites 188 km apart on each other to be zero... 

I don't really understand your suggested solution

*If that’s going to mess up the scaling of the matrix, then a workaround could be setting the distance to 0km in the distance matrix M (giving an inverse-exponential distance of 1), and then transforming the inverse-weighted matrix W in Stan to set all values of 1 to 0. I’ve done this already for the diagonal values in the distance matrix because this would imply the site is its own neighbour (and thereby conflicting with a_site).*

But I'm happy to go with it if you think it will work...This suggestion also makes me doubt my next point...

I also need to be convinced of the wisdom of calculating a distance matrix on samples (rather than sites) as you do in your model (or perhaps I am misunderstanding what you do?).  I think (but am not 100% sure) that modelling variation within sites using a random site variable is different from (and more correct than?) modelling it using a distance matrix where samples from the same site have distance 0.  I think the more correct way to model this (particularly in this set where we have 23 sites with >1 sample) would be have a random site variable and apply the distance matrix to the sites. (This must be possible surely, given the possibility of applying similarity matrices to the similarly random species?)


```{r}
#| label: site_distance_matrices
#| echo: true
#| message: false
#| warning: false
#| results: "hide"
#| eval: true

# 1. Ecuclidean distance matrix
sites$site_no <- 1:nrow(sites)
sites_coords_km <- sf::st_drop_geometry(sites)[c("mgae_7855","mgan_7855")]/1000
dist_euc_km <- as.matrix(stats::dist(sites_coords_km, diag = TRUE, upper = TRUE))
dist_euc_scaled <- scale(dist_euc_km, center = FALSE)

# 2 and 3. The two river distance matrices (Not yet scaled: needs further consideration)

# Add reach length and site number from mwstr database to sites table
sites$site <- sites$reach_length_m <- NA
for(i in 1:nrow(sites)){
  site_stats <- DBI::dbGetQuery(mwstr, 
      paste0("SELECT site, length_m FROM streams WHERE reach = '", 
             sites$reach_v12[i], "';"))
  sites$reach_length_m[i] <- site_stats$length_m
  sites$site[i] <- site_stats$site
} # takes 10 s

# Load the subcs (reach) table from mwstr and convert it to an igraph network object
subcs <- DBI::dbGetQuery(mwstr, "SELECT reach, site, nextds, tot_length_m FROM subcs;")
network <- subcs[c("site","nextds","reach","tot_length_m")]
network$nextds[network$nextds == -1] <- -(sum(network$nextds == -1):1)
network_i <- apply(network[,1:2], 2, as.character)
network_ig <- igraph::graph_from_data_frame(network_i)
network_df <- igraph::as_long_data_frame(network_ig)

# Subset the network object to link only the 223 sites of interest
network_223 <- data.frame(vertex = match(sites$site, network_i[,1]),
                          site = sites$site, 
                          sitecode = sites$sitecode_v12,
                          nextds = NA, dist_km = NA)
network_223 <- network_223[!duplicated(network_223$site),]

for(i in 1:nrow(network_223)){
  all_ds_i <- igraph::subcomponent(network_ig, network_223$vertex[i], 
                                               mode = "out")
  all_ds_223 <- as.numeric(names(all_ds_i)[names(all_ds_i) %in% network_223$site])
  # Next downstream site in pruned network is the second listed subcomponent 
  # (i.e. all downstream)
  network_223$nextds[i] <- all_ds_223[2]
  if(!is.na(network_223$nextds[i])){
  all_ds_nextds <- as.numeric(names(igraph::subcomponent(network_ig, 
                            all_ds_i[names(all_ds_i) == network_223$nextds[i]], 
                            mode = "out")))
  reaches_bn <- setdiff(names(all_ds_i), all_ds_nextds)
  ds_sitecode <- sites$sitecode_v12[sites$site == network_223$nextds[i]]
  # Length of all full reaches between the two sites
  network_223$dist_km[i] <- (sum(subcs$tot_length_m[subcs$site %in% reaches_bn]) - 
    # minus the distance upstream of the site on the upstream reach 
    (1.05 - 0.1* match(right(network_223$sitecode[i],1),letters)) * 
    sites$reach_length_m[sites$sitecode_v12 == network_223$sitecode[i]] -
    # minus the distance downstream of the site on the downstream reach
     (0.1* match(right(ds_sitecode,1),letters) - 0.05) * 
    sites$reach_length_m[sites$sitecode_v12 == ds_sitecode])/1000
  }else{
    network_223$dist_km[i] <- Inf
  }
  } # takes ~10 s

network_223$nextds[is.na(network_223$nextds)] <- -(sum(is.na(network_223$nextds)):1)
network_223_i <- apply(network_223[,c("site","nextds")], 2, as.character)
network_223_ig <- igraph::graph_from_data_frame(network_223_i)
igraph::E(network_223_ig)$weight <- network_223$dist_km

# Downstream river distance matrix
dist_ds_km_223 <- igraph::distances(network_223_ig, 
                                  mode = "out",
                                 weights = network_223$dist_km)
# Remove the terminal sites 
dist_ds_km_223 <- as.matrix(dist_ds_km_223)[1:223,1:223]

# Downstream river distance matrix
dist_us_km_223 <- igraph::distances(network_223_ig, 
                                  mode = "in",
                                 weights = network_223$dist_km)
# Remove the terminal sites 
dist_us_km_223 <- as.matrix(dist_us_km_223)[1:223,1:223]

# sum(dist_ds_km_223 == Inf)/length(dist_ds_km_223) # 98.1% = Inf (i.e. independent)
# sum(dist_us_km_223 == Inf)/length(dist_us_km_223) # same

```

Tom, Chris has not yet got past this point...

### 1.3. Compile environmental predictors

```{r}

### --- Sample metadata variables

# Retrieve environmental data from mwstr database
samples$season <- "A"; samples$season[lubridate::month(samples$date) > 6] <- "S" # Season
# regex to identify Riffle habitat in collection_method
samples$riffle <- 0; samples$riffle[grep("riffle", samples$collection_method)] <- 1 # Riffle habitat

### --- Forest cover

# AF calculated for sites rather than reaches and stored in mwbugs database
sites_af <- DBI::dbReadTable(mwbugs, "sitecode_env")

# Check sites are in sites_af:
sites$sitecode_v12[which(!sites$sitecode_v12 %in% sites_af$sitecode_v12)]
# Some sites are not matched:
# "MUC_987b"  "BRL_728e"  "TNK_2042d" "UOZ_1497c" "SPL_263d"  "TY9_79e"   "WMR_82c"   "BLT_87a" 

# Check if there are some sites on the same reach:
sites_af$sitecode_v12[which(substr(sites_af$sitecode_v12,1,nchar(sites_af$sitecode_v12)-1) %in% 
                              c("MUC_987", "BRL_728", "TNK_2042", "UOZ_1497", "SPL_263", "TY9_79", "WMR_82", "BLT_87"))]
# "BRL_728g"  "MUC_987i"  "UOZ_1497h" "SPL_263b"  "TY9_79a"   "WMR_82d"   "BLT_87c" 

# Correct these sitecodes in sites_af to match those on the same reach (if possible):
sites_af$sitecode_v12_corr <- sites_af$sitecode_v12
sites_af$sitecode_v12_corr[which(sites_af$sitecode_v12 %in% 
                              c("BRL_728g",  "MUC_987i",  "UOZ_1497h", "SPL_263b",  "TY9_79a",   "WMR_82d",   "BLT_87c"))] <-
  c("BRL_728e",  "MUC_987b",  "UOZ_1497c", "SPL_263d",  "TY9_79e",   "WMR_82c",   "BLT_87a")

# Check all sites are now matched
sites$sitecode_v12[which(!sites$sitecode_v12 %in% sites_af$sitecode_v12_corr)]
# Still one unmatched: "TNK_2042d"
# Remove for now
sites <- sites[sites$sitecode_v12 != "TNK_2042d",]
samples <- samples[samples$sitecode_v12 != "TNK_2042d",]
biota <- biota[biota$smpcode %in% samples$smpcode,]

# Join environmental data to sites
# A major storm on 9 June 2021 caused much forest loss https://www.theguardian.com/australia-news/2021/jun/24/dandenong-ranges-power-outage-storm-debris-victoria-emergency-services-warning
# therefore use af_2018 before this date and af_2022 after.
samples$af <- sites_af$af_2022[match(samples$sitecode_v12, sites_af$sitecode_v12_corr)]
samples$af[samples$date < "2021-06-10"] <- sites_af$af_2018[match(samples$sitecode_v12[samples$date < "2021-06-10"],
                         sites_af$sitecode_v12)]

### --- Impervious cover and other environmental variables

# Other variables can be adequately estimated from reach values.
cat_env <- DBI::dbGetQuery(mwstr,paste0("SELECT *, subcs.carea_km2 FROM cat_env ",
                                         "JOIN subcs ON cat_env.site = subcs.site ",
                                         "WHERE cat_env.reach IN ('",
                                         paste(sites$reach_v12, collapse = "', '"), "');"))
subc_env <- DBI::dbGetQuery(mwstr,paste0("SELECT * FROM subc_env WHERE reach IN ('",
                                         paste(sites$reach_v12, collapse = "', '"), "');"))

samples$reach_v12 <- sites$reach_v12[match(samples$sitecode, sites$sitecode)] # Add reach_v12 to samples
samples$carea_km2 <- cat_env$carea_km2[match(samples$reach_v12, cat_env$reach)] # Catchment area
samples$dor <- cat_env$dor_2010[match(samples$reach_v12,cat_env$reach)] # Degree of regulation
samples$ei <- cat_env$ei_2022[match(samples$reach_v12,cat_env$reach)] # Effective imperviousness
samples$ti <- cat_env$ti_2022[match(samples$reach_v12,cat_env$reach)] # Total imperviousness
samples$tf <- cat_env$tf_2022[match(samples$reach_v12,cat_env$reach)] # Total forest
samples$tf[samples$date < "2021-06-10"] <- 
  cat_env$tf_2018[match(samples$reach_v12[samples$date < "2021-06-10"],
                         cat_env$reach)] # Adjust for storm-related forest loss
samples$meanq <- cat_env$meanq_mm[match(samples$reach_v12,cat_env$reach)] # Mean annual runoff
samples$meant <- subc_env$meant_y30_2022[match(samples$reach_v12,subc_env$reach)] # Mean annual temperature
samples$elev <- subc_env$elev_asl_m[match(samples$reach_v12,subc_env$reach)] # Elevation above sea level
samples$sitecode <- as.factor(samples$sitecode)
samples$siteno <- as.numeric(as.factor(samples$sitecode_v12))
samples$coord_X <- sites$mgae_7855[match(samples$sitecode, sites$sitecode)]
samples$coord_Y <- sites$mgan_7855[match(samples$sitecode, sites$sitecode)]

samples <- samples[order(samples$siteno, samples$monthcode),] # Order by site and season

# Disconnect from databases
DBI::dbDisconnect(mwbugs)
DBI::dbDisconnect(mwstr)

# Write to excel file:
writexl::write_xlsx(list("samples" = samples,
                         "biota" = biota,
                         "sites" = sites),
                   path = "temp_data/mw_metabarcoding_225site_model_data.xlsx")

# To WERG drive:
writexl::write_xlsx(list("samples" = samples,
                          "biota" = biota,
                          "sites" = sites),
                    path = "~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/mw_metabarcoding_225site_model_data.xlsx")

```

## 2. Analysis in Stan

### 2.1. Assembling model data for Stan

The following code chunk assembles the model data for input to the Stan model `multitaxon_pa_rSAC.stan`.

```{r}
#| label: stan_data
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

# Load data from excel file
data_list <- rio::import_list("temp_data/mw_metabarcoding_225site_model_data.xlsx")

samples <- data_list$samples
biota <- data_list$biota

### Preparing predictor data

# scale fixed predictor variables
# keeping them as scaled objects for ease of back-transforming later
I <- scale(log(samples$ei + 0.01))
F <- scale(samples$af)
Q <- scale(samples$meanq^0.5)
Q2 <- scale(samples$meanq^0.5)^2
C <- scale(log(samples$carea_km2))
D <- scale(pmin(samples$dor, 1))
T <- scale(samples$meant)
nriff <- scale(samples$riffle)

# Random effects
site <- samples$siteno
site_group <- samples$site_group
season <- as.numeric(as.factor(samples$season)) # Autumn = 1, Spring = 2

# Coordinates
coords <- unique(samples[,c("siteno", "coord_X", "coord_Y")])
coords_XY <- cbind(coords$coord_X, coords$coord_Y)
distm_XY <- scale(stats::dist(coords_XY), center = FALSE)

### Biota presence/absence, count and subsample percentage data
biota_ct <- with(biota, ct(smpcode, shortcode, count))
biota_ct <- biota_ct[match(samples$smpcode,row.names(biota_ct)),] # ensure c and u rows are ordered the same way

### Predictor matrix (u in the model)
# Full predictor set: I, F, Q, C, D, T, nriff, Q2, IF, QF, IQ
# Interactions: IF, QF, IQ (all first order interactions only
# Quadratic: Q2 (for modal response of biota to runoff, Q)
# Most plausible model did not include interactions, should remove:

# --- Predictor matrix u for morph samples
I_mod <- as.vector(I)[match(samples$smpcode,row.names(biota_ct))]
F_mod <- as.vector(F)[match(samples$smpcode,row.names(biota_ct))]
Q_mod <- as.vector(Q)[match(samples$smpcode,row.names(biota_ct))]
Q2_mod <- as.vector(Q2)[match(samples$smpcode,row.names(biota_ct))]
C_mod <- as.vector(C)[match(samples$smpcode,row.names(biota_ct))]
D_mod <- as.vector(D)[match(samples$smpcode,row.names(biota_ct))]
T_mod <- as.vector(T)[match(samples$smpcode,row.names(biota_ct))]
nriff_mod <- as.vector(nriff)[match(samples$smpcode,row.names(biota_ct))]

samples_u <- cbind(I = I_mod,                   # Effective imperviousness
                   F = F_mod,                   # Attenuated Forest cover
                   Q = Q_mod,                   # mean annual runoff
                   C = C_mod,                   # catchment area
                   D = D_mod,                   # Degree of regulation
                   T = T_mod,                   # Temperature  
                   nriff = nriff_mod,           # number of riffles at site
                   Q2 = Q2_mod,                 # square of Q for modal response
                   IF = I_mod * F_mod,          # Interaction betwen I and F
                   FQ = F_mod * (Q_mod + Q2_mod),      # Interaction betwen Q and F
                   IQ = I_mod * (Q_mod + Q2_mod))      # Interaction betwen I and Q            

samples_u <- samples_u[match(samples$smpcode,row.names(biota_ct)),] # ensure c and u rows are ordered the same way
samples_u <- samples_u[,c("I","F","Q","C","D","T","nriff","Q2")] # final predictor set doesn't include interactions
# Final predictor set: I, F, Q, C, D, T, nriff, Q2

### Assemble stan data:
biota_225site_data <- list(n_obs = nrow(biota_ct),                 # no. samples
                    n_pred = ncol(samples_u),                     # no. predictors
                    n_taxa = ncol(biota_ct),                       # no. taxa
                    n_site = as.integer(max(samples$siteno)),     # no. sites
                    n_seasons = as.integer(max(season)),                # no. seasons included in the dataset
                    u = as.matrix(samples_u),               # predictor model matrix
                    y = as.matrix(biota_ct),
                    site = site,                            # site number for each obs
                    season = season,                        # sampling either in Autumn (1) or Spring (2)
                    M = as.matrix(distm_XY))

# Write standata file to WERG drive:
saveRDS(biota_225site_data, file = "~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/225site_standata.rds")

# Store files locally (e.g., for use on SPARTAN):
saveRDS(biota_225site_data, file = "temp_data/225site_standata.rds")



```

### 2.2. Creating directories for model outputs

Normally when running a Stan model, the model outputs are written to a directory on the WERG shared drive, as this drive has sufficient space to store large model outputs and is accessible from multiple computers. However, to run the model on a high-performance computing cluster (such as SPARTAN), it is much easier to set up a local directory, then push the entire local directory to SPARTAN, run the model, then pull the entire directory back to the local computer for analysis.

The following code chunk creates directories for storing model outputs both on the WERG shared drive and locally.

```{r}
#| label: create_output_dirs
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

# Create directory for model outputs on WERG shared drive
## Create local directory for each model set:
system(command = "mkdir -p 225-site_model_objects/")
system(command = "mkdir -p 225-site_model_objects/draws/")
system(command = "mkdir -p 225-site_model_objects/diagnostics/")

```

SPARTAN operates by running batch scripts on a remote server. The batch scripts are run through slurm, a workload manager for Linux clusters. The following is an example slurm batch script that can be used to run the Stan model on SPARTAN (saved as `225site.sbatch`):

```{r}
#| label: create_sbatch_script
#| echo: true
#| message: false
#| warning: false
#| results: "hide"

sbatch_contents <- "#!/bin/bash
#SBATCH -p sapphire                     # Select partition (high CPU)
#SBATCH --job-name=225_site             # Job name
#SBATCH --ntasks=1                      # Number of tasks
#SBATCH --cpus-per-task=4               # Number of CPU cores per task - 4 required for Stan sampler
#SBATCH --mem=500000                    # Memory per node
#SBATCH --time=24:00:00                 # Time limit

# Load the R module
module load foss/2022a R/4.2.2

# Run the R script corresponding to the array job ID
Rscript 225-site_model_objects/225site_model.R


##DO NOT ADD/EDIT BEYOND THIS LINE##
##Job monitor command to list the resource usage
my-job-stats -a -n -s
"

# Write it to a file
writeLines(sbatch_contents, "225-site_model_objects/225site.sbatch")

```

### 2.3. Stan model script

The following is an example R script that can be used to run the Stan model on SPARTAN (saved as `225site_model.R`):

```{r}
#| label: create_stan_rscript 
#| echo: true 
#| message: false 
#| warning: false 
#| results: "hide"

# load stan data

# standata <- readRDS("~/uomShare/wergStaff/tewilkins/git-data/mw_metabarcoding_model/data/biota_225site_standata.rds")

standata <- readRDS("temp_data/biota_225site_standata.rds") # local version

# Stop code from accidentally running on local computer

stop() # unblock on SPARTAN or water-prod

library(StanHeaders, lib="~/R_libs/4.2.2") 
library(rstan, lib="~/R_libs/4.2.2") 
source("functions/rstan_functions.R")

# --- Set project seed

proj_seed <- 12345 
set.seed(proj_seed)

# --- Set cores and load stan model:

# rstan_options(auto_write = TRUE)

options(mc.cores = parallel::detectCores()) 
print("Sampling Stan model...") 
model_fit <- rstan::stan(file = "stancode/multitaxon_pa_rSAC_noSS.stan", 
                         data = standata, 
                         control = list(adapt_delta = 0.9), 
                         seed = proj_seed, 
                         chains = 4, 
                         iter = 4000)

print("Model finished sampling.") 
# --- Save model outputs to WERG shared drive: 
print("Saving model draws...") 
draws <- rstan::extract(model_fit, permuted = TRUE) 
saveRDS(draws, file = "225-site_model_objects/draws/draws.rds") 
saveRDS(draws$y_pred, file = "225-site_model_objects/draws/y_pred.rds")
saveRDS(draws$mu, file = "225-site_model_objects/draws/mu_draws.rds") print("Draws saved.")

# --- Generate diagnostics report

print("Generating diagnostics report...") 
mon <- as.data.frame(monitor(model_fit, print = FALSE)) 
write.csv(mon, file = "225-site_model_objects/diagnostics/mon.csv") 
report <- convergence_report(model_fit, mon) 
print_convergence_report(report)

```
